---
title: "Pairwise Agreement and per-raw-img plots"
format:
  html:
    code-fold: true
---
 
Date: 15/5/2024 - 10/6/2024

## GOHREP
**Goal:**

Investigate changes to avg pairwise agreement without each scorer.

Investigate differences between raw agroinfiltration images

**Hypothesis:**

Excluding the scores of scorers with low % agreement to other scorers will increase the avg pairwise agreement.

Some agroinfiltration images will be more accurately scored than others.

**Rationale:**

Differences in lighting or clarity, or familiarity could cause scorers to score more or less consistently in some agroinfiltration images.

**Experimental plan:**

N/A

### Load required packages.

```{r}
library(readr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(here)
library(RColorBrewer)
```

### Load in the data. Create combined table. Anonymise scorers.

```{r}
path1 <- here("analyses", "03_inter_rater_agreement", "output_combined.csv")
path2 <- here("analyses", "03_inter_rater_agreement", "median_combined.csv")
path3 <- here("analyses", "03_inter_rater_agreement", "scorer_key.csv")

output_data <- read_csv(path1, show_col_types = FALSE)
median_data <- read_csv(path2, show_col_types = FALSE)

scorer_key <- read_csv(path3, show_col_types = FALSE, col_types = cols(ID = col_character(), Scorer_Name = col_character()))

score_data <- output_data %>%
  select(Basename, Row, Col, Pos, Scorer1, Score1, Scorer2, Score2, Scorer3, Score3) %>%
  bind_cols(
      median_data %>%
        select(Median_Score)
  )

id_lookup <- setNames(scorer_key$ID, scorer_key$Scorer_Name)

score_data <- score_data %>%
  mutate(across(c(Scorer1, Scorer2, Scorer3), ~ id_lookup[.]))

score_data

score_data_no_0 <- score_data %>% 
  filter(Median_Score != 0)

score_data_no_0
```

### Create a long table with one scorer and one score in each row.

```{r}
long_score_table <- function(score_data) {
    scorer_data <- score_data %>%
    select(starts_with("Scorer")) %>%
    pivot_longer(cols = everything(), names_to = NULL, values_to = "Scorer") %>%
    filter(!is.na(Scorer))

    score_values <- score_data %>%
    select(starts_with("Score"), -starts_with("Scorer")) %>%
    pivot_longer(cols = everything(), names_to = NULL, values_to = "Score") %>%
    filter(!is.na(Score))

    additional_data <- score_data %>%
    select(Basename, Row, Col, Pos, Median_Score) %>%
    slice(rep(row_number(), each = 3))

    score_data_long <- bind_cols(scorer_data, score_values, additional_data)

    score_data_long$Scorer <- factor(score_data_long$Scorer, levels = sort(as.numeric(unique(score_data_long$Scorer))))

    score_data_long <- score_data_long %>%
    filter(Score != "[]")

    return(score_data_long)
}
score_data_long <- long_score_table(score_data)

score_data_long

score_data_long_no_0 <- long_score_table(score_data_no_0)
```

### What was the percentage agreement between each pair of scorers?

```{r}
pairwise_score_data <- function(score_data_long){
    data <- score_data_long %>%
    mutate(Score = as.numeric(Score)) %>%
    group_by(Basename, Row, Col, Pos) %>%
    mutate(id = row_number()) %>%
    ungroup()

    pairwise_comparisons <- data %>%
    full_join(data, by = c("Basename", "Row", "Col", "Pos")) %>%
    filter(id.x < id.y) %>%
    select(Scorer.x = Scorer.x, Score.x = Score.x, Scorer.y = Scorer.y, Score.y = Score.y) %>%
    group_by(Scorer.x, Score.x, Scorer.y, Score.y) %>%
    summarise(Count = n(), .groups = 'drop')

    filtered_data <- pairwise_comparisons %>%
    filter(!is.na(Score.x) & !is.na(Score.y))

    filtered_copy <- filtered_data
    names(filtered_copy)[names(filtered_copy) %in% c("Scorer.x", "Scorer.y")] <- c("Scorer.y", "Scorer.x")

    filtered_data <- rbind(filtered_data, filtered_copy)

    agreement_data <- filtered_data %>%
    group_by(Scorer.x, Scorer.y) %>%
    summarise(Total = sum(Count),
                Agree = sum(Count * (Score.x == Score.y)),
                Percent_Agree = (Agree / Total) * 100,
                .groups = 'drop')

    agreement_data <- agreement_data[as.integer(agreement_data$Scorer.x) < as.integer(agreement_data$Scorer.y), ]

    all_combinations <- expand.grid(Scorer.x = unique(agreement_data$Scorer.x), Scorer.y = unique(agreement_data$Scorer.y))

    merged_df <- all_combinations %>%
    left_join(agreement_data, by = c("Scorer.x", "Scorer.y")) %>%
    mutate(Total = ifelse(is.na(Total), NA, Total),
            Agree = ifelse(is.na(Agree), NA, Agree),
            Percent_Agree = ifelse(is.na(Percent_Agree), NA, Percent_Agree)) %>%
    filter(as.integer(Scorer.x) < as.integer(Scorer.y))

    return(merged_df)
}
merged_df <- pairwise_score_data(score_data_long)
merged_df_no_0 <- pairwise_score_data(score_data_long_no_0)
```

```{r fig.width=11, fig.height=10}
#| fig-width: 11
#| fig-height: 10

plot_agree <- function(merged_df){
  plot <- ggplot(merged_df, aes(x = Scorer.x, y = factor(Scorer.y, levels = as.character(sort(unique(as.integer(Scorer.y))))), fill = Percent_Agree)) +
  geom_tile(color = "black", size = 1) +
  geom_text(aes(label = ifelse(is.na(Percent_Agree), "", sprintf("%.1f%%", Percent_Agree))), color = "black", size = 5, vjust = 0.5) +
  scale_fill_gradient(low = "blue", high = "red", name = "Percentage Agreement", na.value = "white") +
  labs(x = "Scorer X", y = "Scorer Y") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.key.size = unit(2, "lines"))
}

plot_agree_full <- plot_agree(merged_df)
plot_agree_full_no_zero <- plot_agree(merged_df_no_0)
print(plot_agree_full_no_zero)
```

### Average percentage agreement = p(any two scorers give the same score)

This is the average percentage agreement of each pair of scorers, weighted by the number of CDAs they both scored.

```{r}

average_pairwise_agreement <- function(data) {
  valid_data <- data[!is.na(data$Total) & !is.na(data$Percent_Agree), ]
  
  weighted_sum <- sum(valid_data$Percent_Agree * valid_data$Total)
  
  total_sum <- sum(valid_data$Total)
  
  weighted_average <- weighted_sum / total_sum
  
  return(weighted_average)
}

average_agreement <- average_pairwise_agreement(merged_df)

print("Average Pairwise Agreement Amongst Scorers")
average_agreement
#64.24763

avg_agree_no_0 <- average_pairwise_agreement(merged_df_no_0)
avg_agree_no_0
```

### For each scorer calculate average percentage agreement when their scores are removed
```{r}
remove_scorer <- function(score_data_long, scorer, remove_2_disagree){

    subset_score_data <- subset(score_data_long, Scorer != scorer)
    
    if (remove_2_disagree == TRUE) {

      data_grouped <- subset_score_data %>%
        group_by(Basename, Row, Col, Pos)
      
      groups_with_issues <- data_grouped %>%
        filter(n() == 2) %>%
        mutate(diff_scores = n_distinct(Score) > 1) %>%
        filter(diff_scores) %>%
        ungroup() %>%
        select(Basename, Row, Col, Pos)
      
      subset_score_data <- subset_score_data %>%
        anti_join(groups_with_issues, by = c("Basename", "Row", "Col", "Pos"))
    }

    return(subset_score_data)
}

remove_scorer1 <- remove_scorer(score_data_long, "1", remove_2_disagree = FALSE)
remove_scorer1_rm <- remove_scorer(score_data_long, "1", remove_2_disagree = TRUE)
```

```{r}
unique_scorers <- unique(score_data_long$Scorer)
unique_scorers <- unique_scorers[order(unique_scorers)]
unique_scorers <- c("Original", unique_scorers)

CDA_count <- c()
avg_pairwise_agreement <- c()

for (scorer in unique_scorers) {
  subset_score_data_T <- remove_scorer(score_data_long, scorer, TRUE)

  num_unique_CDAs <- subset_score_data_T %>%
    group_by(Basename, Row, Col, Pos) %>%
    summarise(n = n(), .groups = 'drop_last') %>%
    ungroup() %>%
    nrow()

  pairwise_df_T <- pairwise_score_data(subset_score_data_T)

  pairwise_avg_T <- average_pairwise_agreement(pairwise_df_T)

  CDA_count <- c(CDA_count, num_unique_CDAs)
  avg_pairwise_agreement <- c(avg_pairwise_agreement, pairwise_avg_T)
}

result_df <- data.frame(
    Scorer_Removed = unique_scorers,
    Score_Count = CDA_count,
    Avg_Pairwise_Agreement = avg_pairwise_agreement
)
```

### Percentage agreement increase per removed image

Since we're both trying to maximise agreement whilst maximising the dataset size

```{r}
result_df_2 <- result_df %>%
  mutate(
    Images_Lost = 6336 - Score_Count,
    Change_in_Agreement = Avg_Pairwise_Agreement - 64.24763,
    Agreement_Change_Per_Img_Lost = Change_in_Agreement / Images_Lost,
    Change_Per_Img_x1000 = round(Agreement_Change_Per_Img_Lost * 1000, 2)
  )

result_df_2 <- result_df_2 %>%
  select(Scorer_Removed, Score_Count, Avg_Pairwise_Agreement, Images_Lost, Agreement_Change_Per_Img_Lost, Change_Per_Img_x1000)

result_df_2
```

### Demonstrate random sampling without replacement gives the same answer for original:

```{r}
total_sum <- sum(merged_df$Total, na.rm = TRUE)
print(total_sum)

random_sample_pair_agree <- function(data, sample_num) {
    set.seed(42)
    # Create a set of the unique combinations of Basename, Row, Col, and Pos in data
    CDA_set <- distinct(data, Basename, Row, Col, Pos)

    # Randomly permute the CDA_set once
    CDA_set <- CDA_set[sample(nrow(CDA_set)),]

    # Determine the number of samples to take, limited by the number of available CDAs
    sample_num <- min(sample_num, nrow(CDA_set))

    # Set up vector to store true/false
    agreement_vector <- numeric(0)  # Using numeric() to initialize the vector

    # Loop for the number of samples
    for(i in 1:sample_num) {
        # Subset data by the ith row of the permuted CDA_set
        subset_data <- data %>%
            filter(Basename == CDA_set$Basename[i], 
                   Row == CDA_set$Row[i], 
                   Col == CDA_set$Col[i], 
                   Pos == CDA_set$Pos[i])

        # If there are three rows, randomly remove 1, whose index is randomly sampled from between 1 and 3
        if (nrow(subset_data) == 3) {
            to_remove <- sample(1:3, 1)  # Sample from exactly 1 to 3
            subset_data <- subset_data[-to_remove, ]
        }

        # Check if the remaining two scores agree or not.
        if(nrow(subset_data) == 2 && subset_data$Score[1] == subset_data$Score[2]) {
            # If so, append 1 to the agreement vector
            agreement_vector <- c(agreement_vector, 1)
        } else {
            # Otherwise, append 0 to the agreement vector
            agreement_vector <- c(agreement_vector, 0)
        }
    }
    # Calculate the avg pairwise agreement
    avg_pairwise_agreement <- mean(agreement_vector)

    return(avg_pairwise_agreement)
}

# Usage example (assuming score_data_long is defined)
print("Original: 64.24763%")

agreement_1000 <- random_sample_pair_agree(score_data_long, 1000)
print(paste("Agreement after 1,000 randomly sampled pairs (without replacement):", agreement_1000 * 100, "%")) # 63.9%


```
Why does pairwise agreement increase with most individual scorer removals?

When the scores of a particular score is removed you might expect two outcomes:
- If the average agreement of that scorer with other scorers is less than the overall average, then that overall average would increase.
- If the average agreement of that scorer with other scorers is more than the overall average, then that overall average would decrease.

This follows, but another factor is at play, since having removed those scores, there are a number of CDAs which now have only 2 scores.
- If those scores are disagreeing (36% of the time given the original average), they will be removed.
- If those scores are agreeing (64% of the time given the original average), they will be kept.

What this means is that for every score removed, there is a 36% chance that a further disagreeing pair of scores will be removed.


### Jaccard index between each pair of scorers

This measures how commonly two individuals have both scored the same CDAs relative to all the unique objects they've scored, so it's related to frequency not agreement. If there's a high Jaccard index, then it means that the pair of scorers scored a lot of the same images and not many others. If there's a low Jaccard index, then it means that the pair of scorers scored more images with other scorers than they did this partner.

```{r}
for_jaccard <- score_data_long %>%
mutate(ObjectID = paste(Basename, Row, Col, Pos, sep="_"))

scorer_objects <- for_jaccard %>%
  distinct(Scorer, ObjectID) %>%
  group_by(Scorer) %>%
  summarise(Objects = list(ObjectID))

scorer_objects$Scorer <- as.numeric(scorer_objects$Scorer)

jaccard_index <- function(objects1, objects2) {
  intersect_len = length(intersect(objects1, objects2))
  union_len = length(union(objects1, objects2))
  if (union_len == 0 || intersect_len == 0) {
    return(NA)  # Return NA if there's no overlap or intersection is zero
  } else {
    return(intersect_len / union_len)
  }
}

pairs <- expand.grid(scorer_objects$Scorer, scorer_objects$Scorer)
colnames(pairs) <- c("Scorer1", "Scorer2")
pairs <- pairs[pairs$Scorer1 < pairs$Scorer2, ]

results <- pairs %>%
  rowwise() %>%
  mutate(Jaccard = jaccard_index(scorer_objects$Objects[scorer_objects$Scorer == Scorer1][[1]],
                                 scorer_objects$Objects[scorer_objects$Scorer == Scorer2][[1]]))

```

```{r fig.width=11, fig.height=10}
#| fig-width: 11
#| fig-height: 10
ggplot(results, aes(x = factor(Scorer1), y = factor(Scorer2, levels = as.character((sort(unique(as.integer(Scorer2)))))), fill = Jaccard)) +
  geom_tile(color = "black", size = 1) +
  geom_text(aes(label = ifelse(is.na(Jaccard), "", round(Jaccard, 2))), color = "black", size = 5, vjust = 0.5) +
  scale_fill_gradient(low = "blue", high = "red", name = "Count", na.value = "white") +
  labs(x = "Scorer X", y = "Scorer Y") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.key.size = unit(2, "lines"))
```

A Jaccard index of 0.5 means that for that pair of scorers, half of all the scores they collected were of the same CDA.

In general the Jaccard index between all pairs of scorers was low, peaking at 0.5. This means that whilst there was some overlap in the CDAs scored by each pair, there were many CDAs that were scored with different pairs. 

### How frequently did each pair of scorers score the same image?
```{r fig.width=11, fig.height=10}
#| fig-width: 11
#| fig-height: 10
ggplot(merged_df, aes(x = Scorer.x, y = factor(Scorer.y, levels = as.character(sort(unique(as.integer(Scorer.y))))), fill = Total)) +
  geom_tile(color = "black", size = 1) +
  geom_text(aes(label = ifelse(is.na(Total), "", Total)), color = "black", size = 5, vjust = 0.5) +
  scale_fill_gradient(low = "blue", high = "red", name = "Count", na.value = "white") +
  labs(x = "Scorer X", y = "Scorer Y") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.key.size = unit(2, "lines"))
```

### Agreement by raw agroinfiltration image

Are larger images scored better?

```{r}
# Function to calculate percentage agreement for each image using Median_Score and count unique objects
calculate_image_agreement_with_median_and_counts <- function(data) {
  # Calculate the number of unique positions (Row, Col, Pos) per Basename
  object_counts <- data %>%
    group_by(Basename) %>%
    summarise(Objects_Count = n_distinct(paste(Row, Col, Pos, sep = "_")), .groups = 'drop')
  
  # Calculate agreement using Median_Score
  agreement_data <- data %>%
    group_by(Basename, Row, Col, Pos) %>%
    summarise(Agreement = mean(Score == as.character(Median_Score)), .groups = 'drop') %>%
    group_by(Basename) %>%
    summarise(Avg_Percent_Agreement = mean(Agreement) * 100, .groups = 'drop')

  # Combine the agreement data with the object counts
  result_data <- left_join(agreement_data, object_counts, by = "Basename")
  
  return(result_data)
}

# Apply the function to your dataframe
image_agreements_with_counts <- calculate_image_agreement_with_median_and_counts(score_data_long)
```

```{r fig.width=11, fig.height=10}
#| fig-width: 11
#| fig-height: 10
ggplot(image_agreements_with_counts, aes(x = Objects_Count, y = Avg_Percent_Agreement)) +
  geom_point(alpha = 0.8, color = "black", size = 2) +
  labs(x = "CDA Count", y = "Percent Agreement with Median") +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 1)) + 
  scale_x_continuous(breaks = pretty(image_agreements_with_counts$Objects_Count, n = 10, min.n = 10))
```

All raw images with <60% agreement with the median are smaller images with fewer than 20 CDAs.

The largest images with >100 CDAs have relatively similar levels of agreement with the median, between 50-70%

### Cluster scorers + dendrogram.

It is challenging to use regular hierarchical clustering, since not all scorers scored all CDAs.

For this reason, I'm using the pairwise percentage agreements as the agreement matrix, and substituting 0% agreement for NA values. This means scorers that did not share any images will be treated as far apart, giving priority to data that does exist.

I tried this with 0% and 50% agreement, and although the dendrogram did differ in both cases in terms of distance, the same scorers clustered more closely: ((1, 9), (4, 8)), (6, 7, 10) and (5, 2, 3).

```{r}
  data <- score_data_long %>%
  mutate(Score = as.numeric(Score)) %>%
  group_by(Basename, Row, Col, Pos) %>%
  mutate(id = row_number()) %>%
  ungroup()

  pairwise_comparisons <- data %>%
  full_join(data, by = c("Basename", "Row", "Col", "Pos")) %>%
  filter(id.x < id.y) %>%
  select(Scorer.x = Scorer.x, Score.x = Score.x, Scorer.y = Scorer.y, Score.y = Score.y) %>%
  group_by(Scorer.x, Score.x, Scorer.y, Score.y) %>%
  summarise(Count = n(), .groups = 'drop')

  filtered_data <- pairwise_comparisons %>%
  filter(!is.na(Score.x) & !is.na(Score.y))

  filtered_copy <- filtered_data
  names(filtered_copy)[names(filtered_copy) %in% c("Scorer.x", "Scorer.y")] <- c("Scorer.y", "Scorer.x")

  filtered_data <- rbind(filtered_data, filtered_copy)

  agreement_data <- filtered_data %>%
  group_by(Scorer.x, Scorer.y) %>%
  summarise(Total = sum(Count),
              Agree = sum(Count * (Score.x == Score.y)),
              Percent_Agree = (Agree / Total) * 100,
              .groups = 'drop')

agreement_data
```

```{r fig.width=11, fig.height=10}
#| fig-width: 11
#| fig-height: 10
pairwise_data <- agreement_data

pairwise_data$Scorer.x <- as.numeric(as.character(pairwise_data$Scorer.x))
pairwise_data$Scorer.y <- as.numeric(as.character(pairwise_data$Scorer.y))

max_scorer_id <- max(pairwise_data$Scorer.x, pairwise_data$Scorer.y)
agreement_matrix <- matrix(NA, nrow = max_scorer_id, ncol = max_scorer_id)
agreement_matrix[cbind(pairwise_data$Scorer.x, pairwise_data$Scorer.y)] <- pairwise_data$Percent_Agree / 100
agreement_matrix[cbind(pairwise_data$Scorer.y, pairwise_data$Scorer.x)] <- pairwise_data$Percent_Agree / 100
diag(agreement_matrix) <- 1  # Scorers perfectly agree with themselves

# Convert agreement to distance
distance_matrix <- 1 - agreement_matrix

# Because NAs can't be included, for the pairs that are never compared I need to set their distance to a large value.
# This way, they will never be the closest in the hierarchy.
distance_matrix[is.na(distance_matrix)] <- 1 # Effectively states that scorers disagree always

# Perform hierarchical clustering
hc <- hclust(as.dist(distance_matrix), method = "ward.D2")

# Plot the dendrogram
plot(hc, hang = -1, sub = "", xlab = "Scorers", ylab = "Distance")
```

The unit distance is proportion of disagreement so a distance of 0.25 between scorers 1 and 9 mean there is a 25% disagreement.

### Circular plot showing average median score for each leaf position

```{r fig.width=11, fig.height=10}
#| fig-width: 11
#| fig-height: 10
create_circular_plot <- function(data, num_positions) {
  # Filter out the Basenames with the exact number of positions per leaf
  images_with_positions <- data %>%
    group_by(Basename) %>%
    filter(max(Pos, na.rm = TRUE) == num_positions) %>%
    ungroup()

  # Calculate the average Median_Score per position for these images
  average_scores_per_position <- images_with_positions %>%
    group_by(Pos) %>%
    summarise(Avg_Median_Score = mean(Median_Score, na.rm = TRUE)) %>%
    ungroup()

  # Create the circular plot
  plot <- ggplot(average_scores_per_position, aes(x = factor(Pos, levels = 1:num_positions), y = 1, fill = Avg_Median_Score)) +
    geom_col(width = 1) +
    coord_polar(start = 0) +
    annotate("segment", x = (num_positions/2)+0.5, xend = (num_positions/2)+0.5, y = 1, yend = 1.5, colour = "black", size = 3) +
    scale_fill_gradient(low = "blue", high = "red", name = "Average Median Score") +
    labs(x = "Position", y = "Average Median Score") +
    theme_void()

  return(plot)
}

five_leaf <- create_circular_plot(score_data, 5)
print("Leaves with 5 positions")
print(five_leaf)
six_leaf <- create_circular_plot(score_data, 6)
print("Leaves with 6 positions")
print(six_leaf)
eight_leaf <- create_circular_plot(score_data, 8)
print("Leaves with 8 positions")
print(eight_leaf)

```

It seems that the bottom of the leaf is consistently scored on average lower than the top of the leaf by around half a score.

### Clustering raw agroinfiltration images (using pixel values or using scores)

If I cluster by scores I could see whethe

```{r}