---
title: "Sequential and image-level structure of scoring disagreement"
format:
  html:
    code-fold: true
---

Date: 2026-06-15

## GOHREP

**Goal:**

Determine whether scorers' over- and under-scoring errors are *independent* per CDA, or whether they
**cluster** — either in time (a sequential momentum / anchoring effect, where having just scored high
makes the next score more likely to be high) or by image (an experiment-level effect, where a scorer's
deviation is consistent within a leaf/image but shifts between images).

**Hypothesis:**

1. In the rescore task (originals presented in *randomised* order), drift `new − old` will show little
   sequential autocorrelation — there is no stimulus structure to anchor to.
2. In the original scoring (CDAs scored in image order), residual disagreement will appear autocorrelated,
   but this will be an **image-blocking** effect (consecutive CDAs share a leaf) rather than genuine
   temporal momentum.
3. Between-scorer disagreement will carry a modest but real **image-level** component: two scorers'
   relative offset is more consistent within an image than between images.

**Rationale:**

This follows up a question raised in `06_intra_rater_reliability`: do over-/under-scoring patterns occur
in groups? The danger is mistaking experimental structure for scorer psychology. Original scoring is not
random — infiltration layouts place controls (0s) together, leaves differ in lighting/background, and
neighbouring spots on a leaf share treatment — so consecutive scores are correlated for reasons that have
nothing to do with sequential anchoring. We therefore (a) use the *randomised-order* rescore as a clean
temporal test, (b) decompose the original-scoring autocorrelation into within- vs across-image, and
(c) test image-level structure directly with a confound-free pairwise comparison.

**Experimental plan:**

1. Rescore drift autocorrelation (randomised order → clean temporal test).
2. Original-scoring residual autocorrelation, decomposed into within-image vs across-image consecutive pairs.
3. Between-scorer pairwise disagreement: variance decomposition (ICC) by image, with and without 0-median CDAs.

---

## 1. Setup

### 1.1 Packages

```{r}
# 1.1
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
library(scales)
```

### 1.2 Keys, data and helpers

```{r}
# 1.2
source(here("src", "plotting_helpers.R"))

# Canonical three-scorer panel (one row per CDA).
canon <- read_csv(
  here("analyses", "01_generate_raw_data", "04_median_and_centre_coords", "combined_cda_data_median.csv"),
  show_col_types = FALSE
) %>%
  select(Basename, Row, Col, Pos, Scorer1, Score1, Scorer2, Score2, Scorer3, Score3, Median_Score) %>%
  mutate(
    across(c(Scorer1, Scorer2, Scorer3), as.integer),
    across(c(Score1, Score2, Score3, Median_Score), as.numeric)
  )

save_plot <- make_save_plot("08_disagreement_structure")

# Wald-Wolfowitz runs test on the sign of a series (p-value only; zeros dropped).
runs_p <- function(x) {
  s <- sign(x[x != 0])
  n <- length(s)
  n1 <- sum(s > 0)
  n2 <- sum(s < 0)
  if (n1 == 0 || n2 == 0 || n < 3) {
    return(NA_real_)
  }
  runs <- 1 + sum(s[-1] != s[-n])
  mu <- 1 + 2 * n1 * n2 / n
  v <- 2 * n1 * n2 * (2 * n1 * n2 - n) / (n^2 * (n - 1))
  2 * pnorm(-abs((runs - mu) / sqrt(v)))
}
```

---

## 2. Temporal test 1 — rescore drift (randomised order)

The rescore originals were presented in a randomised order, so consecutive trials are unrelated images.
Any autocorrelation in the drift `new − old` is therefore a **pure sequential effect**. The cost is
power: only ~100 trials per scorer.

```{r}
# 2.1
# rescore_anon/ holds one file per scorer, already anonymised (filename and
# `scorer` column both give the same ID).
rescore_files <- list.files(
  here("analyses", "06_intra_rater_reliability", "rescore_anon"),
  pattern = "[.]csv$", full.names = TRUE
)
# Rows are kept in file order = presentation order.
rescore <- bind_rows(lapply(rescore_files, function(f) {
  read_csv(f, show_col_types = FALSE) %>%
    mutate(
      Scorer = factor(scorer), seq = row_number(),
      old_score = as.integer(old_score), new_score = as.integer(new_score),
      drift = new_score - old_score
    )
})) %>%
  filter(!is.na(drift)) %>%
  group_by(Scorer) %>%
  mutate(cum = cumsum(drift - mean(drift))) %>%
  ungroup()

rescore_stats <- rescore %>%
  group_by(Scorer) %>%
  summarise(
    n = n(),
    lag1_acf = cor(head(drift, -1), tail(drift, -1)),
    runs_p = runs_p(drift),
    mean_drift = mean(drift), .groups = "drop"
  )
print(rescore_stats)

rescore_lag <- rescore %>%
  group_by(Scorer) %>%
  mutate(lag_drift = lag(drift)) %>%
  ungroup()
m_rescore <- lm(drift ~ lag_drift + Scorer, data = rescore_lag)
cat(sprintf(
  "\nPooled lag-1 drift autocorrelation: coef = %.3f, p = %.3f\n",
  coef(m_rescore)["lag_drift"], summary(m_rescore)$coefficients["lag_drift", 4]
))
```

```{r fig.width=14, fig.height=3.5}
#| fig-width: 14
#| fig-height: 3.5
# 2.2
p <- ggplot(rescore, aes(seq, cum)) +
  geom_hline(yintercept = 0, color = "grey60") +
  geom_line(color = "steelblue") +
  facet_wrap(~Scorer, nrow = 1, labeller = label_both) +
  labs(x = "Rescoring order", y = "Cumulative\nmean-centred drift") +
  theme_minimal() +
  theme(
    panel.spacing = unit(0.5, "lines"),
    axis.text = element_text(size = 11), axis.title = element_text(size = 13),
    strip.text = element_text(size = 12)
  )
p
save_plot(p, "rescore_drift_sequence.png", width = 14, height = 3.5)
```

**What this shows:**

- The cumulative drift wanders in long excursions — but a cumulative sum of *independent* noise is a random walk that wanders identically, so the eye is unreliable here; the formal tests are what count.
- Every scorer's runs test is non-significant and the pooled lag-1 autocorrelation is small and only marginal (≈0.08, p≈0.07). With ~100 trials this is underpowered, but there is no clear sequential clustering when the stimulus order is randomised.

**Discussion:**
_Add notes here._

---

## 3. Temporal test 2 — original scoring (image-ordered)

The original scoring was done in image order, with thousands of scores per scorer — far more power, but
consecutive CDAs usually share an image. Each score is compared to the **leave-one-out consensus** (the
median, i.e. mean, of the *other two* scorers on that CDA), giving a residual `score − consensus`. Scores
are matched to the canonical panel by image and CDA position; the leave-one-out reference removes the
self-inclusion that would otherwise force a score to match its own median.

```{r}
# 3
# data/cdascorer_anon/ holds one file per scorer, named Scorer<ID>.csv - see
# the anonymisation script under scripts/.
read_raw_set <- function(set) {
  files <- list.files(here("data", "cdascorer_anon", set), pattern = "[.]csv$", full.names = TRUE)
  bind_rows(lapply(files, function(f) {
    scorer_id <- as.integer(sub("^Scorer", "", tools::file_path_sans_ext(basename(f))))
    read_csv(f, show_col_types = FALSE) %>%
      transmute(
        Set = set, Scorer = scorer_id, seq = row_number(),
        Basename = basename(img), Row = row, Col = col, Pos = pos, score = as.numeric(score)
      )
  }))
}

raw_resid <- bind_rows(read_raw_set("set1"), read_raw_set("set2")) %>%
  inner_join(canon %>% select(Basename, Row, Col, Pos, Score1, Score2, Score3),
    by = c("Basename", "Row", "Col", "Pos")
  ) %>%
  filter(!is.na(Score1), !is.na(Score2), !is.na(Score3)) %>%
  # keep only rows where the raw score is genuinely one of the three panel scores,
  # so (sum - score)/2 is exactly the median of the other two.
  filter(abs(score - Score1) < 1e-9 | abs(score - Score2) < 1e-9 | abs(score - Score3) < 1e-9) %>%
  mutate(
    loo_ref = (Score1 + Score2 + Score3 - score) / 2,
    resid = score - loo_ref
  )

cat(
  "Matched raw scores:", nrow(raw_resid), "across",
  n_distinct(paste(raw_resid$Scorer, raw_resid$Set)), "scorer-sessions\n"
)

session_stats <- raw_resid %>%
  group_by(Scorer, Set) %>%
  arrange(seq, .by_group = TRUE) %>%
  summarise(
    n = n(), lag1_acf = cor(head(resid, -1), tail(resid, -1)),
    runs_p = runs_p(resid), .groups = "drop"
  )
print(as.data.frame(session_stats))
```

### 3.1 Within-image vs across-image autocorrelation

If the autocorrelation is genuine temporal momentum it should persist when consecutive scores come from
*different* images; if it is image-blocking it should vanish at image boundaries.

```{r}
# 3.1.1
raw_lag <- raw_resid %>%
  group_by(Scorer, Set) %>%
  arrange(seq, .by_group = TRUE) %>%
  mutate(lag_resid = lag(resid), lag_base = lag(Basename)) %>%
  ungroup() %>%
  filter(!is.na(lag_resid)) %>%
  mutate(boundary = if_else(lag_base == Basename, "Same image", "Different image"))

cat(
  "Consecutive pairs from the same image:",
  percent(mean(raw_lag$boundary == "Same image"), accuracy = 0.1), "\n\n"
)

boundary_fit <- bind_rows(lapply(c("Same image", "Different image"), function(g) {
  sub <- raw_lag %>% filter(boundary == g)
  f <- lm(resid ~ lag_resid + interaction(Scorer, Set), data = sub)
  ci <- confint(f, "lag_resid")
  tibble(
    boundary = g, coef = unname(coef(f)["lag_resid"]),
    lo = ci[1], hi = ci[2],
    p = summary(f)$coefficients["lag_resid", 4], n = nrow(sub)
  )
}))
boundary_fit$boundary <- factor(boundary_fit$boundary, levels = c("Same image", "Different image"))
print(boundary_fit)
```

```{r fig.width=6, fig.height=4.5}
#| fig-width: 6
#| fig-height: 4.5
# 3.1.2
p <- ggplot(boundary_fit, aes(boundary, coef)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(ymin = lo, ymax = hi), color = "steelblue", linewidth = 0.9, size = 0.9) +
  labs(x = NULL, y = "Lag-1 residual autocorrelation") +
  theme_minimal() +
  theme(axis.text = element_text(size = 13), axis.title = element_text(size = 14))
p
save_plot(p, "raw_image_boundary_autocorrelation.png", width = 6, height = 4.5)
```

**What this shows:**

- Pooled across the original scoring the residual is strongly positively autocorrelated — but this is **entirely a within-image effect**. ~97% of consecutive scores share an image; the lag-1 autocorrelation is large and highly significant for same-image pairs yet indistinguishable from zero across image boundaries (CI spans 0).
- This matches §2: when the context changes — randomised rescore order, or crossing an image boundary — there is no residual momentum. The apparent clustering is image-blocking, not sequential anchoring.

**Discussion:**
_Add notes here._

---

## 4. Image-level structure of disagreement (confound-free)

The temporal tests show disagreement is organised at the image level, but the raw-score autocorrelation
conflates several things (the experimental layout, the abundance of agreed 0s, per-image bias). A cleaner
test, free of temporal order and of the self-vs-median issue, looks at the **difference between two
scorers** on the same CDA: the true severity cancels, leaving only their relative disagreement. If that
difference is consistent within an image but shifts between images, disagreement has an image-level
component. This is the intraclass correlation (ICC) of the pairwise difference with image as the grouping
factor: the fraction of disagreement variance attributable to *which image* a CDA is on.

```{r}
# 4
# All within-CDA scorer pairs, oriented lower-ID minus higher-ID, with image label.
mk_pair <- function(d, a, b) {
  d %>%
    transmute(Basename, Median_Score,
      sa = .data[[paste0("Scorer", a)]], sb = .data[[paste0("Scorer", b)]],
      ya = .data[[paste0("Score", a)]], yb = .data[[paste0("Score", b)]]
    ) %>%
    filter(!is.na(sa), !is.na(sb), !is.na(ya), !is.na(yb)) %>%
    mutate(
      pair = paste(pmin(sa, sb), pmax(sa, sb), sep = "-"),
      diff = if_else(sa < sb, ya - yb, yb - ya)
    )
}
pairs <- bind_rows(mk_pair(canon, 1, 2), mk_pair(canon, 1, 3), mk_pair(canon, 2, 3)) %>%
  select(pair, Basename, Median_Score, diff)

# Random-effects (one-way ANOVA) ICC of the difference by image, for one scorer-pair.
icc_by_pair <- function(df) {
  g <- df %>%
    group_by(Basename) %>%
    summarise(ni = n(), mi = mean(diff), .groups = "drop") %>%
    filter(ni >= 2)
  df <- df %>% semi_join(g, by = "Basename")
  k <- nrow(g)
  N <- nrow(df)
  if (k < 3 || N < 10) {
    return(tibble(k = k, N = N, icc = NA_real_, btw_sd = NA, wth_sd = NA))
  }
  grand <- mean(df$diff)
  ss_b <- sum(g$ni * (g$mi - grand)^2)
  g_mi <- g %>% select(Basename, mi)
  ss_w <- df %>%
    left_join(g_mi, by = "Basename") %>%
    summarise(s = sum((diff - mi)^2)) %>%
    pull(s)
  ms_b <- ss_b / (k - 1)
  ms_w <- ss_w / (N - k)
  n0 <- (N - sum(g$ni^2) / N) / (k - 1)
  var_b <- max(0, (ms_b - ms_w) / n0)
  var_w <- ms_w
  tibble(k = k, N = N, icc = var_b / (var_b + var_w), btw_sd = sqrt(var_b), wth_sd = sqrt(var_w))
}

run_icc <- function(dat, label) {
  res <- dat %>%
    group_by(pair) %>%
    group_modify(~ icc_by_pair(.x)) %>%
    ungroup() %>%
    filter(!is.na(icc)) %>%
    mutate(subset = label)
  cat(sprintf(
    "[%s] %d pairs | median ICC = %.3f | within-image SD = %.2f | between-image SD = %.2f\n",
    label, nrow(res), median(res$icc), median(res$wth_sd), median(res$btw_sd)
  ))
  res
}
icc_all <- run_icc(pairs, "All CDAs (incl. 0s)")
icc_nz <- run_icc(pairs %>% filter(Median_Score > 0), "Non-zero median only")
icc_res <- bind_rows(icc_all, icc_nz) %>%
  mutate(subset = factor(subset, levels = c("All CDAs (incl. 0s)", "Non-zero median only")))
```

### 4.1 ICC of pairwise disagreement by image

```{r fig.width=7, fig.height=5}
#| fig-width: 7
#| fig-height: 5
# 4.1
p <- ggplot(icc_res, aes(subset, icc, fill = subset)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.5) +
  geom_jitter(width = 0.12, height = 0, size = 1.6, alpha = 0.7) +
  scale_fill_manual(values = c(
    "All CDAs (incl. 0s)" = "grey65",
    "Non-zero median only" = "#3182bd"
  ), guide = "none") +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = NULL, y = "ICC of pairwise difference by image\n(per scorer-pair)") +
  theme_minimal() +
  theme(axis.text = element_text(size = 12), axis.title = element_text(size = 13))
p
save_plot(p, "pairwise_image_icc.png", width = 7, height = 5)
```

**What this shows:**

- Each point is one scorer-pair; the ICC is the share of their disagreement variance explained by image.
- The median ICC is modest but clearly non-zero (≈0.14 with 0s, ≈0.22 without): two scorers' relative offset has a real per-image component, while most disagreement remains CDA-specific (within-image SD ≫ between-image SD).
- Crucially, **removing the abundant agreed-0 CDAs raises the ICC** rather than lowering it — so the image-level structure is not an artefact of zero-heavy control regions; it is strongest among genuinely contested CDAs.

**Discussion:**
_Add notes here._

### 4.2 Illustration: per-image mean difference for the best-sampled pairs

```{r fig.width=12, fig.height=4.5}
#| fig-width: 12
#| fig-height: 4.5
# 4.2
top_pairs <- pairs %>%
  count(pair, Basename) %>%
  count(pair, name = "n_images") %>%
  slice_max(n_images, n = 3) %>%
  pull(pair)

img_means <- pairs %>%
  filter(pair %in% top_pairs) %>%
  group_by(pair, Basename) %>%
  summarise(
    ni = n(), mean_diff = mean(diff),
    se = ifelse(n() > 1, sd(diff) / sqrt(n()), NA), .groups = "drop"
  ) %>%
  filter(ni >= 5) %>%
  group_by(pair) %>%
  arrange(mean_diff, .by_group = TRUE) %>%
  mutate(img_rank = row_number()) %>%
  ungroup()

p <- ggplot(img_means, aes(img_rank, mean_diff)) +
  geom_hline(yintercept = 0, color = "grey60") +
  geom_pointrange(aes(ymin = mean_diff - 1.96 * se, ymax = mean_diff + 1.96 * se),
    color = "steelblue", size = 0.3
  ) +
  facet_wrap(~pair, scales = "free_x", labeller = label_both) +
  labs(
    x = "Image (ranked by mean difference)",
    y = "Mean score difference\n(scorer A − scorer B)"
  ) +
  theme_minimal() +
  theme(
    panel.spacing = unit(0.6, "lines"),
    axis.text = element_text(size = 11), axis.title = element_text(size = 13),
    strip.text = element_text(size = 11)
  )
p
save_plot(p, "pairwise_per_image_mean_diff.png", width = 12, height = 4.5)
```

**What this shows:**

- For the best-sampled scorer-pairs, each point is the mean difference on one image with its 95% CI.
- The per-image means span a range (a between-image gradient) wider than the within-image error bars, illustrating the ICC: the two scorers' relative offset shifts from image to image while being comparatively settled within each image.

**Discussion:**
_Add notes here._

---

## 5. Discussion — confounds and interpretation

**Same image = same experiment.** Original scoring is not a random sequence. CDAs on one image are spots
on one leaf from one infiltration, so they share construct, leaf age, lighting and background. Two
effects follow: (i) the underlying severities are correlated, and (ii) a scorer's *reading* of that leaf
(exposure, contrast, what counts as "necrotic" under that lighting) is correlated. The §3 within-image
autocorrelation and the §4 ICC both capture this image-level structure; neither is evidence of temporal
anchoring, which the randomised rescore (§2) and the across-image boundary (§3.1) both rule out.

**The abundance of 0s and non-random layout.** Infiltration layouts group controls together, so a scorer
genuinely does see runs of 0s — but those are near-universally agreed, so they contribute *agreement*,
not disagreement structure. This is why the analyses use **residuals / pairwise differences** rather than
raw scores (which would be trivially autocorrelated by the design), and why the §4 ICC is computed both
with and without 0s. The result is reassuring: excluding 0s *increases* the image-level ICC, so the
structure lives in the contested mid-range CDAs, not in the clustered controls.

**Is class balancing necessary?** For the *temporal* tests, no — and it would be harmful: down-sampling
or re-weighting score classes destroys the true adjacency of the sequence, so a "lag-1" on a balanced
series no longer means consecutive-in-time. The appropriate controls there are analysing residuals (not
raw scores) and the within/across-image split, both of which are done. For the *image-level* ICC, balance
is handled implicitly: the difference between two scorers already removes the severity (class) signal, and
reporting the non-zero subset alongside the full set shows the conclusion is not driven by the dominant
0 class.

**Overall.** Over-/under-scoring does **not** have temporal momentum across contexts — a scorer's last
score does not pull their next one once the image changes. What is real is a modest **image-level**
structure to disagreement: two scorers' relative calibration drifts somewhat from image to image
(plausibly per-leaf lighting/background and per-session adaptation) while staying more consistent within
an image. Most disagreement, however, remains CDA-specific — the irreducible ambiguity of scoring an
individual cell-death area.

**Discussion:**
_Add notes here._
