---
title: "EDA Plots"
format:
  html:
    code-fold: true
---
 
Date: 13/4/24

## GOHREP
**Goal:**

Create plots to visualise the collected scores for exploratory data analysis

**Hypothesis:**

NA - The aim of EDA is to direct hypothesis generation

**Rationale:**

NA

**Experimental plan:**

Plot a range of representations of the scoring data using R and ggplot

### Load packages dplyr and ggplot2
```{r}
#install.packages('IRkernel')
#IRkernel::installspec()

library(readr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)

#install.packages("here")
library(here)
```

### Load in the data and add the median value from median_combined to output_combined
```{r}
path1 <- here("analyses", "02_score_distributions", "output_combined.csv")
path2 <- here("analyses", "02_score_distributions", "median_combined.csv")
path3 <- here("analyses", "02_score_distributions", "scorer_key.csv")

output_data <- read_csv(path1)
median_data <- read_csv(path2)
scorer_key <- read_csv(path3, col_types = cols(ID = col_character(), Scorer_Name = col_character()))

score_data <- output_data %>%
  select(Basename, Row, Col, Pos, Scorer1, Score1, Scorer2, Score2, Scorer3, Score3) %>%
  bind_cols(
      median_data %>%
        select(Median_Score)
  )

id_lookup <- setNames(scorer_key$ID, scorer_key$Scorer_Name)

score_data <- score_data %>%
  mutate(across(c(Scorer1, Scorer2, Scorer3), ~ id_lookup[.]))

head(score_data)

score_data <- score_data %>%
  mutate(across(everything(), ~na_if(.x, '[]')))

outcsvpath <- here("analyses", "02_score_distributions", "original_CDA_data.csv")
write.csv(score_data, file = outcsvpath, row.names = FALSE)

```

### Create an additional data table that just contains Scorer, Score, and Median
```{r}
scorer_data <- score_data %>%
  select(starts_with("Scorer")) %>%
  pivot_longer(cols = everything(), names_to = NULL, values_to = "Scorer") %>%
  filter(!is.na(Scorer))

score_values <- score_data %>%
  select(starts_with("Score"), -starts_with("Scorer")) %>%
  pivot_longer(cols = everything(), names_to = NULL, values_to = "Score") %>%
  filter(!is.na(Score))

additional_data <- score_data %>%
  select(Basename, Row, Col, Pos, Median_Score) %>%
  slice(rep(row_number(), each = 3))

score_data_long <- bind_cols(scorer_data, score_values, additional_data)

score_data_long$Scorer <- factor(score_data_long$Scorer, levels = sort(as.numeric(unique(score_data_long$Scorer))))

score_data_long
```

### Plot 1: Faceted scatter plot - facet x = scorer (discrete), facet y = scorer (discrete), x = score (discrete), y = score (discrete), black line y=x, red line fit points

```{r}

# Assuming 'data' is your dataframe
data <- score_data_long %>%
  mutate(Score = as.numeric(Score)) %>%
  group_by(Basename, Row, Col, Pos) %>%
  mutate(id = row_number()) %>%
  ungroup()

# Assuming 'data' is your dataframe and has been prepared as mentioned earlier
pairwise_comparisons <- data %>%
  full_join(data, by = c("Basename", "Row", "Col", "Pos")) %>%
  filter(id.x < id.y) %>%
  select(Scorer.x = Scorer.x, Score.x = Score.x, Scorer.y = Scorer.y, Score.y = Score.y) %>%
  group_by(Scorer.x, Score.x, Scorer.y, Score.y) %>%
  summarise(Count = n(), .groups = 'drop')

pairwise_comparisons

filtered_data <- pairwise_comparisons %>%
  filter(!is.na(Score.x) & !is.na(Score.y))

filtered_copy <- filtered_data
names(filtered_copy)[names(filtered_copy) %in% c("Scorer.x", "Scorer.y")] <- c("Scorer.y", "Scorer.x")
filtered_data <- rbind(filtered_data, filtered_copy)

filtered_data <- filtered_data[as.integer(filtered_data$Scorer.x) >= as.integer(filtered_data$Scorer.y), ]
filtered_data
```

```{r fig.width=20, fig.height=18}
#| fig-width: 8
#| fig-height: 7

#plot1path <- here("analyses", "02_score_distributions", "plot1.png")
#png(plot1path, width = 20, height = 18, units = "cm", res = 300)

# Plotting with bubble size representing frequency
ggplot(filtered_data, aes(x = Score.x, y = Score.y)) +
  geom_point(aes(size = Count, color = Count), stat = "identity", alpha = 0.8) +  # Points with dynamic size and color
  geom_abline(data = filtered_data, aes(intercept = 0, slope = 1), linetype = "dashed") +  # Y=X line
  scale_size_continuous(name = "Frequency", range = c(0.5, 4)) +  # Adjust size range as needed
  scale_color_gradient(name = "Frequency", low = "blue", high = "red") +  # Color gradient
  facet_grid(Scorer.x ~ Scorer.y, scales = "free", switch = "both") +  # Facet labels on the left and bottom
  theme_bw() +
  labs(x = "Scorer 2", y = "Scorer 1") +  # Titles and labels
  theme(panel.spacing = unit(0.25, "lines"),  # Reduce spacing between facets
        strip.placement = "outside",  # Place facet labels outside the plot area
        panel.grid.major = element_blank(),  # Remove major grid lines
        panel.grid.minor = element_blank(),  # Remove minor grid lines
        panel.background = element_blank())  # Remove panel background

#dev.off()
```


```{r fig.width=20, fig.height=18}
#| fig-width: 8
#| fig-height: 7

#plot1path <- here("analyses", "02_score_distributions", "plot1.png")
#png(plot1path, width = 20, height = 18, units = "cm", res = 300)

# Plotting with bubble size representing frequency
ggplot(filtered_data, aes(x = Score.x, y = Score.y)) +
  geom_point(aes(size = Count, color = Count), stat = "identity", alpha = 0.8, size = 4) +  # Points with dynamic size and color
  geom_abline(data = filtered_data, aes(intercept = 0, slope = 1), linetype = "dashed") +  # Y=X line
  scale_color_gradient(trans = "log10", name = "Frequency", low = "white", high = "black") +  # Color gradient
  facet_grid(Scorer.x ~ Scorer.y, scales = "free", switch = "both") +  # Facet labels on the left and bottom
  theme_bw() +
  labs(x = "Scorer 2", y = "Scorer 1") +  # Titles and labels
  theme(panel.spacing = unit(0.25, "lines"),  # Reduce spacing between facets
        strip.placement = "outside",  # Place facet labels outside the plot area
        panel.grid.major = element_blank(),  # Remove major grid lines
        panel.grid.minor = element_blank(),  # Remove minor grid lines
        panel.background = element_blank())  # Remove panel background

#dev.off()
```

### Plot 2: The same but with heatmaps

```{r fig.width=20, fig.height=18}
#| fig-width: 20
#| fig-height: 18

#plot2path <- here("analyses", "02_score_distributions", "plot2.png")
#png(plot2path, width = 20, height = 18, units = "cm", res = 300)

# Creating the heatmap plot
ggplot(filtered_data, aes(x = Score.x, y = Score.y)) +
  geom_tile(aes(fill = Count), color = "black", size = 1) +  # Use fill for frequency
  geom_abline(data = filtered_data, aes(intercept = 0, slope = 1), linetype = "dashed") +  # Y=X line
  scale_fill_gradient(trans = 'log10', name = "Log10\nFrequency", low = "white", high = "darkblue") +  # Color gradient for frequency
  facet_grid(Scorer.x ~ Scorer.y, switch = "both", scales = "free") +  # Facet setup
  theme_bw() +
  labs(x = "Scorer 2", y = "Scorer 1") +  # Labels
  theme(panel.spacing = unit(0.25, "lines"),  # Panel spacing
        strip.placement = "outside",  # Facet label placement
        panel.grid.major = element_blank(),  # Remove major grid lines
        panel.grid.minor = element_blank(),  # Remove minor grid lines
        panel.background = element_blank())  # Remove panel background

#dev.off()

```

### Plot 3: Bubble plot - x = scorer (discrete), y = score (discrete), circle size = frequency (discrete), plot points with jitter.
```{r fig.width=10, fig.height=15}
#| fig-width: 10
#| fig-height: 15
score_data_long_freq <- score_data_long %>%
  group_by(Scorer, Score) %>%
  summarise(Count = n(), .groups = 'drop')

score_data_long_freq

ggplot(score_data_long_freq, aes(x = Scorer, y = Score)) +
  geom_point(aes(size = Count, color = Count), stat = "identity") +  # Use actual count for size and color
  geom_text(aes(label = Count), vjust = 4, color = "black", size = 5) +  # Add labels for counts
  scale_size_continuous(name = "Frequency", range = c(0.06, 25)) +  # Control size of points based on Count
  scale_color_gradient(name = "Frequency", low = "blue", high = "red") +  # Apply color gradient based on Count
  labs(x = "Scorer", y = "Score") +  # Axis labels
  theme_minimal()
```

### Plot 4: Heatmap - x = scorer (discrete), y = score (discrete), circle size = frequency (discrete), plot points with jitter.

```{r fig.width=10, fig.height=10}
#| fig-width: 10
#| fig-height: 10
ggplot(score_data_long_freq, aes(x = Scorer, y = Score, fill = Count)) +
  geom_tile(color="black") +
  geom_text(aes(label = Count), color = "black", size = 5, vjust = 0.5) +
  scale_fill_gradient(name = "Frequency", low = "white", high = "red") +
  labs(x = "Scorer", y = "Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 0.5, hjust=1))
```

### Plot 5: Violin plot - x = scorer, y = score, violin and points with jitter
```{r fig.width=10, fig.height=10}
#| fig-width: 10
#| fig-height: 10
score_data_filtered <- score_data_long %>%
  filter(!grepl("\\[\\]", Score))
score_data_filtered$Score <- as.integer(score_data_filtered$Score)

ggplot(score_data_filtered, aes(x = Scorer, y = Score)) +
  geom_jitter(color = "darkgreen", alpha = 0.5, width = 0.2, height = 0.1) +  # Add jittered points with custom color, transparency, and width
  geom_violin(fill = "palegreen1", color = "black", alpha = 0.5) +  # Add violin plot with custom fill, border color, and transparency
  labs(x = "Scorer", y = "Score") +  # Axis labels
  theme_minimal()
```

### Plot 6: Heatmap of % agreement between scorers
```{r}
filtered_data <- pairwise_comparisons %>%
  filter(!is.na(Score.x) & !is.na(Score.y))

filtered_copy <- filtered_data
names(filtered_copy)[names(filtered_copy) %in% c("Scorer.x", "Scorer.y")] <- c("Scorer.y", "Scorer.x")

filtered_data <- rbind(filtered_data, filtered_copy)

agreement_data <- filtered_data %>%
  group_by(Scorer.x, Scorer.y) %>%
  summarise(Total = sum(Count),
            Agree = sum(Count * (Score.x == Score.y)),
            Percent_Agree = (Agree / Total) * 100,
            .groups = 'drop')

agreement_data <- agreement_data[as.integer(agreement_data$Scorer.x) < as.integer(agreement_data$Scorer.y), ]
```

```{r}
# Create a data frame with all unique combinations of Scorer.x and Scorer.y
all_combinations <- expand.grid(Scorer.x = unique(agreement_data$Scorer.x), Scorer.y = unique(agreement_data$Scorer.y))

# Merge with the original data frame using a left join
merged_df <- all_combinations %>%
  left_join(agreement_data, by = c("Scorer.x", "Scorer.y")) %>%
  mutate(Total = ifelse(is.na(Total), 0, Total),  # Replace NA values in Total column with 0
         Agree = ifelse(is.na(Agree), NA, Agree),  # Replace NA values in Agree column with NA
         Percent_Agree = ifelse(is.na(Percent_Agree), NA, Percent_Agree)) %>%  # Replace NA values in Percent_Agree column with NA
  filter(as.integer(Scorer.x) < as.integer(Scorer.y))
```

```{r fig.width=11, fig.height=10}
#| fig-width: 11
#| fig-height: 10

#pathout <- here("analyses", "02_score_distributions", "pairwiseagree.png")

#png(pathout, height=1000, width=1150)
#ggplot(merged_df, aes(x = Scorer.x, y = factor(Scorer.y, levels = as.character(sort(unique(as.integer(Scorer.y))))), fill = Percent_Agree)) +
#  geom_tile(color = "black", size = 1) +
#  geom_text(aes(label = ifelse(is.na(Percent_Agree), "", sprintf("%.1f%%", Percent_Agree))), color = "white", size = 8, vjust = 0.5, family = "sans", fontface = "bold") +
#  scale_fill_gradient(low = "blue", high = "red", name = "Percentage Agreement", na.value = "white") +
#  labs(x = "Scorer X", y = "Scorer Y") +
#  scale_x_discrete(limits = 1:10) +
#  scale_y_discrete(limits = 1:10) +
#  coord_cartesian(xlim = c(1, 10), ylim = c(1, 10)) +
#  theme_minimal() +
#  theme(panel.grid.major = element_blank(),
#        panel.grid.minor = element_blank(),
#        panel.background = element_blank(),
#        legend.key.size = unit(2, "lines"),
#        text = element_text(size=20),
#        axis.text.x = element_text(size = 20),
#        axis.text.y = element_text(size = 20))
#dev.off()

pathout <- here("analyses", "02_score_distributions", "pairwiseagree.png")
ggplot(merged_df, aes(x = Scorer.x, y = factor(Scorer.y, levels = as.character(sort(unique(as.integer(Scorer.y))))), fill = Percent_Agree)) +
  geom_tile(color = "black", size = 1) +  # Use tiles to create the heatmap
  geom_text(aes(label = ifelse(is.na(Percent_Agree), "", sprintf("%.1f%%", Percent_Agree))), color = "white", size = 8, vjust = 0.5, family = "sans", fontface = "bold") +
  scale_fill_gradient(low = "blue", high = "red", name = "Percentage Agreement", na.value = "white") +
  labs(x = "Scorer X", y = "Scorer Y") +
  scale_x_discrete(limits = 1:10) +
  scale_y_discrete(limits = 1:10) +
  coord_cartesian(xlim = c(1, 10), ylim = c(1, 10)) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        legend.key.size = unit(2, "lines"),
        text = element_text(size=20),
        axis.text.x = element_text(size = 20),
        axis.text.y = element_text(size = 20))
dev.off()
```

### Plot 7:

### Plot 8: Cumulative frequency - x = score, y = frequency, line colour = scorer

```{r fig.width=10, fig.height=10}
#| fig-width: 10
#| fig-height: 10
# Calculate cumulative frequency for each scorer
cumulative_data <- score_data_filtered %>%
  group_by(Scorer, Score) %>%
  summarise(Freq = n())

cumulative_data <- cumulative_data %>%
  group_by(Scorer) %>%
  mutate(Cum_Freq = cumsum(Freq)) %>%
  ungroup()

ggplot(cumulative_data, aes(x = Score, y = Cum_Freq, color = Scorer, group = Scorer)) +
  geom_line(size = 2, alpha = 0.7) +
  labs(x = "Score", y = "Cumulative Frequency", color = "Scorer") +  # Axis and legend labels
  theme_minimal()
```

### Plot 9:
```{r fig.width=10, fig.height=10}
#| fig-width: 10
#| fig-height: 10
score_data_filtered <- score_data_long %>%
  filter(!grepl("\\[\\]", Score))
score_data_filtered$Score <- as.integer(score_data_filtered$Score)

score_data_filtered_with_distance2 <- score_data_filtered %>%
  mutate(Distance_to_Median = abs(Score - Median_Score))

# Group by Scorer and Distance_to_Median and calculate the count of occurrences
score_data_filtered_freq2 <- score_data_filtered_with_distance2 %>%
  group_by(Scorer, Distance_to_Median) %>%
  summarise(Count = n(), .groups = 'drop')

# Plot the bubble plot with Distance_to_Median on the y-axis
ggplot(score_data_filtered_freq2, aes(x = Scorer, y = Distance_to_Median)) +
  geom_point(aes(size = Count, color = Count), stat = "identity") +  # Use actual count for size and color
  geom_text(aes(label = Count), vjust = 4, color = "black", size = 5) +  # Add labels for counts
  scale_size_continuous(name = "Frequency", range = c(3, 25)) +  # Control size of points based on Count
  scale_color_gradient(name = "Frequency", low = "blue", high = "red") +  # Apply color gradient based on Count
  labs(x = "Scorer", y = "Distance to Median", color = "Frequency") +  # Axis and legend labels
  theme_minimal() + 
  scale_y_continuous(breaks = unique(score_data_filtered_with_distance2$Distance_to_Median), minor_breaks = NULL)
```


### Plot 10:

### Plot 11: Scatter - x = scorer, y = dist from median
```{r fig.width=10, fig.height=15}
#| fig-width: 10
#| fig-height: 15
score_data_filtered_with_distance <- score_data_filtered %>%
  mutate(Distance_to_Median = Score - Median_Score)

# Group by Scorer and Distance_to_Median and calculate the count of occurrences
score_data_filtered_freq <- score_data_filtered_with_distance %>%
  group_by(Scorer, Distance_to_Median) %>%
  summarise(Count = n(), .groups = 'drop')

# Plot the bubble plot with Distance_to_Median on the y-axis
ggplot(score_data_filtered_freq, aes(x = Scorer, y = Distance_to_Median)) +
  geom_point(aes(size = Count, color = Count), stat = "identity") +  # Use actual count for size and color
  geom_text(aes(label = Count), vjust = 4, color = "black", size = 5) +  # Add labels for counts
  scale_size_continuous(name = "Frequency", range = c(3, 25)) +  # Control size of points based on Count
  scale_color_gradient(name = "Frequency", low = "blue", high = "red") +  # Apply color gradient based on Count
  labs(x = "Scorer", y = "Distance to Median", color = "Frequency") +  # Axis and legend labels
  theme_minimal() + 
  scale_y_continuous(breaks = unique(score_data_filtered_with_distance$Distance_to_Median), minor_breaks = NULL)
```

### Plot 11B:
```{r fig.width=10, fig.height=15}
#| fig-width: 10
#| fig-height: 15
ggplot(score_data_filtered_freq, aes(x = Scorer, y = Distance_to_Median, fill = Count)) +
  geom_tile(color="black") +
  geom_text(aes(label = Count), color = "black", size = 5, vjust = 0.5) +
  scale_fill_gradient(name = "Frequency", low = "white", high = "red") +
  labs(x = "Scorer", y = "Distance to Median", fill = "Frequency") +
  theme_minimal() + 
  theme(axis.text.x = element_text(vjust = 0.5, hjust = 1))  
```

### Plot 11C:
```{r fig.width=10, fig.height=15}
#| fig-width: 10
#| fig-height: 15
# Group by Median and Scorer and calculate the count of occurrences
score_data_filtered_freq3 <- score_data_filtered_with_distance %>%
  group_by(Median_Score, Distance_to_Median) %>%
  summarise(Count = n(), .groups = 'drop')

# Generate the bubble plot with median on the x-axis
ggplot(score_data_filtered_freq3, aes(x = Median_Score, y = Distance_to_Median)) +
  geom_point(aes(size = Count, color = Count), stat = "identity") +  # Use actual count for size and color
  geom_text(aes(label = Count), vjust = 4, color = "black", size = 5) +  # Add labels for counts
  scale_size_continuous(name = "Frequency", range = c(3, 25)) +  # Control size of points based on Count
  scale_color_gradient(name = "Frequency", low = "blue", high = "red") +  # Apply color gradient based on Count
  labs(x = "Median", y = "Scorer", color = "Frequency") +  # Axis and legend labels
  theme_minimal() + 
  scale_x_continuous(breaks = unique(score_data_filtered_with_distance$Median_Score), minor_breaks = NULL)
```

### Plot 11D:
```{r fig.width=10, fig.height=10}
#| fig-width: 10
#| fig-height: 10
# Group by Median and Scorer and calculate the count of occurrences
score_data_filtered_freq4 <- score_data_filtered_with_distance2 %>%
  group_by(Median_Score, Distance_to_Median) %>%
  summarise(Count = n(), .groups = 'drop')

# Generate the bubble plot with median on the x-axis
ggplot(score_data_filtered_freq4, aes(x = Median_Score, y = Distance_to_Median)) +
  geom_point(aes(size = Count, color = Count), stat = "identity") +  # Use actual count for size and color
  geom_text(aes(label = Count), vjust = 4, color = "black", size = 5) +  # Add labels for counts
  scale_size_continuous(name = "Frequency", range = c(3, 25)) +  # Control size of points based on Count
  scale_color_gradient(name = "Frequency", low = "blue", high = "red") +  # Apply color gradient based on Count
  labs(x = "Median", y = "Scorer", color = "Frequency") +  # Axis and legend labels
  theme_minimal() + 
  scale_x_continuous(breaks = unique(score_data_filtered_with_distance$Median_Score), minor_breaks = NULL)
```

### Plot 12: Cumulative frequency - x = abs diff from median, y = cumulative frequency, line colour = scorer
