---
title: "EDA Plots No 0"
format:
  html:
    code-fold: true
---
 
Date: 14/5/2024

## GOHREP
**Goal:**

Repeat preivous EDA but removing any CDAs with a median score of 0 (the large majority) to look for patterns amongst the other scores.

**Hypothesis:**

Removing the 0s will allow patterns to be seen more clearly amongst the remaining scores.

**Rationale:**

NA

**Experimental plan:**

1. Copy the code from the original EDA, but add a line to remove Median_Score == 0
2. Remove any plots for which this doesn't make sense e.g. those split by Median_Score

### Load required packages.

```{r}
library(readr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(here)
library(RColorBrewer)
```

### Load in the data. Create combined table. Anonymise scorers.

```{r}
path1 <- here("analyses", "02_score_distributions", "output_combined.csv")
path2 <- here("analyses", "02_score_distributions", "median_combined.csv")
path3 <- here("analyses", "02_score_distributions", "scorer_key.csv")

output_data <- read_csv(path1, show_col_types = FALSE)
median_data <- read_csv(path2, show_col_types = FALSE)
scorer_key <- read_csv(path3, show_col_types = FALSE, col_types = cols(ID = col_character(), Scorer_Name = col_character()))

score_data <- output_data %>%
  select(Basename, Row, Col, Pos, Scorer1, Score1, Scorer2, Score2, Scorer3, Score3) %>%
  bind_cols(
      median_data %>%
        select(Median_Score)
  )

id_lookup <- setNames(scorer_key$ID, scorer_key$Scorer_Name)

score_data <- score_data %>%
  mutate(across(c(Scorer1, Scorer2, Scorer3), ~ id_lookup[.]))

score_data_no0 <- score_data %>%
  filter(Median_Score != 0)

head(score_data_no0)
```

### Create a long table with one scorer and one score in each row.

```{r}
gen_long <- function(score_data){
  scorer_data <- score_data %>%
    select(starts_with("Scorer")) %>%
    pivot_longer(cols = everything(), names_to = NULL, values_to = "Scorer") %>%
    filter(!is.na(Scorer))

  score_values <- score_data %>%
    select(starts_with("Score"), -starts_with("Scorer")) %>%
    pivot_longer(cols = everything(), names_to = NULL, values_to = "Score") %>%
    filter(!is.na(Score))

  additional_data <- score_data %>%
    select(Basename, Row, Col, Pos, Median_Score) %>%
    slice(rep(row_number(), each = 3))

  score_data_long <- bind_cols(scorer_data, score_values, additional_data)

  score_data_long$Scorer <- factor(score_data_long$Scorer, levels = sort(as.numeric(unique(score_data_long$Scorer))))
  return(score_data_long)
}

score_data_long <- gen_long(score_data)
score_data_long_no0 <- gen_long(score_data_no0)
```

### What were the frequencies of severity scores collected by each scorer?

```{r fig.width=10, fig.height=12}
#| fig-width: 10
#| fig-height: 12
score_data_long_freq <- score_data_long %>%
  group_by(Scorer, Score) %>%
  summarise(Count = n(), .groups = 'drop')

my_palette <- colorRampPalette(c("lightgoldenrod1", "red"))(431)

ggplot(score_data_long_freq, aes(x = Scorer, y = Score)) +
  geom_point(aes(size = Count, color = Count), stat = "identity") +
  geom_text(aes(label = Count), vjust = 4, color = "black", size = 5) +
  scale_size_continuous(name = "Count", range = c(0.06, 25)) +
  scale_colour_gradientn(colours = my_palette) + 
  labs(x = "Scorer", y = "Score") +
  theme_minimal()
```

What does this plot show?

- After excluding 0s, the pattens of score frequencies stand out more.

Hypothesis forming:

1. N/A

### What were the percentages of scores collected by each scorer?

```{r fig.width=11, fig.height=10}
#| fig-width: 10
#| fig-height: 11

score_data_long_prop <- score_data_long_freq %>%
  group_by(Scorer) %>%
  mutate(total_count = sum(Count))

score_data_long_prop <- score_data_long_prop %>%
  mutate(proportion = Count / total_count)

my_palette <- colorRampPalette(c("lightgoldenrod1", "red"))(345)

ggplot(score_data_long_prop, aes(x = Scorer, y = Score, fill = proportion)) +
  geom_tile(color = "black") +
  geom_text(aes(label = sprintf("%.1f%%", proportion * 100)), color = "black", size = 5, vjust = 0.5) +
  scale_fill_gradientn(name = "Percentage", colours = my_palette) +
  labs(x = "Scorer", y = "Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 0.5, hjust = 1))
```

What does this plot show?

- This highlights the differences in percentages of scores collected by each scorer.
- Scorers 1 and 5 both scored a high proportion of 6s, and a low proportion of 1s and 2s.
- Scorer 10 scored low proportion of 6s and a high proportion of 4s.
- Despite CDAs of median 0 being removed, Scorer 4 scored 0 5.1% of the time.
- Scorer 5 did not score any 0s when the Median was not 0.

Hypothesis forming:

1. N/A

### Was there a difference in total scores collected by scorers?

```{r fig.width=7, fig.height=10}
#| fig-width: 7
#| fig-height: 10

# Calculate cumulative frequency for each scorer
score_data_filtered <- score_data_long %>%
  filter(!grepl("\\[\\]", Score))

score_data_filtered$Score <- as.integer(score_data_filtered$Score)

cumulative_data <- score_data_filtered %>%
  group_by(Scorer, Score) %>%
  summarise(Freq = n())

cumulative_data <- cumulative_data %>%
  group_by(Scorer) %>%
  mutate(Cum_Freq = cumsum(Freq)) %>%
  ungroup()

cumulative_data$Score <- factor(cumulative_data$Score)

total_cumulative <- cumulative_data %>%
  filter(Score == 6) %>%
  group_by(Scorer) %>%
  summarize(total_cumulative = max(Cum_Freq))

total_cumulative

# Plot
ggplot(cumulative_data, aes(x = Score, y = Cum_Freq, color = Scorer, group = Scorer)) +
  geom_line(size = 2) +
  geom_text(data = total_cumulative, aes(label = total_cumulative, x = 7.5, y = total_cumulative), vjust = 0, hjust = 1.2, color = "black", size = 3) +
  labs(x = "Score", y = "Cumulative Frequency", color = "Scorer") +
  theme_minimal() +
  scale_x_discrete(breaks = unique(cumulative_data$Score)) +
  scale_color_brewer(palette = "Set3")

```

As a table:

| Scorer | Frequency of Non-Zero Scores Collected |
|--------|---------------------------------------|
|    1   |                 648                   |
|    2   |                1408                   |
|    3   |                1207                   |
|    4   |                1339                   |
|    5   |                 529                   |
|    6   |                1174                   |
|    7   |                 580                   |
|    8   |                1275                   |
|    9   |                1356                   |
|   10   |                1355                   |


What does this plot show?

- A similar trend shows here to the previous plot including median 0s: Scorers 1, 5, and 7 collected fewer scores.
- However, it is interesting that after removing 0s, the number of scores collected by scorer 5 is much closer to that of scorers 1 and 7 compared to previously, highlighting did not score as many images that received a median score of 0.

Hypothesis forming:

1. N/A


### What was the percentage agreement between each pair of scorers?

```{r}

gen_merged_df <- function(score_data_long){
  # Assuming 'data' is your dataframe
  data <- score_data_long %>%
    mutate(Score = as.numeric(Score)) %>%
    group_by(Basename, Row, Col, Pos) %>%
    mutate(id = row_number()) %>%
    ungroup()

  # Assuming 'data' is your dataframe and has been prepared as mentioned earlier
  pairwise_comparisons <- data %>%
    full_join(data, by = c("Basename", "Row", "Col", "Pos")) %>%
    filter(id.x < id.y) %>%
    select(Scorer.x = Scorer.x, Score.x = Score.x, Scorer.y = Scorer.y, Score.y = Score.y) %>%
    group_by(Scorer.x, Score.x, Scorer.y, Score.y) %>%
    summarise(Count = n(), .groups = 'drop')

  filtered_data <- pairwise_comparisons %>%
    filter(!is.na(Score.x) & !is.na(Score.y))

  filtered_copy <- filtered_data
  names(filtered_copy)[names(filtered_copy) %in% c("Scorer.x", "Scorer.y")] <- c("Scorer.y", "Scorer.x")

  filtered_data <- rbind(filtered_data, filtered_copy)

  agreement_data <- filtered_data %>%
    group_by(Scorer.x, Scorer.y) %>%
    summarise(Total = sum(Count),
              Agree = sum(Count * (Score.x == Score.y)),
              Percent_Agree = (Agree / Total) * 100,
              .groups = 'drop')

  agreement_data <- agreement_data[as.integer(agreement_data$Scorer.x) < as.integer(agreement_data$Scorer.y), ]

  all_combinations <- expand.grid(Scorer.x = unique(agreement_data$Scorer.x), Scorer.y = unique(agreement_data$Scorer.y))

  # Merge with the original data frame using a left join
  merged_df <- all_combinations %>%
    left_join(agreement_data, by = c("Scorer.x", "Scorer.y")) %>%
    mutate(Total = ifelse(is.na(Total), 0, Total),  # Replace NA values in Total column with 0
          Agree = ifelse(is.na(Agree), NA, Agree),  # Replace NA values in Agree column with NA
          Percent_Agree = ifelse(is.na(Percent_Agree), NA, Percent_Agree)) %>%  # Replace NA values in Percent_Agree column with NA
    filter(as.integer(Scorer.x) < as.integer(Scorer.y))

  return(merged_df)
}

merged_df <- gen_merged_df(score_data_long)
merged_df_no0 <- gen_merged_df(score_data_long_no0)
```

```{r}
print(merged_df)
print(merged_df_no0)
```

```{r fig.width=11, fig.height=10}
#| fig-width: 11
#| fig-height: 10

min_value <- min(c(merged_df$Percent_Agree, merged_df_no0$Percent_Agree), na.rm = TRUE)
max_value <- max(c(merged_df$Percent_Agree, merged_df_no0$Percent_Agree), na.rm = TRUE)


# With 0
pathout <- here("analyses", "02_score_distributions", "pairwiseagree_norm.png")
png(pathout, height=1000, width=1150)
ggplot(merged_df, aes(x = Scorer.x, y = factor(Scorer.y, levels = as.character(sort(unique(as.integer(Scorer.y))))), fill = Percent_Agree)) +
  geom_tile(color = "black", size = 1) +  # Use tiles to create the heatmap
  geom_text(aes(label = ifelse(is.na(Percent_Agree), "", sprintf("%.1f%%", Percent_Agree))), color = "white", size = 8, vjust = 0.5, family = "sans", fontface = "bold") +
  scale_fill_gradient(low = "blue", high = "red", name = "Percentage Agreement", na.value = "white", limits = c(min_value, max_value)) +
  labs(x = "Scorer X", y = "Scorer Y") +
  scale_x_discrete(limits = 1:10) +
  scale_y_discrete(limits = 1:10) +
  coord_cartesian(xlim = c(1, 10), ylim = c(1, 10)) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        legend.key.size = unit(2, "lines"),
        text = element_text(size=20),
        axis.text.x = element_text(size = 20),
        axis.text.y = element_text(size = 20))
dev.off()
```

```{r}
# No 0
pathout <- here("analyses", "02_score_distributions", "pairwiseagreeno0_norm.png")
png(pathout, height=1000, width=1150)
ggplot(merged_df_no0, aes(x = Scorer.x, y = factor(Scorer.y, levels = as.character(sort(unique(as.integer(Scorer.y))))), fill = Percent_Agree)) +
  geom_tile(color = "black", size = 1) +
  geom_text(aes(label = ifelse(is.na(Percent_Agree), "", sprintf("%.1f%%", Percent_Agree))), color = "white", size = 8, vjust = 0.5, family = "sans", fontface = "bold") +
  scale_fill_gradient(low = "blue", high = "red", name = "Percentage Agreement", na.value = "white", limits = c(min_value, max_value)) +
  labs(x = "Scorer X", y = "Scorer Y") +
  scale_x_discrete(limits = 1:10) +
  scale_y_discrete(limits = 1:10) +
  coord_cartesian(xlim = c(1, 10), ylim = c(1, 10)) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        legend.key.size = unit(2, "lines"),
        text = element_text(size=20),
        axis.text.x = element_text(size = 20),
        axis.text.y = element_text(size = 20))
dev.off()
```

```{r}
subset_data <- merged_df %>%
  filter(Scorer.x == 5, Scorer.y == 6)
```


What does this plot show?

- The minimum agreement has decreased from 34.3% to 16.1% (between the same pair of scorers - Scorer 5 and Scorer 6).
- Without 0s, the maximum % agreement is 57.1% between Scorer 2 and Scorer 7, down from 76.6 (previously between Scorer 10 and Scorer 7).
- Scorer 10 and Scorer 8 is interesting, because with 0s they agreed 38.9% of the time and without there was only a 4.2% decrease to 34.7%. This means that they either both scored highly (and were unlikely to give out 0s), or they both scored relatively fewer 0-score CDAs than other pairs.
- The largest decrease in accuracy after removing 0s was between Scorer 4 and Scorer 1, with a decrease of 30.1% from 70.7% to 40.6%, indicating that the majority of agreement was on 0s, and agreement over other scores was far less likely.

Hypothesis forming:

1. N/A

### Agreement patterns between scorer pairs

```{r fig.width=11, fig.height=10}
#| fig-width: 11
#| fig-height: 10
filtered_data <- pairwise_comparisons %>%
  filter(!is.na(Score.x) & !is.na(Score.y))

filtered_copy <- filtered_data
names(filtered_copy)[names(filtered_copy) %in% c("Scorer.x", "Scorer.y")] <- c("Scorer.y", "Scorer.x")
names(filtered_copy)[names(filtered_copy) %in% c("Score.x", "Score.y")] <- c("Score.y", "Score.x")
filtered_data <- rbind(filtered_data, filtered_copy)

filtered_data <- filtered_data[as.integer(filtered_data$Scorer.x) <= as.integer(filtered_data$Scorer.y), ]

my_palette <- colorRampPalette(c("lightblue", "red"))(100)

filtered_data$Scorer.y <- factor(filtered_data$Scorer.y, levels = rev(levels(filtered_data$Scorer.y)))

```

```{r}
library(tidyr)
#pathout <- here("analyses", "02_score_distributions", "pair_patterns_no0.png")
#png(pathout, height=1000, width=1150)
levels_x <- levels_y <- as.character(1:10)
filtered_data <- filtered_data %>%
  mutate(Scorer.x = factor(Scorer.x, levels = levels_x),
         Scorer.y = factor(Scorer.y, levels = levels_y))
complete_data <- complete(filtered_data,
                          Scorer.x = factor(levels_x, levels = levels_x),
                          Scorer.y = factor(levels_y, levels = levels_y),
                          fill = list(Count = 0))
complete_data <- complete_data %>% mutate(Scorer.y = forcats::fct_rev(Scorer.y))
real_data <- complete_data %>% filter(Count > 0)

```

```{r}
pathout <- here("analyses", "02_score_distributions", "pair_patterns_no0.png")
png(pathout, height=1000, width=1150)
ggplot(complete_data, aes(x = Score.x, y = Score.y)) +
  geom_tile(aes(fill = Count), color = "black", size = 0.5) +
  geom_abline(data = real_data, aes(intercept = 0, slope = 1), linetype = "dashed") +
  scale_fill_gradientn(colours = my_palette, name = "Frequency") +
  facet_grid(Scorer.y ~ Scorer.x, switch = "both") +
  theme_bw() +
  labs(x = "Scorer X", y = "Scorer Y") + 
  theme(panel.spacing = unit(0.25, "lines"),
        strip.placement = "outside",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        legend.key.size = unit(2, "lines"),
        text = element_text(size=20),
        axis.text.x = element_text(size = 20),
        axis.text.y = element_text(size = 20)
        )
dev.off()
```

```{r}
subset_data <- filtered_data %>%
  filter(Scorer.x == 5, Scorer.y == 6)

pathsingle <- here("analyses", "02_score_distributions", "singletrend.png")
png(pathsingle, height=600, width=650)
ggplot(subset_data, aes(x = Score.x, y = Score.y)) +
  geom_tile(aes(fill = Count), color = "black", size = 0.5) +
  geom_abline(data = subset_data, aes(intercept = 0, slope = 1), linetype = "dashed") +
  scale_fill_gradientn(colours = my_palette, name = "Frequency") +
  facet_grid(Scorer.y ~ Scorer.x, switch = "both") +
  theme_bw() +
  labs(x = "Scorer X", y = "Scorer Y") + 
  theme(panel.spacing = unit(0.25, "lines"),
        strip.placement = "outside",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        legend.key.size = unit(2, "lines"),
        text = element_text(size=20),
        axis.text.x = element_text(size = 20),
        axis.text.y = element_text(size = 20)
        ) +
  scale_x_continuous(limits = c(-0.5, max(subset_data$Score.x) + 1), expand = c(0, 0)) +
  scale_y_continuous(limits = c(-0.5, max(subset_data$Score.y) + 1), expand = c(0, 0))
dev.off()


```

What does this plot show?

- Like previously, there are trends showing some scorers to score higher or lower than others, for example Scorer 5 typically scores higher than the median, and Scorer 4 often scores lower than the median.
- The most frequent pattern is still that of agreement though.

Hypothesis forming:

1. N/A