---
title: "Image information and input size selection"
format:
  html:
    code-fold: true
---

Date: 13/7/2026

## GOHREP

**Goal:**

1. Measure how much image information (Shannon entropy) each cropped CDA image
   retains as it is downsampled to progressively smaller side lengths, in order
   to justify a standardised input size for model training.
2. Characterise how Shannon entropy, the green/red colour balance, and the mean
   green level of a CDA relate to its severity score.

**Hypothesis:**

1. Shannon entropy stays roughly constant from the raw crop down to a moderate
   side length, then falls once the image becomes too small to hold the same
   range of pixel intensities. A side length just above that drop-off preserves
   almost all of the information while minimising the number of pixels (and
   therefore the training compute).
2. Under the UV-fluorescence imaging used here, healthy tissue autofluoresces
   red and cell death shifts the balance towards green, so the green/red ratio
   and the mean green level should both increase with severity score. Shannon
   entropy alone is expected to overlap heavily between scores.

**Rationale:**

Downsampling reduces the number of pixels and therefore the number of distinct
intensity values an image can represent, which caps its Shannon entropy. If
entropy is preserved down to a given side length, that size still carries the
information a model could use; shrinking further discards information. Locating
the side length at which entropy begins to fall gives a principled, information-
theoretic basis for choosing the input resolution (64 × 64 px) used elsewhere in
this project (see `07_visual_complexity`).

This analysis reproduces, for the full multi-set dataset, figures first made on
Set 1 alone during the Year 1 probationary review.

---

## 1. Setup

### 1.1 Load required packages

```{r}
#| output: false
# 1.1
library(tiff)
suppressMessages(library(EBImage)) # image resizing
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(readr)
library(stringr)
library(patchwork)
library(here)
library(RColorBrewer)
```

### 1.2 Helper functions and colour palettes

```{r}
# 1.2
source(here("src", "plotting_helpers.R"))

save_plot <- make_save_plot("10_image_entropy")
```

### 1.3 Configuration and metric functions

```{r}
# 1.3
# Cropped CDA images are stored one folder per median severity score (0-6);
# the folder name is the score. Each image is resized to a range of square
# side lengths and its Shannon entropy computed at every size. "Raw" is the
# original crop (variable native size, see 07_visual_complexity).
data_dir <- here("data", "cropped_images")
sizes <- c(128, 64, 32, 16, 8, 4)
size_levels <- c("Raw", as.character(sizes)) # x-axis order: large -> small

# Shannon entropy of an image, replicating the default behaviour of
# skimage.measure.shannon_entropy used in the original Year 1 analysis:
# histogram every value in the array (flattened across the RGB channels) as an
# 8-bit intensity, then H = -sum(p_i * log2(p_i)). More distinct, evenly used
# intensities -> higher entropy; the maximum is 8 bits.
shannon_entropy <- function(arr) {
  v <- as.integer(round(pmin(pmax(arr, 0), 1) * 255))
  counts <- tabulate(v + 1L, nbins = 256L)
  p <- counts[counts > 0] / length(v)
  -sum(p * log2(p))
}

# Per-image metrics: entropy at every side length, plus two colour metrics
# computed on the raw crop over non-dark pixels (max(R,G,B) >= 0.1, which
# excludes the black background and holes in the leaf, as in the original).
#   - log_gr: log of the green:red pixel-count ratio. Under this imaging
#     healthy tissue is red and cell death shifts towards green, so this is
#     expected to rise with score. A +1 pseudocount avoids log(0).
#   - mean_green: mean green-channel intensity (0-255) of the non-dark pixels.
image_metrics <- function(path) {
  img <- readTIFF(path, info = FALSE)
  if (length(dim(img)) == 2) img <- array(img, dim = c(dim(img), 3))
  img <- img[, , 1:3, drop = FALSE]

  ent <- c(Raw = shannon_entropy(img))
  for (s in sizes) {
    r <- pmin(pmax(EBImage::resize(img, w = s, h = s), 0), 1)
    ent[as.character(s)] <- shannon_entropy(r)
  }

  R <- img[, , 1]
  G <- img[, , 2]
  B <- img[, , 3]

  # Non-dark pixels (drops black background/holes) for the colour metrics.
  nondark <- pmax(R, G, B) >= 0.1
  n_red <- sum(R[nondark] > G[nondark])
  n_green <- sum(G[nondark] >= R[nondark])

  # Per-pixel brightness and two leaf-region definitions (see 5.5):
  #  - otsu_*: an image-adaptive two-class threshold (Otsu, from the image's
  #    own brightness histogram). It cleanly removes a dark background but, in
  #    high-score CDAs, tends to select the bright lesion rather than the whole
  #    leaf, so it is used only to illustrate that difficulty.
  #  - leaf_*: a low fixed cut that removes only the near-black background,
  #    keeping all tissue - a less score-dependent lighting proxy.
  bright_px <- (R + G + B) / 3
  thr <- EBImage::otsu(bright_px, range = c(0, 1))
  otsu_mask <- bright_px > thr
  if (sum(otsu_mask) < 0.02 * length(otsu_mask)) otsu_mask <- bright_px >= 0.1
  leaf_mask <- bright_px >= 0.10

  c(ent,
    log_gr = log((n_green + 1) / (n_red + 1)),
    mean_green = mean(G[nondark]) * 255,
    otsu_frac = mean(otsu_mask),
    otsu_bright = mean(bright_px[otsu_mask]) * 255,
    leaf_frac = mean(leaf_mask),
    leaf_bright = mean(bright_px[leaf_mask]) * 255
  )
}
```

---

## 2. Compute per-image metrics

The full dataset is a few thousand crops, each resized six times, so the
per-image metrics are computed once and cached to `entropy_cache/`. Delete that
file to force a recompute.

```{r}
# 2.1
cache_path <- here("analyses", "10_image_entropy", "entropy_cache", "cda_metrics.rds")

if (file.exists(cache_path)) {
  metrics <- readRDS(cache_path)
} else {
  scores <- list.dirs(data_dir, recursive = FALSE, full.names = FALSE)
  scores <- scores[grepl("^[0-9]+$", scores)]

  rows <- list()
  for (sc in scores) {
    files <- list.files(file.path(data_dir, sc),
      pattern = "\\.tif$",
      full.names = TRUE
    )
    for (f in files) {
      m <- image_metrics(f)
      rows[[length(rows) + 1]] <- tibble::tibble(
        image_id = tools::file_path_sans_ext(basename(f)),
        score = as.integer(sc),
        size = size_levels,
        entropy = as.numeric(m[size_levels]),
        log_gr = unname(m["log_gr"]),
        mean_green = unname(m["mean_green"]),
        otsu_frac = unname(m["otsu_frac"]),
        otsu_bright = unname(m["otsu_bright"]),
        leaf_frac = unname(m["leaf_frac"]),
        leaf_bright = unname(m["leaf_bright"])
      )
    }
    message(sprintf("score %s: %d images", sc, length(files)))
  }
  metrics <- dplyr::bind_rows(rows)
  saveRDS(metrics, cache_path)
}

# Long form (one row per image x size) for the size plots, and a per-image
# table (one row per image) for the score/colour plots.
metrics <- metrics %>%
  mutate(
    size = factor(size, levels = size_levels),
    score_f = factor(score, levels = 0:6)
  )

per_image <- metrics %>%
  filter(size == "Raw") %>%
  transmute(image_id, score, score_f,
    entropy_raw = entropy, log_gr, mean_green,
    otsu_frac, otsu_bright, leaf_frac, leaf_bright
  )

cat(sprintf(
  "Loaded metrics for %d CDA images across %d scores.\n",
  dplyr::n_distinct(metrics$image_id), dplyr::n_distinct(metrics$score)
))
per_image %>%
  count(score, name = "n") %>%
  as.data.frame()
```

---

## 3. Image information versus input size

Before comparing entropy across input sizes, the native sizes of the cropped CDA
images are summarised. CDAScorer stores each CDA as a square crop, but the side
length varies with the size of the annotated bounding box. The mean, variance and
range of the native side length are printed below, and their distribution is
shown as a histogram (dashed line: the 64 px training size).

```{r}
# 3.0.1  Native crop side lengths (square crops, so side = height = width). Cached.
side_cache <- here("analyses", "10_image_entropy", "entropy_cache", "side_lengths.rds")
if (file.exists(side_cache)) {
  side_lengths <- readRDS(side_cache)
} else {
  side_files <- list.files(data_dir,
    pattern = "\\.tif$", recursive = TRUE,
    full.names = TRUE
  )
  side_lengths <- vapply(
    side_files,
    function(f) dim(readTIFF(f, info = FALSE))[1], integer(1)
  )
  saveRDS(side_lengths, side_cache)
}
cat(sprintf(
  "Native crop side length (px): mean = %.1f, variance = %.1f, range = %d-%d (n = %d)\n",
  mean(side_lengths), var(side_lengths),
  min(side_lengths), max(side_lengths), length(side_lengths)
))
```

```{r fig.width=8, fig.height=4.5}
#| fig-width: 8
#| fig-height: 4.5
# 3.0.2  Histogram of native crop side lengths
p_sides <- ggplot(data.frame(side = side_lengths), aes(x = side)) +
  geom_histogram(binwidth = 20, fill = "grey60", colour = "white") +
  geom_vline(xintercept = 64, linetype = "dashed", colour = "#2166AC") +
  labs(x = "Native cropped-image side length (pixels)", y = "Number of CDAs") +
  theme_minimal() +
  theme(axis.text = element_text(size = 13), axis.title = element_text(size = 15))
p_sides
save_plot(p_sides, "crop_side_lengths.png", width = 8, height = 4.5, dir = "thesis_plots")
```

### 3.1 Entropy across image sizes (parallel coordinates)

Each faint line is one CDA, tracing its Shannon entropy as it is downsampled
from the raw crop to a 4 × 4 px thumbnail. The thick line is the per-size
median. Entropy is essentially flat from Raw down to 32 px, dips slightly at
16 px, then falls sharply at 8 and 4 px — the point at which images are too
small to hold their intensity range. 64 × 64 px sits comfortably on the flat
region.

```{r fig.width=8, fig.height=5}
#| fig-width: 8
#| fig-height: 5
# 3.1
size_median <- metrics %>%
  group_by(size) %>%
  summarise(entropy = median(entropy), .groups = "drop")

p_parallel <- ggplot(metrics, aes(x = size, y = entropy)) +
  geom_line(aes(group = image_id), alpha = 0.03, linewidth = 0.3) +
  geom_line(
    data = size_median, aes(group = 1), colour = "#2166AC",
    linewidth = 1.2
  ) +
  geom_point(data = size_median, colour = "#2166AC", size = 2.5) +
  labs(x = "CDA image side length in pixels", y = "Shannon entropy") +
  theme_minimal() +
  theme(
    axis.text  = element_text(size = 13),
    axis.title = element_text(size = 15)
  )
p_parallel
save_plot(p_parallel, "parallel_coords.png",
  width = 8, height = 5,
  dir = "thesis_plots"
)
```

### 3.2 Entropy distribution at each image size (ridge plot)

The same information as a set of stacked densities, from the raw crop (top) to
4 × 4 px (bottom). The distribution holds its shape and position down to 16 px,
then shifts left and collapses at 8 and 4 px.

```{r fig.width=8, fig.height=5}
#| fig-width: 8
#| fig-height: 5
# 3.2
# Manual ridgeline (the ggridges package is not available in this environment):
# a scaled kernel density per size, offset vertically by its row so the ridges
# overlap slightly, with Raw drawn at the top.
ridge_levels <- rev(size_levels) # 4 (bottom) ... Raw (top)
x_grid <- seq(min(metrics$entropy) - 0.3, max(metrics$entropy) + 0.3,
  length.out = 512
)

ridge_df <- metrics %>%
  group_by(size) %>%
  group_modify(~ {
    d <- density(.x$entropy, from = min(x_grid), to = max(x_grid), n = 512)
    tibble(x = d$x, dens = d$y)
  }) %>%
  ungroup() %>%
  group_by(size) %>%
  mutate(dens = dens / max(dens)) %>% # normalise each ridge to unit height
  ungroup() %>%
  mutate(
    row = match(as.character(size), ridge_levels) - 1,
    scale = 1.6, # overlap factor
    ymin = row,
    ymax = row + scale * dens
  )

# Mean entropy per size, placed on each ridge's baseline and joined into a line
# so the leftward shift at small sizes is visible directly on the ridges.
mean_df <- metrics %>%
  group_by(size) %>%
  summarise(mean_ent = mean(entropy), .groups = "drop") %>%
  mutate(row = match(as.character(size), ridge_levels) - 1)

p_ridge <- ggplot(ridge_df, aes(group = size)) +
  geom_ribbon(aes(x = x, ymin = ymin, ymax = ymax),
    fill = "grey75",
    colour = NA
  ) +
  geom_line(aes(x = x, y = ymax), colour = "black", linewidth = 0.5) +
  geom_line(
    data = mean_df, aes(x = mean_ent, y = row, group = 1),
    colour = "#2166AC", linewidth = 0.9
  ) +
  geom_point(
    data = mean_df, aes(x = mean_ent, y = row),
    colour = "#2166AC", size = 2.4
  ) +
  scale_y_continuous(
    breaks = seq_along(ridge_levels) - 1,
    labels = ridge_levels
  ) +
  labs(x = "Shannon entropy", y = "CDA image side length in pixels") +
  theme_minimal() +
  theme(
    axis.text  = element_text(size = 13),
    axis.title = element_text(size = 15),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )
p_ridge
save_plot(p_ridge, "ridge_plot.png", width = 8, height = 5, dir = "thesis_plots")
```

### 3.3 Fraction of raw entropy retained at each size

A numeric summary of the two plots above: the median entropy at each size, and
the fraction of the raw median it retains. 64 × 64 px keeps almost all of the
raw information; the loss only becomes substantial below 16 px.

```{r}
# 3.3
raw_median <- size_median$entropy[size_median$size == "Raw"]
size_median %>%
  mutate(fraction_of_raw = entropy / raw_median) %>%
  arrange(factor(size, levels = size_levels)) %>%
  mutate(
    entropy = round(entropy, 3),
    fraction_of_raw = round(fraction_of_raw, 3)
  ) %>%
  as.data.frame()
```

---

## 4. Image information versus severity score

### 4.1 Shannon entropy by score

Raw-crop entropy against severity score. Entropy rises slightly with score but
overlaps heavily between adjacent scores, so — as found in the Year 1 review —
Shannon entropy alone does not cleanly separate the severity classes.

```{r fig.width=8, fig.height=5}
#| fig-width: 8
#| fig-height: 5
# 4.1
p_ent_score <- ggplot(per_image, aes(x = score_f, y = entropy_raw)) +
  geom_boxplot(aes(fill = score_f), outlier.shape = NA, alpha = 0.85) +
  geom_jitter(width = 0.18, height = 0, alpha = 0.15, size = 0.5) +
  scale_fill_manual(name = "Severity Score", values = sev_pal) +
  labs(x = "Severity score", y = "Shannon entropy") +
  theme_minimal() +
  theme(
    axis.text    = element_text(size = 13),
    axis.title   = element_text(size = 15),
    legend.title = element_text(size = 13),
    legend.text  = element_text(size = 11)
  )
p_ent_score
save_plot(p_ent_score, "entropy_score_boxplots.png", width = 8, height = 5)
```

---

## 5. Colour versus severity score

### 5.1 Green/red ratio by score

The log green:red pixel-count ratio rises steadily with severity score,
consistent with cell death shifting the fluorescence balance from red (healthy
tissue) towards green.

```{r fig.width=8, fig.height=5}
#| fig-width: 8
#| fig-height: 5
# 5.1
p_ratio <- ggplot(per_image, aes(x = score_f, y = log_gr)) +
  geom_boxplot(aes(fill = score_f), outlier.shape = NA, alpha = 0.85) +
  geom_jitter(width = 0.18, height = 0, alpha = 0.15, size = 0.5) +
  scale_fill_manual(name = "Severity Score", values = sev_pal) +
  labs(x = "Severity score", y = "log(Green / Red ratio)") +
  theme_minimal() +
  theme(
    axis.text    = element_text(size = 13),
    axis.title   = element_text(size = 15),
    legend.title = element_text(size = 13),
    legend.text  = element_text(size = 11)
  )
p_ratio
save_plot(p_ratio, "ratio_score_boxplots.png", width = 8, height = 5)
```

### 5.2 Mean green level by score

The mean green-channel intensity of the non-dark pixels also increases with
severity score, giving a simpler, single-channel view of the same colour shift.

```{r fig.width=8, fig.height=5}
#| fig-width: 8
#| fig-height: 5
# 5.2
p_green <- ggplot(per_image, aes(x = score_f, y = mean_green)) +
  geom_boxplot(aes(fill = score_f), outlier.shape = NA, alpha = 0.85) +
  geom_jitter(width = 0.18, height = 0, alpha = 0.15, size = 0.5) +
  scale_fill_manual(name = "Severity Score", values = sev_pal) +
  labs(x = "Severity score", y = "Mean green intensity (0-255)") +
  theme_minimal() +
  theme(
    axis.text    = element_text(size = 13),
    axis.title   = element_text(size = 15),
    legend.title = element_text(size = 13),
    legend.text  = element_text(size = 11)
  )
p_green
save_plot(p_green, "green_score_boxplots.png", width = 8, height = 5)
```

### 5.3 Mean green level by project

The green level of a CDA depends not only on its severity but on the lighting
and exposure of the source image it was cropped from, which varies between
imaging sessions and projects. Here each point is one CDA and boxes are grouped
by project (projects ordered by median green): the between-project differences
in green level are a source of lighting variation that is unrelated to score.

```{r}
# 5.3.1
# Map each CDA to its source image and project. The cropped filenames are
# formatted <Basename>_<Row>_<Col>_<Pos>.tif, so stripping the trailing
# _Row_Col_Pos recovers the source-image basename (same approach as
# 07_visual_complexity). image_info.csv records the source Path once per folder
# (filled downwards); the top-level folder is the project, anonymised via the
# shared key.
image_info <- read_csv(
  here("analyses", "01_generate_raw_data", "01_allocation_info", "image_info.csv"),
  show_col_types = FALSE
) %>%
  fill(Path) %>%
  mutate(
    Project = str_split_fixed(Path, "/", 2)[, 1],
    stem = str_remove(Image, "[.](tif|TIF)$")
  )

project_key <- read_csv(
  here("anonymisation_keys", "project_anonymisation_key.csv"),
  show_col_types = FALSE
)
project_lut <- setNames(project_key$Anon_Label, project_key$Project)
stem_to_project <- setNames(project_lut[image_info$Project], image_info$stem)

per_image <- per_image %>%
  mutate(
    source  = str_remove(image_id, "_[0-9]+_[0-9]+_[0-9]+$"),
    project = unname(stem_to_project[source])
  )

n_unmapped <- sum(is.na(per_image$project))
cat(sprintf(
  "CDAs: %d | source images: %d | projects: %d | unmapped: %d\n",
  nrow(per_image), n_distinct(per_image$source),
  n_distinct(per_image$project), n_unmapped
))
```

```{r fig.width=8, fig.height=5}
#| fig-width: 8
#| fig-height: 5
# 5.3.2
project_order <- per_image %>%
  group_by(project) %>%
  summarise(med = median(mean_green), .groups = "drop") %>%
  arrange(med) %>%
  pull(project)

p_green_proj <- per_image %>%
  mutate(project = factor(project, levels = project_order)) %>%
  ggplot(aes(x = project, y = mean_green)) +
  geom_boxplot(aes(fill = project), outlier.shape = NA, alpha = 0.85) +
  geom_jitter(width = 0.18, height = 0, alpha = 0.12, size = 0.5) +
  scale_fill_brewer(palette = "Set1", guide = "none") +
  labs(x = "Project", y = "Mean green intensity (0-255)") +
  theme_minimal() +
  theme(
    axis.text  = element_text(size = 13),
    axis.title = element_text(size = 15)
  )
p_green_proj
save_plot(p_green_proj, "green_project_boxplots.png", width = 8, height = 5)
```

### 5.4 Lighting-normalised green level by score

To remove the between-project lighting offset seen in §5.3, each CDA's green
level is centred on its project mean and shifted back to the global mean
(`green_norm = green - project_mean + global_mean`). This corrects for
per-project brightness without touching the score differences that exist
*within* a project. The two panels below compare raw and project-normalised
green by score.

The correction is applied at project level rather than per source image on
purpose: an individual source image holds only ~30 CDAs whose scores are
correlated, so its own green mean partly encodes score, and normalising there
removes real signal (η² below). A project spans many source images and the full
score range, so its mean is a cleaner estimate of the lighting baseline.

```{r fig.width=11, fig.height=5}
#| fig-width: 11
#| fig-height: 5
# 5.4.1
global_mean_green <- mean(per_image$mean_green)

per_image <- per_image %>%
  group_by(project) %>%
  mutate(green_norm = mean_green - mean(mean_green) + global_mean_green) %>%
  ungroup() %>%
  group_by(source) %>%
  mutate(green_norm_source = mean_green - mean(mean_green) + global_mean_green) %>%
  ungroup()

# eta^2: fraction of green variance explained by score. Higher = scores more
# separable. Compared across raw green and the two normalisation levels.
eta_sq <- function(y, g) {
  grand <- mean(y)
  ss_between <- sum(tapply(y, g, function(v) length(v) * (mean(v) - grand)^2))
  ss_between / sum((y - grand)^2)
}
cat(sprintf(
  "eta^2 (green ~ score):  raw = %.3f | project-normalised = %.3f | source-normalised = %.3f\n",
  eta_sq(per_image$mean_green, per_image$score),
  eta_sq(per_image$green_norm, per_image$score),
  eta_sq(per_image$green_norm_source, per_image$score)
))
cat(sprintf(
  "eta^2 (green ~ project): %.3f  (share of green variance that is between-project)\n",
  eta_sq(per_image$mean_green, per_image$project)
))

green_panel <- function(yvar, title) {
  ggplot(per_image, aes(x = score_f, y = .data[[yvar]])) +
    geom_boxplot(aes(fill = score_f), outlier.shape = NA, alpha = 0.85) +
    geom_jitter(width = 0.18, height = 0, alpha = 0.12, size = 0.4) +
    scale_fill_manual(name = "Severity Score", values = sev_pal) +
    coord_cartesian(ylim = range(c(per_image$mean_green, per_image$green_norm))) +
    labs(title = title, x = "Severity score", y = "Mean green intensity (0-255)") +
    theme_minimal() +
    theme(
      axis.text = element_text(size = 12), axis.title = element_text(size = 14),
      plot.title = element_text(size = 14)
    )
}

green_norm_raw <- green_panel("mean_green", "Raw")
green_norm_proj <- green_panel("green_norm", "Project-normalised")
p_green_norm <- (green_norm_raw | green_norm_proj) + plot_layout(guides = "collect")
p_green_norm
save_plot(p_green_norm, "green_score_normalised_boxplots.png", width = 11, height = 5)
```

### 5.5 Isolating the leaf region to estimate lighting

The green normalisation in §5.4 uses green to estimate lighting, which is
circular. A cleaner lighting proxy is the **brightness** of the leaf itself
(mean of the R, G, B channels), measured only over leaf pixels so that the
black background — including holes and cell-death voids at the leaf edge — does
not drag the estimate down.

To decide which pixels are leaf, two thresholds on each CDA's own brightness
histogram are compared:

- **Otsu** — the image-adaptive threshold that maximises between-class variance,
  i.e. the natural split in the brightness distribution.
- **Leaf (background-only cut)** — a low fixed cut (brightness ≥ 0.1) that
  removes only the near-black background and keeps *all* tissue.

The panels below show, for one CDA per score, the original crop, the Otsu mask,
and the background-only leaf mask.

```{r fig.width=9, fig.height=11}
#| fig-width: 9
#| fig-height: 11
# 5.5.1
set.seed(7)
seg_examples <- lapply(c(0, 2, 4, 6), function(sc) {
  files <- list.files(file.path(data_dir, sc), pattern = "\\.tif$", full.names = TRUE)
  f <- sample(files, 1)
  img <- readTIFF(f, info = FALSE)
  if (length(dim(img)) == 2) img <- array(img, dim = c(dim(img), 3))
  img <- img[, , 1:3, drop = FALSE]
  bright_px <- (img[, , 1] + img[, , 2] + img[, , 3]) / 3
  thr <- EBImage::otsu(bright_px, range = c(0, 1))
  otsu_mask <- bright_px > thr
  if (sum(otsu_mask) < 0.02 * length(otsu_mask)) otsu_mask <- bright_px >= 0.1
  leaf_mask <- bright_px >= 0.10
  list(
    sc = sc, img = img, otsu = otsu_mask, leaf = leaf_mask,
    otsu_frac = mean(otsu_mask), leaf_frac = mean(leaf_mask)
  )
})

mask_img <- function(img, mask) {
  out <- img
  for (k in 1:3) {
    ch <- out[, , k]
    ch[!mask] <- 0
    out[, , k] <- ch
  }
  out
}
draw_seg <- function() {
  par(mfrow = c(length(seg_examples), 3), mar = c(0.5, 2, 2.2, 0.5))
  for (e in seg_examples) {
    plot(as.raster(e$img))
    title(sprintf("score %d: crop", e$sc))
    plot(as.raster(mask_img(e$img, e$otsu)))
    title(sprintf("Otsu region (%.0f%%)", 100 * e$otsu_frac))
    plot(as.raster(mask_img(e$img, e$leaf)))
    title(sprintf("leaf region (%.0f%%)", 100 * e$leaf_frac))
  }
}
png(plot_output_path("10_image_entropy", "supp_plots", "leaf_segmentation_examples.png"),
  width = 1500, height = 1800, res = 150
)
draw_seg()
invisible(dev.off())
draw_seg()
```

The Otsu mask cleanly removes the black background where there is one (scores 2,
4), but because healthy tissue fluoresces red (darker) while lesions are bright
yellow-green, in higher-score CDAs Otsu selects the **lesion** rather than the
whole leaf. Its foreground brightness therefore tracks score, making it a poor
lighting proxy; the background-only leaf brightness does not.

```{r}
# 5.5.2
# eta^2 (~ score) quantifies how much each brightness measure is driven by score
# rather than lighting. A good lighting proxy should be LOW.
cat(sprintf(
  "eta^2 (~ score):  Otsu brightness = %.3f  |  leaf brightness = %.3f\n",
  eta_sq(per_image$otsu_bright, per_image$score),
  eta_sq(per_image$leaf_bright, per_image$score)
))
cat(sprintf(
  "eta^2 (~ project): leaf brightness = %.3f  (more between-project variation than green's %.3f)\n",
  eta_sq(per_image$leaf_bright, per_image$project),
  eta_sq(per_image$mean_green, per_image$project)
))
```

Leaf brightness is far less score-driven than Otsu brightness and carries more
between-project (lighting) variation than green, so it is used as the lighting
baseline. Grouping it by project shows the same lighting differences seen for
green in §5.3.

```{r fig.width=8, fig.height=5}
#| fig-width: 8
#| fig-height: 5
# 5.5.3
lb_order <- per_image %>%
  group_by(project) %>%
  summarise(med = median(leaf_bright), .groups = "drop") %>%
  arrange(med) %>%
  pull(project)

p_leafbright_proj <- per_image %>%
  mutate(project = factor(project, levels = lb_order)) %>%
  ggplot(aes(x = project, y = leaf_bright)) +
  geom_boxplot(aes(fill = project), outlier.shape = NA, alpha = 0.85) +
  geom_jitter(width = 0.18, height = 0, alpha = 0.12, size = 0.5) +
  scale_fill_brewer(palette = "Set1", guide = "none") +
  labs(x = "Project", y = "Mean leaf brightness (0-255)") +
  theme_minimal() +
  theme(axis.text = element_text(size = 13), axis.title = element_text(size = 15))
p_leafbright_proj
save_plot(p_leafbright_proj, "leaf_brightness_project_boxplots.png", width = 8, height = 5)
```

### 5.6 Normalising green by leaf brightness

Each CDA's green is now divided by the leaf-brightness baseline of its source
image (or project), scaled to the global mean — a multiplicative correction for
per-source/-project lighting gain that, unlike §5.4, does not use green to
estimate the lighting. The panels compare raw green with the best correction
(per source image).

```{r fig.width=11, fig.height=5}
#| fig-width: 11
#| fig-height: 5
# 5.6.1
global_leaf <- mean(per_image$leaf_bright)

per_image <- per_image %>%
  group_by(source) %>%
  mutate(green_lb_source = mean_green * global_leaf / mean(leaf_bright)) %>%
  ungroup() %>%
  group_by(project) %>%
  mutate(green_lb_project = mean_green * global_leaf / mean(leaf_bright)) %>%
  ungroup()

cat("eta^2 (green ~ score), higher = more separable:\n")
cat(sprintf("  raw green                          = %.3f\n", eta_sq(per_image$mean_green, per_image$score)))
cat(sprintf("  green-mean normalised (project, 5.4) = %.3f\n", eta_sq(per_image$green_norm, per_image$score)))
cat(sprintf("  leaf-brightness normalised (project) = %.3f\n", eta_sq(per_image$green_lb_project, per_image$score)))
cat(sprintf("  leaf-brightness normalised (source)  = %.3f\n", eta_sq(per_image$green_lb_source, per_image$score)))

green_panel2 <- function(yvar, title) {
  ggplot(per_image, aes(x = score_f, y = .data[[yvar]])) +
    geom_boxplot(aes(fill = score_f), outlier.shape = NA, alpha = 0.85) +
    geom_jitter(width = 0.18, height = 0, alpha = 0.12, size = 0.4) +
    scale_fill_manual(name = "Severity Score", values = sev_pal) +
    coord_cartesian(ylim = range(c(per_image$mean_green, per_image$green_lb_source))) +
    labs(title = title, x = "Severity score", y = "Mean green intensity (0-255)") +
    theme_minimal() +
    theme(
      axis.text = element_text(size = 12), axis.title = element_text(size = 14),
      plot.title = element_text(size = 14)
    )
}

green_lb_raw <- green_panel2("mean_green", "Raw")
green_lb_norm <- green_panel2("green_lb_source", "Leaf-brightness normalised (per source)")
p_green_lb <- (green_lb_raw | green_lb_norm) + plot_layout(guides = "collect")
p_green_lb
save_plot(p_green_lb, "green_score_leafnorm_boxplots.png", width = 11, height = 5)
```

Correcting green for leaf brightness gives the largest improvement of any
normalisation tried (η² ≈ 0.52 → 0.58), confirming that between-source lighting
does add real noise. But the gain is modest and the adjacent-score boxes still
overlap heavily: even the best hand-crafted, lighting-corrected single feature
cannot separate neighbouring severity classes. This is the motivation for a
learned model (see `07_visual_complexity`) rather than hand-engineered features.

### 5.7 Green/red ratio versus entropy

The Year 1 review also asked whether entropy and colour carry the same
information. They do not: the green:red ratio and Shannon entropy are only
weakly related, so colour separates the scores (§5.1) where entropy alone does
not (§4.1).

```{r fig.width=8, fig.height=5}
#| fig-width: 8
#| fig-height: 5
# 5.7
r_gr_ent <- cor(per_image$log_gr, per_image$entropy_raw)

p_scatter <- ggplot(per_image, aes(x = entropy_raw, y = log_gr)) +
  geom_point(aes(colour = score_f), alpha = 0.3, size = 0.7) +
  scale_colour_manual(name = "Severity Score", values = sev_pal) +
  labs(
    x = "Shannon entropy", y = "log(Green / Red ratio)",
    subtitle = sprintf("Pearson r = %.2f", r_gr_ent)
  ) +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  theme_minimal() +
  theme(
    axis.text    = element_text(size = 13),
    axis.title   = element_text(size = 15),
    legend.title = element_text(size = 13),
    legend.text  = element_text(size = 11)
  )
p_scatter
save_plot(p_scatter, "ratio_entropy_scatter.png", width = 8, height = 5)
```

---

## 6. Interpretation and Discussion

- **Input size.** Shannon entropy is preserved from the raw crop down to about
  16 px and only falls substantially at 8 and 4 px (§3.1–3.3). 64 × 64 px
  therefore retains essentially all of the information in the raw crops while
  keeping the pixel count — and the training compute — low, supporting the
  64 × 64 input size used for model training (see `07_visual_complexity`).
- **Entropy and score.** Entropy overlaps heavily between adjacent scores
  (§4.1), so it is not a usable single feature for classifying severity.
- **Colour and score.** Both the green:red ratio (§5.1) and the mean green
  level (§5.2) increase monotonically with severity, and colour is only weakly
  correlated with entropy (§5.7) — colour carries score information that
  entropy does not.
- **Lighting is a real but only partly removable confound.** Green level varies
  between projects (§5.3; ~16% of green variance is between-project), confirming
  that imaging/lighting differences add noise on top of the score signal. Three
  intuition-led corrections were tried:
  - *Centring green on the project mean* (§5.4): removes the offset and modestly
    tightens the within-score spread, but the gain is small because green's
    score signal is partly correlated with which project a CDA is from.
  - *Source-image centring* (§5.4): over-corrects — a source image's own green
    mean partly encodes the scores of its ~30 CDAs — so it removes real signal.
  - *Segmenting the leaf and normalising green by leaf brightness* (§5.5–5.6):
    the best proxy, because leaf brightness is far less score-driven than an
    Otsu foreground (which latches onto the lesion) and carries more
    between-project variation than green. This gives the largest improvement
    (η² ≈ 0.52 → 0.58).
- **The features are not enough.** Even the best lighting-corrected single
  feature still leaves neighbouring severity classes heavily overlapped
  (§5.6). Hand-engineered colour/brightness/entropy features capture part of
  the severity signal but cannot cleanly separate adjacent scores — the
  motivation for a learned model over hand-crafted features.
```
