---
title: "Intra-Rater Reliability"
format:
  html:
    code-fold: true
---

Date: 2026-06-05

## GOHREP

**Goal:**

Quantify how consistently each rater scores the same CDA on two separate occasions (intra-rater reliability). Assess whether rescoring changes agreement with the original ground truth (median score). Investigate how rescoring disagreements would change ground truth labels (median score).

**Hypothesis:**

1. Raters will be more consistent with themselves on extreme scores (1, 6) and less consistent on intermediate scores (2–5).
2. Rescoring will not systematically improve or worsen agreement with the median — any changes will reflect genuine scoring uncertainty rather than learning.

**Rationale:**

Only non-zero median CDAs were included in the rescoring set, since 0s are almost always agreed upon. Each rater was presented the same CDAs they originally scored, in a randomised order, without access to their original scores.

---

## 1. Setup

### 1.1 Load required packages

```{r}
#| output: false
# 1.1
library(readr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)
library(patchwork)
library(scales)
library(here)
library(irr)
```

### 1.2 Load and anonymise rescore data

```{r}
# 1.2.1
source(here("src", "agreement_metrics.R"))
source(here("src", "plotting_helpers.R"))

# rescore_anon/ holds one file per rater, already anonymised (filename and
# `scorer` column both give the same ID) - see the anonymisation script under scripts/.
rescore_files <- list.files(
  here("analyses", "06_intra_rater_reliability", "rescore_anon"),
  pattern = "[.]csv$", full.names = TRUE
)

rescore_data <- bind_rows(lapply(rescore_files, read_csv, show_col_types = FALSE)) %>%
  mutate(
    Scorer        = scorer,
    old_score     = as.integer(old_score),
    new_score     = as.integer(new_score),
    median_score  = as.integer(median_score),
    score_change  = new_score - old_score
  ) %>%
  select(
    Scorer, basename, row, col, pos,
    old_score, new_score, median_score, score_change
  )

rescore_data$Scorer <- factor(rescore_data$Scorer,
  levels = sort(unique(rescore_data$Scorer))
)

cat("Raters:", paste(levels(rescore_data$Scorer), collapse = ", "), "\n")

cat("CDAs per rater:\n")
print(count(rescore_data, Scorer))

cat("\nRows with missing old_score (original score was '[]'):\n")
print(rescore_data %>% filter(is.na(old_score)) %>% count(Scorer))
```

```{r}
# 1.2.2
save_plot <- make_save_plot("06_intra_rater_reliability")
```

---

## 2. Summary statistics

```{r}
# 2.1
summary_stats <- rescore_data %>%
  group_by(Scorer) %>%
  summarise(
    N = n(),
    N_with_old = sum(!is.na(old_score)),
    intra_exact = mean(old_score == new_score, na.rm = TRUE) * 100,
    intra_within1 = mean(abs(old_score - new_score) <= 1, na.rm = TRUE) * 100,
    orig_vs_median = mean(old_score == median_score, na.rm = TRUE) * 100,
    rescore_vs_median = mean(new_score == median_score, na.rm = TRUE) * 100,
    agreement_change = rescore_vs_median - orig_vs_median,
    mean_abs_change = mean(abs(score_change), na.rm = TRUE),
    .groups = "drop"
  )

write_csv(
  summary_stats,
  here("analyses", "06_intra_rater_reliability", "intra_rater_summary.csv")
)

print(summary_stats)
```

```{r}
# 2.2
# Pooled across all rescored CDAs (all raters combined), the off-by-one rate
# is the near-miss rate (|old - new| <= 1) minus the exact rate (old = new):
# how often a rater's repeat score differed from their original by exactly
# one severity point.
intra_pooled_diffs <- rescore_data %>%
  filter(!is.na(old_score), !is.na(new_score)) %>%
  mutate(absdiff = abs(old_score - new_score))

cat(sprintf("Rescore comparisons:                      %d\n", nrow(intra_pooled_diffs)))
cat(sprintf("Exact self-agreement (diff = 0):          %.2f%%\n", mean(intra_pooled_diffs$absdiff == 0) * 100))
cat(sprintf("Near-miss self-agreement (diff <= 1):     %.2f%%\n", mean(intra_pooled_diffs$absdiff <= 1) * 100))
cat(sprintf("Off-by-one disagreement (diff == 1):      %.2f%%\n", mean(intra_pooled_diffs$absdiff == 1) * 100))
```

---

## 3. Score class distribution of rescored CDAs

```{r fig.width=12, fig.height=5}
#| fig-width: 12
#| fig-height: 5
# 3
scorer_levels <- levels(rescore_data$Scorer)

rescore_prop <- rescore_data %>%
  count(Scorer, Median_Score = median_score, name = "n") %>%
  group_by(Scorer) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  mutate(source = "Rescored sample")

original_alloc <- read_csv(
  here(
    "analyses", "01_generate_raw_data", "04_median_and_centre_coords",
    "combined_cda_data_median.csv"
  ),
  show_col_types = FALSE
) %>%
  select(Scorer1, Score1, Scorer2, Score2, Scorer3, Score3, Median_Score) %>%
  mutate(Median_Score = as.integer(Median_Score)) %>%
  filter(Median_Score > 0) %>%
  pivot_longer(c(Scorer1, Score1, Scorer2, Score2, Scorer3, Score3),
    names_to = c(".value", "num"),
    names_pattern = "(Scorer|Score)([0-9]+)"
  ) %>%
  filter(!is.na(Scorer)) %>%
  mutate(Scorer = as.integer(Scorer)) %>%
  filter(as.character(Scorer) %in% scorer_levels) %>%
  count(Scorer, Median_Score, name = "n") %>%
  group_by(Scorer) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  mutate(
    Scorer = factor(Scorer, levels = scorer_levels),
    source = "Original allocation"
  )

# Grey "Original" bar left of each score; coloured rescored bar (by severity) right. A single fill
# scale carries both: grey for the original allocation, YlOrRd for the rescored severity classes.
class_prop <- bind_rows(original_alloc, rescore_prop) %>%
  mutate(source = factor(source, levels = c("Original allocation", "Rescored sample"))) %>%
  complete(Scorer, Median_Score = 1:6, source, fill = list(prop = 0, n = 0)) %>%
  mutate(fillkey = if_else(source == "Original allocation",
    "Original", as.character(Median_Score)
  ))

fill_vals <- c("Original" = "grey65", sev_pal[as.character(1:6)])

p <- ggplot(class_prop, aes(x = factor(Median_Score), y = prop, fill = fillkey, group = source)) +
  geom_col(
    position = position_dodge(width = 0.9), width = 0.8,
    color = "black", linewidth = 0.3
  ) +
  geom_text(aes(label = scales::percent(prop, accuracy = 1)),
    position = position_dodge(width = 0.9), vjust = -0.35, size = 2.4
  ) +
  scale_fill_manual(
    values = fill_vals,
    breaks = c("Original", as.character(1:6)), name = NULL
  ) +
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.12))) +
  facet_wrap(~Scorer, nrow = 1, labeller = as_labeller(function(x) paste("Rater:", x))) +
  labs(x = "Median Severity Score", y = "Proportion of non-zero CDAs") +
  theme_minimal() +
  theme(
    panel.spacing = unit(0.5, "lines"),
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 14),
    strip.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.position = "right"
  )
p
save_plot(p, "rescore_class_distribution.png", width = 12, height = 5)
```

---

## 4. Intra-rater consistency: old vs new score

### 4.1 Exact and within-±1 intra-rater agreement per rater

```{r fig.width=7, fig.height=5}
#| fig-width: 7
#| fig-height: 5
# 4.1
intra_plot_data <- summary_stats %>%
  mutate(extra_within1 = intra_within1 - intra_exact) %>%
  pivot_longer(
    cols = c(intra_exact, extra_within1),
    names_to = "metric", values_to = "value"
  ) %>%
  mutate(metric = factor(metric,
    levels = c("extra_within1", "intra_exact"),
    labels = c("Near-miss agreement (±1)", "Exact agreement")
  ))

# Panel-wide means, drawn as black dashed reference lines matching the
# scorer_accuracy figure: the lower line is the mean exact self-agreement,
# the upper line the mean near-miss self-agreement (within +/-1). They are
# labelled on a secondary right-hand y-axis so they add no legend entries.
mean_exact <- mean(summary_stats$intra_exact) / 100
mean_within1 <- mean(summary_stats$intra_within1) / 100

p <- ggplot(intra_plot_data, aes(x = Scorer, y = value / 100, fill = metric)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = mean_exact, linetype = "dashed", colour = "black", linewidth = 0.6) +
  geom_hline(yintercept = mean_within1, linetype = "dashed", colour = "black", linewidth = 0.6) +
  scale_fill_manual(
    values = c(
      "Exact agreement" = "#238b45",
      "Near-miss agreement (±1)" = "#fec44f"
    ),
    name = NULL,
    breaks = c("Exact agreement", "Near-miss agreement (±1)")
  ) +
  scale_y_continuous(
    labels = scales::percent, limits = c(0, 1),
    sec.axis = sec_axis(~.,
      breaks = c(mean_exact, mean_within1),
      labels = c(
        sprintf("Mean exact\n(%.1f%%)", 100 * mean_exact),
        sprintf("Mean near-miss\n(%.1f%%)", 100 * mean_within1)
      )
    )
  ) +
  labs(x = "Rater", y = "Intra-rater agreement") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 13),
    axis.text.y.right = element_text(size = 11, hjust = 0),
    axis.title = element_text(size = 15),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 11),
    legend.position = "top"
  )
p
save_plot(p, "intra_rater_agreement.png", width = 7, height = 5, dir = "thesis_plots")
```

The per-rater values shown in panel A above, as a table (exact intra-rater
self-agreement, and total self-agreement within ±1 of the rater's original
score; n = rescored CDAs with a recoverable original score):

```{r}
# 4.1 (table)
intra_agreement_values <- summary_stats %>%
  arrange(Scorer) %>%
  transmute(
    Rater = as.character(Scorer),
    `Rescored (n)` = as.character(N_with_old),
    `Exact Agreement` = scales::percent(intra_exact / 100, accuracy = 0.1),
    `Near-miss (±1)` = scales::percent(intra_within1 / 100, accuracy = 0.1)
  )

mean_agreement_row <- tibble(
  Rater = "Mean",
  `Rescored (n)` = "",
  `Exact Agreement` = scales::percent(mean(summary_stats$intra_exact) / 100, accuracy = 0.1),
  `Near-miss (±1)` = scales::percent(mean(summary_stats$intra_within1) / 100, accuracy = 0.1)
)

bind_rows(intra_agreement_values, mean_agreement_row) %>%
  knitr::kable(caption = "Per-rater intra-rater self-agreement (panel A values)")
```

### 4.2 Intra-rater self-agreement by original score

```{r fig.width=12, fig.height=5}
#| fig-width: 12
#| fig-height: 5
# 4.2
self_agree <- rescore_data %>%
  filter(!is.na(old_score), !is.na(new_score)) %>%
  group_by(Scorer, old_score) %>%
  summarise(exact = mean(new_score == old_score), n = n(), .groups = "drop")

p <- ggplot(self_agree, aes(x = factor(old_score), y = exact, fill = factor(old_score))) +
  geom_col(color = "black", linewidth = 0.3) +
  geom_text(aes(label = scales::percent(exact, accuracy = 1)), vjust = -0.4, size = 3) +
  scale_fill_manual(values = sev_pal, name = "Original score") +
  scale_y_continuous(
    labels = scales::percent, limits = c(0, 1.05),
    expand = expansion(mult = c(0, 0.05))
  ) +
  facet_wrap(~Scorer, nrow = 1, labeller = as_labeller(function(x) paste("Rater:", x))) +
  labs(x = "Original Severity Score", y = "Self-agreement (exact)") +
  theme_minimal() +
  theme(
    panel.spacing = unit(0.5, "lines"),
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 14),
    strip.text = element_text(size = 12),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 11)
  )
p
save_plot(p, "intra_rater_self_agreement.png", width = 12, height = 5)
```

### 4.3 Confusion matrix of original vs rescore per rater

```{r fig.width=10, fig.height=7}
#| fig-width: 10
#| fig-height: 7
# 4.3
cm_data <- rescore_data %>%
  filter(!is.na(old_score), !is.na(new_score)) %>%
  count(Scorer, old_score, new_score) %>%
  complete(Scorer, old_score = 0:6, new_score = 0:6, fill = list(n = 0))

p <- ggplot(cm_data, aes(x = factor(old_score), y = factor(new_score), fill = n)) +
  geom_tile(color = "grey85", linewidth = 0.3) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey40") +
  geom_text(aes(label = ifelse(n > 0, n, "")), size = 2.5) +
  scale_fill_gradient(low = "white", high = "#238b45", name = "Count") +
  facet_wrap(~Scorer, ncol = 3, labeller = as_labeller(function(x) paste("Rater:", x))) +
  coord_equal() +
  labs(x = "Original", y = "Rescore") +
  theme_minimal() +
  theme(
    panel.spacing = unit(0.4, "lines"),
    panel.grid = element_blank(),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 13),
    strip.text = element_text(size = 11),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )
p
save_plot(p, "intra_rater_confusion_matrix.png", width = 10, height = 7)
```

---

## 5. Score change distribution

### 5.1 Distribution of new_score − old_score per rater

```{r fig.width=12, fig.height=5}
#| fig-width: 12
#| fig-height: 5
# 5.1
change_freq <- rescore_data %>%
  filter(!is.na(score_change)) %>%
  count(Scorer, score_change) %>%
  group_by(Scorer) %>%
  mutate(proportion = n / sum(n))

dir_pal <- c("positive" = "firebrick", "zero" = "grey60", "negative" = "#3182bd")

p <- ggplot(
  change_freq,
  aes(
    x = score_change, y = proportion,
    fill = case_when(
      score_change > 0 ~ "positive",
      score_change < 0 ~ "negative",
      TRUE ~ "zero"
    )
  )
) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.3) +
  scale_fill_manual(
    values = dir_pal, name = "Direction",
    labels = c("negative" = "Lower", "zero" = "Same", "positive" = "Higher")
  ) +
  scale_x_continuous(breaks = -6:6) +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~Scorer, nrow = 1, labeller = as_labeller(function(x) paste("Rater:", x))) +
  labs(x = "Score change (rescore − original)", y = "Proportion") +
  theme_minimal() +
  theme(
    panel.spacing = unit(0.5, "lines"),
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 14),
    strip.text = element_text(size = 12),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 11)
  )
p
save_plot(p, "score_change_distribution.png", width = 12, height = 5, dir = "thesis_plots")
```

The per-rater direction of change summarising panel B (proportion of rescores
that were higher, the same as, or lower than the rater's original score), and —
among the self-disagreements only (excluding unchanged scores) — the split
between scoring higher versus lower than their own previous score:

```{r}
# 5.1 (table)
change_dir <- rescore_data %>%
  filter(!is.na(score_change)) %>%
  group_by(Scorer) %>%
  summarise(
    n = n(),
    higher = mean(score_change > 0),
    same = mean(score_change == 0),
    lower = mean(score_change < 0),
    n_disagree = sum(score_change != 0),
    n_higher = sum(score_change > 0),
    n_lower = sum(score_change < 0),
    higher_of_disagree = sum(score_change > 0) / sum(score_change != 0),
    lower_of_disagree = sum(score_change < 0) / sum(score_change != 0),
    .groups = "drop"
  ) %>%
  arrange(Scorer)

change_dir %>%
  transmute(
    Rater = as.character(Scorer),
    `Rescored (n)` = n,
    Higher = scales::percent(higher, accuracy = 0.1),
    Same = scales::percent(same, accuracy = 0.1),
    Lower = scales::percent(lower, accuracy = 0.1),
    `Disagreements (n)` = n_disagree,
    `Higher (n)` = n_higher,
    `Higher | disagree` = scales::percent(higher_of_disagree, accuracy = 0.1),
    `Lower (n)` = n_lower,
    `Lower | disagree` = scales::percent(lower_of_disagree, accuracy = 0.1)
  ) %>%
  knitr::kable(caption = "Per-rater direction of rescore change (panel B values)")
```

---

## 6. Would rescoring change the ground truth label?

```{r}
# 6
canonical_scores <- read_csv(
  here(
    "analyses", "01_generate_raw_data", "04_median_and_centre_coords",
    "combined_cda_data_median.csv"
  ),
  show_col_types = FALSE
) %>%
  select(Basename, Row, Col, Pos, Scorer1, Score1, Scorer2, Score2, Scorer3, Score3) %>%
  mutate(
    across(c(Scorer1, Scorer2, Scorer3), as.integer),
    across(c(Score1, Score2, Score3), as.numeric)
  )

median_change <- rescore_data %>%
  filter(!is.na(old_score)) %>%
  left_join(canonical_scores,
    by = c(
      "basename" = "Basename", "row" = "Row",
      "col" = "Col", "pos" = "Pos"
    )
  ) %>%
  rowwise() %>%
  mutate(
    scorer_id  = as.integer(as.character(Scorer)),
    s1         = ifelse(scorer_id == Scorer1, new_score, Score1),
    s2         = ifelse(scorer_id == Scorer2, new_score, Score2),
    s3         = ifelse(scorer_id == Scorer3, new_score, Score3),
    new_median = median(c(s1, s2, s3), na.rm = TRUE),
    changed    = new_median != median_score
  ) %>%
  ungroup()

change_summary <- median_change %>%
  group_by(Scorer) %>%
  summarise(
    n = n(),
    n_changed = sum(changed, na.rm = TRUE),
    pct_changed = n_changed / n * 100,
    .groups = "drop"
  )

print(change_summary)
```


### 6.1 Proportion of CDAs where the median label would change per rater

```{r fig.width=7, fig.height=5}
#| fig-width: 7
#| fig-height: 5
# 6.1
p <- ggplot(change_summary, aes(x = Scorer, y = pct_changed / 100)) +
  geom_bar(stat = "identity", fill = "steelblue", color = "black", linewidth = 0.4) +
  geom_text(aes(label = sprintf("%d / %d", n_changed, n)),
    vjust = -0.4, size = 4.5
  ) +
  scale_y_continuous(
    labels = scales::percent, limits = c(0, 0.5),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(x = "Rater", y = "% CDAs with changed median label") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 15)
  )
p
save_plot(p, "median_label_changes.png", width = 7, height = 5, dir = "thesis_plots")
```

---

## 7. Within- vs between-rater variation

```{r}
# 7
# This section compares two kinds of disagreement on the same set of
# non-zero-median CDAs:
#   1. Within-rater: for each rescored CDA, the difference between the new
#      and old score.
#   2. Between-rater: for each CDA, the difference between each pair of
#      original raters (Scorer1 - Scorer2, Scorer1 - Scorer3, Scorer2 -
#      Scorer3) - pairwise rater-vs-rater comparison, not a comparison to
#      the median.
# Both are restricted to non-zero-median CDAs. If the within- and
# between-rater spreads are similar, most label noise is irreducible scoring
# uncertainty rather than systematic between-rater bias.
between_raw <- read_csv(
  here(
    "analyses", "01_generate_raw_data", "04_median_and_centre_coords",
    "combined_cda_data_median.csv"
  ),
  show_col_types = FALSE
) %>%
  select(Basename, Row, Col, Pos, Score1, Score2, Score3, Median_Score) %>%
  mutate(across(c(Score1, Score2, Score3, Median_Score), as.numeric)) %>%
  filter(Median_Score > 0)

# Within-rater differences (one per rescored CDA) and between-rater pairwise differences.
within_diff <- rescore_data %>%
  filter(!is.na(score_change)) %>%
  transmute(diff = as.numeric(score_change))

between_diff <- between_raw %>%
  transmute(d12 = Score1 - Score2, d13 = Score1 - Score3, d23 = Score2 - Score3) %>%
  pivot_longer(everything(), values_to = "diff") %>%
  filter(!is.na(diff)) %>%
  transmute(diff)

src_levels <- c("Within-rater", "Between-rater")
src_pal <- setNames(c("firebrick", "steelblue"), src_levels)

cat(
  "Within-rater differences:", nrow(within_diff),
  " | SD =", round(sd(within_diff$diff), 3), "\n"
)
cat(
  "Between-rater differences:", nrow(between_diff),
  " | SD =", round(sd(between_diff$diff), 3), "\n"
)
```

### 7.1 Absolute difference distribution

```{r fig.width=8, fig.height=5}
#| fig-width: 8
#| fig-height: 5
# 7.1
dist_abs <- bind_rows(
  within_diff %>% transmute(adiff = abs(diff), source = src_levels[1]),
  between_diff %>% transmute(adiff = abs(diff), source = src_levels[2])
) %>%
  count(source, adiff) %>%
  group_by(source) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  mutate(source = factor(source, levels = src_levels))

p <- ggplot(dist_abs, aes(x = factor(adiff), y = prop, fill = source)) +
  geom_col(position = "dodge", color = "black", linewidth = 0.3) +
  scale_fill_manual(values = src_pal, name = NULL) +
  scale_y_continuous(
    labels = scales::percent,
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(x = "Score difference", y = "Proportion") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 15),
    legend.text = element_text(size = 12),
    legend.position = "top"
  )
p
save_plot(p, "within_vs_between_abs.png", width = 8, height = 5, dir = "thesis_plots")
```

### 7.2 Disagreement by score class

```{r fig.width=9, fig.height=5}
#| fig-width: 9
#| fig-height: 5
# 7.2
# Error bars are +/-1 standard error of the mean.
within_class <- rescore_data %>%
  filter(!is.na(score_change)) %>%
  mutate(ad = abs(score_change)) %>%
  group_by(score = median_score) %>%
  summarise(mad = mean(ad), se = sd(ad) / sqrt(n()), n = n(), .groups = "drop") %>%
  mutate(source = src_levels[1])

between_class <- between_raw %>%
  rowwise() %>%
  mutate(mad_cda = mean(abs(c(Score1 - Score2, Score1 - Score3, Score2 - Score3)),
    na.rm = TRUE
  )) %>%
  ungroup() %>%
  group_by(score = Median_Score) %>%
  summarise(mad = mean(mad_cda), se = sd(mad_cda) / sqrt(n()), n = n(), .groups = "drop") %>%
  mutate(source = src_levels[1 + 1])

class_cmp <- bind_rows(within_class, between_class) %>%
  mutate(source = factor(source, levels = src_levels))

p <- ggplot(class_cmp, aes(
  x = factor(score), y = mad,
  color = source, group = source
)) +
  geom_errorbar(aes(ymin = mad - se, ymax = mad + se), width = 0.2, linewidth = 0.7) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 3) +
  scale_color_manual(values = src_pal, name = NULL) +
  labs(
    x = "Median Severity Score",
    y = "Mean absolute disagreement"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 15),
    legend.text = element_text(size = 12)
  )
p
save_plot(p, "within_vs_between_by_class.png", width = 9, height = 5, dir = "thesis_plots")
```

### 7.3 Reliability coefficient: intra- vs inter-rater Krippendorff's α

```{r}
# 7.3.1
# Interval Krippendorff's alpha - the same metric now used for the full
# inter-rater panel in 05_inter_rater_agreement.qmd (05 was switched from
# ordinal to interval; see its section 7.1 for the rationale) - applied here
# to each rater's two occasions (treating old_score and new_score as two
# raters). Using one metric for both keeps intra- and inter-rater reliability
# on a single comparable scale; the interval metric applies a fixed,
# interpretable spacing between severity scores rather than one derived from
# this sample's score frequencies. It also handles missing values natively.
# Per-rater 95% CIs are bootstrapped over CDAs with replacement (n-of-n) - the
# standard nonparametric bootstrap. Each rater rescored only ~100 CDAs, so a
# fixed-size subsample without replacement (as 05's headline figure uses) is
# not possible here, and a with-replacement bootstrap is in any case the
# correct sampling distribution for a CI on the estimate.
#
# `inter_alpha` below is NOT reused from 05_inter_rater_agreement.qmd - it is
# recomputed independently, from the original combined_cda_data_median.csv
# panel, restricted to non-zero-median CDAs only (to match what was available
# for rescoring; only non-zero-median CDAs were rescored - see GOHREP). That
# restriction is why it legitimately differs from 05's headline alpha, which is
# computed over all CDAs (the abundant, near-unanimous median-0 CDAs inflate
# 05's figure). It is given its own with-replacement bootstrap CI so that intra-
# and inter-rater alpha are compared on identical footing - same metric, same
# CDA restriction, same bootstrap scheme - with the comparison read off the
# overlap of the two CIs.
set.seed(42)

# Per-rater intra-rater alpha (old vs new as two raters) with bootstrap CIs.
intra_alpha <- rescore_data %>%
  filter(!is.na(old_score), !is.na(new_score)) %>%
  group_by(Scorer) %>%
  group_modify(~ {
    m <- rbind(.x$old_score, .x$new_score)
    a <- kripp.alpha(m, method = "interval")$value
    boot <- bootstrap_kripp_alpha(
      m,
      method = "interval", replace = TRUE, seed = NULL, suppress_warnings = TRUE,
      cache_path = here(
        "analyses", "06_intra_rater_reliability", "bootstrap_cache",
        paste0("intra_alpha_scorer_", .y$Scorer, ".rds")
      )
    )
    tibble(
      alpha = a,
      lo = quantile(boot, 0.025, na.rm = TRUE),
      hi = quantile(boot, 0.975, na.rm = TRUE)
    )
  }) %>%
  ungroup()

# Panel inter-rater alpha on the same non-zero-median CDAs (reference).
inter_long <- read_csv(
  here(
    "analyses", "01_generate_raw_data", "04_median_and_centre_coords",
    "combined_cda_data_median.csv"
  ),
  show_col_types = FALSE
) %>%
  filter(as.numeric(Median_Score) > 0) %>%
  mutate(CDA_ID = paste(Basename, Row, Col, Pos, sep = "_")) %>%
  select(CDA_ID, Scorer1, Score1, Scorer2, Score2, Scorer3, Score3) %>%
  pivot_longer(c(Scorer1, Score1, Scorer2, Score2, Scorer3, Score3),
    names_to = c(".value", "num"),
    names_pattern = "(Scorer|Score)([0-9]+)"
  ) %>%
  filter(!is.na(Scorer), !is.na(Score)) %>%
  mutate(Scorer = as.integer(Scorer), Score = as.numeric(Score))

inter_mat <- inter_long %>%
  select(Scorer, CDA_ID, Score) %>%
  pivot_wider(names_from = CDA_ID, values_from = Score) %>%
  tibble::column_to_rownames("Scorer") %>%
  as.matrix()

inter_alpha <- kripp.alpha(inter_mat, method = "interval")$value

# With-replacement (n-of-n) bootstrap CI for the inter-rater reference, matching
# the intra-rater bootstrap scheme above. Cached and reused in section 7.5.
inter_alpha_boot <- bootstrap_kripp_alpha(
  inter_mat,
  method = "interval", replace = TRUE, suppress_warnings = TRUE,
  cache_path = here(
    "analyses", "06_intra_rater_reliability", "bootstrap_cache",
    "alpha_inter_nonzero_interval.rds"
  )
)
inter_alpha_lo <- quantile(inter_alpha_boot, 0.025)
inter_alpha_hi <- quantile(inter_alpha_boot, 0.975)

cat(
  "Panel inter-rater alpha (interval, non-zero-median CDAs):",
  round(inter_alpha, 3),
  sprintf("95%% CI [%.3f, %.3f]\n", inter_alpha_lo, inter_alpha_hi)
)
print(intra_alpha)
```

```{r fig.width=7, fig.height=5}
#| fig-width: 7
#| fig-height: 5
# 7.3.2
p <- ggplot(intra_alpha, aes(x = Scorer, y = alpha)) +
  annotate("rect",
    xmin = -Inf, xmax = Inf, ymin = inter_alpha_lo, ymax = inter_alpha_hi,
    fill = "grey70", alpha = 0.35
  ) +
  geom_hline(yintercept = inter_alpha, linetype = "dashed", color = "grey40") +
  annotate("text",
    x = Inf, y = inter_alpha, vjust = -0.6, hjust = 1.05,
    label = sprintf("Inter-rater α = %.3f (95%% CI %.3f–%.3f)", inter_alpha, inter_alpha_lo, inter_alpha_hi),
    size = 4, color = "grey30"
  ) +
  geom_pointrange(aes(ymin = lo, ymax = hi), color = "firebrick", linewidth = 0.7) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Rater", y = "Krippendorff's α (interval)") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 15)
  )
p
save_plot(p, "intra_vs_inter_alpha.png", width = 7, height = 5)
```

### 7.4 Bootstrap distribution of pooled intra-rater α (raw scores)

```{r}
# 7.4.1
intra_pooled <- rescore_data %>%
  filter(!is.na(old_score), !is.na(new_score))

intra_raw_mat <- rbind(
  old = intra_pooled$old_score,
  new = intra_pooled$new_score
)

alpha_intra_raw <- kripp.alpha(intra_raw_mat, method = "interval")$value

alpha_intra_raw_boot <- bootstrap_kripp_alpha(
  intra_raw_mat,
  method = "interval", replace = TRUE, suppress_warnings = TRUE,
  cache_path = here("analyses", "06_intra_rater_reliability", "bootstrap_cache", "alpha_intra_raw.rds")
)

cat("Pooled intra-rater α (interval, raw):", round(alpha_intra_raw, 4), "\n")
cat("Bootstrap mean α:", round(mean(alpha_intra_raw_boot), 4), "\n")
cat(
  "95% CI: [", round(quantile(alpha_intra_raw_boot, 0.025), 4), ",",
  round(quantile(alpha_intra_raw_boot, 0.975), 4), "]\n"
)
```

```{r fig.width=7, fig.height=5}
#| fig-width: 7
#| fig-height: 5
# 7.4.2
boot_df <- data.frame(alpha = alpha_intra_raw_boot)

# The inter-rater and 0.8-threshold vlines can fall close together, so their
# lines would otherwise be hard to tell apart. Embedding each line's actual
# value in its own legend label, rather than a fixed string, keeps them
# unambiguous even when they visually overlap.
boot_mean_lab <- sprintf("Bootstrap mean = %.3f", mean(alpha_intra_raw_boot))
inter_lab <- sprintf("Inter-rater α (panel) = %.3f", inter_alpha)
threshold_lab <- "α = 0.8 reliability threshold (fixed convention)"

p <- ggplot(boot_df, aes(x = alpha)) +
  geom_histogram(binwidth = 0.005, fill = "cornflowerblue", color = "white") +
  geom_vline(aes(
    xintercept = mean(alpha_intra_raw_boot), color = boot_mean_lab,
    linetype = boot_mean_lab
  ), linewidth = 1) +
  geom_vline(aes(
    xintercept = inter_alpha, color = inter_lab,
    linetype = inter_lab
  ), linewidth = 0.8) +
  geom_vline(aes(
    xintercept = 0.8, color = threshold_lab,
    linetype = threshold_lab
  ), linewidth = 0.8) +
  scale_color_manual(
    name = NULL,
    values = setNames(
      c("firebrick", "darkgreen", "black"),
      c(boot_mean_lab, inter_lab, threshold_lab)
    )
  ) +
  scale_linetype_manual(
    name = NULL,
    values = setNames(
      c("dashed", "solid", "dotted"),
      c(boot_mean_lab, inter_lab, threshold_lab)
    )
  ) +
  labs(x = "Krippendorff's α (interval)", y = "Count") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 15),
    legend.text = element_text(size = 11),
    legend.position = "bottom"
  )
p
save_plot(p, "intra_krippendorff_alpha_raw.png", width = 7, height = 5, dir = "thesis_plots")
```

### 7.5 Intra- vs inter-rater α distributions

```{r}
# 7.5.1
# Both bootstrap distributions on one axis: the pooled intra-rater
# (test-retest) alpha from 7.4, and the panel inter-rater alpha bootstrapped
# over the same non-zero-median CDAs (resampling the inter_mat columns with
# replacement). This shows directly how much the two reliabilities overlap.
# Same call (and cache) as the inter-rater bootstrap in 7.3.1, so this reads
# the cached result rather than recomputing.
alpha_inter_boot <- bootstrap_kripp_alpha(
  inter_mat,
  method = "interval", replace = TRUE, suppress_warnings = TRUE,
  cache_path = here("analyses", "06_intra_rater_reliability", "bootstrap_cache", "alpha_inter_nonzero_interval.rds")
)

cat(
  "Intra-rater α: mean", round(mean(alpha_intra_raw_boot), 3),
  " 95% CI [", round(quantile(alpha_intra_raw_boot, 0.025), 3), ",",
  round(quantile(alpha_intra_raw_boot, 0.975), 3), "]\n"
)
cat(
  "Inter-rater α: mean", round(mean(alpha_inter_boot), 3),
  " 95% CI [", round(quantile(alpha_inter_boot, 0.025), 3), ",",
  round(quantile(alpha_inter_boot, 0.975), 3), "]\n"
)
```

```{r fig.width=7, fig.height=5}
#| fig-width: 7
#| fig-height: 5
# 7.5.2
alpha_dist <- bind_rows(
  tibble(alpha = alpha_intra_raw_boot, source = "Intra-rater (test–retest)"),
  tibble(alpha = alpha_inter_boot, source = "Inter-rater (panel)")
) %>%
  mutate(source = factor(source, levels = c("Intra-rater (test–retest)", "Inter-rater (panel)")))

dist_pal <- c("Intra-rater (test–retest)" = "firebrick", "Inter-rater (panel)" = "steelblue")

p <- ggplot(alpha_dist, aes(x = alpha, fill = source, color = source)) +
  geom_density(alpha = 0.4, linewidth = 0.8) +
  scale_fill_manual(values = dist_pal, name = NULL) +
  scale_color_manual(values = dist_pal, name = NULL) +
  labs(x = "Krippendorff's α (interval)", y = "Density") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 13), axis.title = element_text(size = 15),
    legend.text = element_text(size = 12), legend.position = "top"
  )
p
save_plot(p, "intra_vs_inter_alpha_distributions.png", width = 7, height = 5)
```

---

## 8. Interpretation and Discussion

_Add notes here._
