---
title: "Segmenting cell death area from leaf background"
format:
  html:
    code-fold: true
---

Date: 13/7/2026

## GOHREP

**Goal:**

Test whether the cell death area (CDA) of a leaf can be separated from the
healthy leaf background universally, with one method and no per-image manual
threshold tuning, using the distribution of pixel values.

**Hypothesis:**

An earlier attempt (`PhD_DL_Models/analyses/0009_Non_DL_Green_Channel_Analysis`)
watershedded on image brightness and found the threshold had to be retuned per
image. The reason is colour. Under this UV-fluorescence imaging the healthy leaf
is red (high R, low G) and cell death is yellow-green (G at or above R), and red
leaf and green CDA have similar brightness. Segmenting on a greenness channel
(G - R) instead should separate CDA from leaf with a threshold that Otsu can set
automatically per image, because the two tissues form distinct modes in greenness
even when they do not in brightness.

**Rationale:**

Otsu's method finds the threshold that best splits a bimodal distribution. If
CDA and leaf are two modes in greenness, Otsu adapts to each image with no
manual input. Where there is no CDA (median-0 leaves) there is only one mode, so
segmentation is only meaningful once an image is known to be non-zero. 0 and
non-0 are therefore treated separately here.

---

## 1. Setup

### 1.1 Load required packages

```{r}
#| output: false
# 1.1
library(tiff)
suppressMessages(library(EBImage))
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
library(here)
library(RColorBrewer)
```

### 1.2 Helper functions and colour palettes

```{r}
# 1.2
source(here("src", "plotting_helpers.R"))
save_plot <- make_save_plot("10_segmentation_and_feature_extraction")
data_dir <- here("data", "cropped_images_medoid")
```

### 1.3 Segmentation functions

```{r}
# 1.3
load_img <- function(path) {
  img <- readTIFF(path, info = FALSE)
  if (length(dim(img)) == 2) img <- array(img, dim = c(dim(img), 3))
  img[, , 1:3, drop = FALSE]
}

# Leaf mask: non-black pixels (mean RGB >= 0.15), dropping the near-black
# background and holes. 09_image_entropy uses a comparable non-dark threshold
# on peak brightness.
leaf_mask <- function(img) (img[, , 1] + img[, , 2] + img[, , 3]) / 3 >= 0.15

# Greenness channel: green minus red. Healthy (red) leaf is strongly negative;
# yellow-green CDA is near zero or positive.
greenness <- function(img) img[, , 2] - img[, , 1]

# Otsu threshold on a value vector, plus the "separability" = between-class
# variance at the threshold / total variance (a 2-class eta^2 in [0,1]). High
# separability => the values are well split into two modes (bimodal).
otsu_sep <- function(x, nbins = 64) {
  rng <- range(x)
  if (diff(rng) < 1e-6) {
    return(c(thr = rng[1], sep = 0, frac_hi = 0))
  }
  br <- seq(rng[1], rng[2], length.out = nbins + 1)
  mids <- (br[-1] + br[-length(br)]) / 2
  h <- tabulate(cut(x, br, include.lowest = TRUE), nbins)
  p <- h / sum(h)
  mu <- sum(p * mids)
  tot <- sum(p * (mids - mu)^2)
  w1 <- cumsum(p)
  m1 <- cumsum(p * mids)
  bcv <- (mu * w1 - m1)^2 / (w1 * (1 - w1) + 1e-12)
  k <- which.max(bcv)
  c(thr = br[k + 1], sep = max(bcv) / (tot + 1e-12), frac_hi = mean(x > br[k + 1]))
}

# Full spatial segmentation of one image: Otsu on the (blurred) greenness within
# the leaf, followed by morphological cleanup.
segment_cda <- function(img, sigma = 2) {
  leaf <- leaf_mask(img)
  g <- greenness(img)
  gs <- (g - min(g)) / (max(g) - min(g) + 1e-9) # rescale to 0-1 for EBImage
  gb <- EBImage::gblur(gs, sigma = sigma)
  thr <- EBImage::otsu(matrix(gb[leaf]), range = c(0, 1))
  mask <- leaf & (gb > thr)
  # Clean up with a median filter, whose radius scales with image size so the
  # smoothing is constant across crop resolutions: base 5 px at the 264-px median
  # crop, applied at radius 2*base. A median filter removes thin protrusions
  # (veins, spurs) far more effectively than a morphological opening of the same
  # scale while barely shrinking the bulk, at 0.95 circularity for 98.6% of the
  # area retained (see the cleanup comparison below).
  b <- max(1, round(5 * mean(dim(img)[1:2]) / 264))
  mask <- EBImage::fillHull(EBImage::medianFilter(mask, 2 * b))
  # The median filter and fillHull ignore the leaf boundary, so they can bridge
  # across background notches and pull off-leaf (near-black) pixels into the mask.
  # Re-apply the leaf mask so the CDA can never extend onto the background.
  mask <- mask & leaf
  # Keep only the largest connected component, dropping small floating islands.
  lab <- EBImage::bwlabel(mask)
  if (max(lab) > 1) mask <- lab == which.max(tabulate(lab[lab > 0]))
  # Re-applying the leaf mask above punches holes back into the CDA wherever the
  # leaf is dark inside the lesion (veins, specks below the 0.15 brightness cut).
  # Fill those enclosed holes so the CDA is a single solid region with no interior
  # dots in its outline. fillHull only fills fully-enclosed holes, so it cannot
  # re-extend the mask back onto the background.
  mask <- EBImage::fillHull(mask)
  list(leaf = leaf, greenness = gb, mask = mask, cda_frac = mean(mask))
}

# Render a single-channel raster (e.g. the greenness map) as RGB, recolouring
# pixels outside `keep` (e.g. off-leaf background) a flat colour. Used so the
# Otsu-excluded background reads visually distinct from the leaf pixels that
# actually feed the threshold, rather than looking like more greenness data.
recolour_outside <- function(chan, keep, colour = c(0.12, 0.56, 1)) {
  out <- array(chan, dim = c(dim(chan), 3))
  for (k in 1:3) {
    ch <- out[, , k]
    ch[!keep] <- colour[k]
    out[, , k] <- ch
  }
  out
}

# Overlay a mask on an image by turning masked pixels yellow-magenta.
overlay <- function(img, mask, colour = c(1, 0, 1)) {
  out <- img
  for (k in 1:3) {
    ch <- out[, , k]
    ch[mask] <- colour[k]
    out[, , k] <- ch
  }
  out
}

# Bold column headers (top) and rotated row labels (left) for the base-graphics
# example grids, drawn once into the outer margins instead of per-panel titles.
colhead <- function(labels, ncol, cex = 1) {
  for (c in seq_along(labels)) {
    mtext(labels[c],
      side = 3, outer = TRUE, line = 0.3, at = (c - 0.5) / ncol,
      font = 2, cex = cex
    )
  }
}
rowlab <- function(label, r, nrow, cex = 1) {
  mtext(label,
    side = 2, outer = TRUE, las = 0, line = 1,
    at = 1 - (r - 0.5) / nrow, cex = cex
  )
}

# Draw a coloured outline of a mask onto an image (a boundary band of the given
# half-width), leaving the interior visible - clearer than a filled overlay.
outline <- function(img, mask, colour = c(0.12, 0.56, 1), width = 2) {
  br <- EBImage::makeBrush(2 * width + 1, "disc")
  edge <- EBImage::dilate(mask, br) & !EBImage::erode(mask, br)
  out <- img
  for (k in 1:3) {
    ch <- out[, , k]
    ch[edge] <- colour[k]
    out[, , k] <- ch
  }
  out
}

# Shannon entropy of pixel values in [0, 1], matching 09_image_entropy's
# shannon_entropy(): quantise to 8-bit intensities and take H = -sum(p*log2 p).
# region_entropy() applies it over the pixels of a mask, flattened across the
# three RGB channels, exactly as the whole-image version does.
shannon_entropy_vec <- function(v) {
  q <- as.integer(round(pmin(pmax(v, 0), 1) * 255))
  counts <- tabulate(q + 1L, nbins = 256L)
  p <- counts[counts > 0] / length(q)
  -sum(p * log2(p))
}
region_entropy <- function(img, mask) {
  if (!any(mask)) {
    return(NA_real_)
  }
  shannon_entropy_vec(c(img[, , 1][mask], img[, , 2][mask], img[, , 3][mask]))
}
```

---

## 2. Why brightness fails and greenness works

- Otsu's threshold on brightness falls at 66 of 255 for the score-2 crop and 137
  for the score-6 crop, so no single fixed brightness cut can serve both images.
- On greenness the same two thresholds are -0.16 and -0.15, so a single greenness
  cut transfers between crops with no retuning.

```{r fig.width=11.538, fig.height=6.154, dpi=130}
#| fig-width: 11.538
#| fig-height: 6.154
#| fig-dpi: 130
# Display device matches the saved png() below (1500x800 px at res 130), so the
# figure renders here exactly as it is saved.
# 2.1
set.seed(2)
demo <- c(
  sample(list.files(file.path(data_dir, 2), pattern = "\\.tif$", full.names = TRUE), 1),
  sample(list.files(file.path(data_dir, 6), pattern = "\\.tif$", full.names = TRUE), 1)
)
demo_scores <- c(2, 6)
draw_channels <- function() {
  par(mfrow = c(2, 3), mar = c(3.2, 3.4, 1, 1), mgp = c(2, 0.6, 0), oma = c(0, 3, 2.4, 0))
  for (r in seq_along(demo)) {
    img <- load_img(demo[r])
    leaf <- leaf_mask(img)
    bright <- ((img[, , 1] + img[, , 2] + img[, , 3]) / 3)[leaf]
    g <- greenness(img)[leaf]
    plot(as.raster(img))
    hist(bright * 255,
      breaks = 40, col = "grey70", main = "",
      xlab = "Brightness (0-255)", ylab = "Pixels"
    )
    hist(g,
      breaks = 40, col = "darkolivegreen3", main = "",
      xlab = "Greenness (G - R)", ylab = "Pixels"
    )
    abline(v = otsu_sep(g)["thr"], col = "red", lwd = 2)
    rowlab(sprintf("Severity Score %d", demo_scores[r]), r, 2)
  }
  colhead(c("Crop", "Whole-leaf brightness", "Whole-leaf greenness (G - R)"), 3, 0.95)
}
png(plot_output_path("10_segmentation_and_feature_extraction", "supp_plots", "seg_brightness_vs_greenness.png"),
  width = 1500, height = 800, res = 130
)
draw_channels()
invisible(dev.off())
draw_channels()

# The point of the figure is threshold stability, not the shape of one histogram:
# report where Otsu lands in each channel for the two crops.
for (r in seq_along(demo)) {
  img <- load_img(demo[r])
  leaf <- leaf_mask(img)
  bright <- ((img[, , 1] + img[, , 2] + img[, , 3]) / 3)[leaf]
  cat(sprintf(
    "score %d: Otsu threshold on brightness = %.0f/255, on greenness = %+.2f\n",
    demo_scores[r], 255 * otsu_sep(bright)["thr"],
    otsu_sep(greenness(img)[leaf])["thr"]
  ))
}
```

---

## 3. Per-image segmentation via Otsu on greenness

- For every non-zero score the leaf-greenness histogram is bimodal and the Otsu
  threshold lands in the valley between the modes, so the CDA is isolated with no
  manual input.
- The score-0 crop has no lesion, yet Otsu still splits it, and dataset-wide the
  segmented region covers 49% of the leaf at score 0 against 46% at score 6, so
  the segmented area alone cannot say whether a lesion is there.

```{r fig.width=10.714, fig.height=10.714, dpi=140}
#| fig-width: 10.714
#| fig-height: 10.714
#| fig-dpi: 140
# Display device matches the saved png() below (1500x1500 px at res 140), so the
# figure renders here exactly as it is saved.
# 3.1
set.seed(7)
seg_scores <- c(0, 2, 4, 6)
seg_files <- sapply(seg_scores, function(sc) {
  sample(list.files(file.path(data_dir, sc), pattern = "\\.tif$", full.names = TRUE), 1)
})

draw_seg <- function() {
  par(
    mfrow = c(length(seg_files), 4), mar = c(2.4, 2, 0.6, 0.6), mgp = c(1.3, 0.4, 0),
    oma = c(0, 2.6, 2.4, 0)
  )
  for (i in seq_along(seg_files)) {
    img <- load_img(seg_files[i])
    s <- segment_cda(img)
    g <- greenness(img)[s$leaf]
    plot(as.raster(img))
    plot(as.raster(recolour_outside(s$greenness, s$leaf)))
    hist(g, breaks = 40, col = "darkolivegreen3", main = "", xlab = "Greenness (G - R)", ylab = "Frequency")
    abline(v = otsu_sep(g)["thr"], col = "red", lwd = 2)
    plot(as.raster(outline(img, s$mask)))
    rowlab(sprintf("Score %d", seg_scores[i]), i, length(seg_scores))
  }
  colhead(c("Crop", "Greenness (G - R)", "Greenness histogram", "Segmented CDA"), 4, 0.9)
}
png(plot_output_path("10_segmentation_and_feature_extraction", "supp_plots", "seg_otsu_examples.png"),
  width = 1500, height = 1500, res = 140
)
draw_seg()
invisible(dev.off())
draw_seg()
```

### 3.1 Segmentation of six CDAs per severity score

- Six randomly sampled crops per score, 42 in all, so this is a random audit of
  the segmentation rather than a curated set of successes.
- The blue outline follows the visible lesion closely from score 2 upward, while
  at scores 0-1 it encloses whole regions of healthy or merely infiltrated
  tissue, which is where the feature overlap later shows up.

```{r fig.width=13.333, fig.height=13.333, dpi=150}
#| fig-width: 13.333
#| fig-height: 13.333
#| fig-dpi: 150
# Display device matches the saved png() below (2000x2000 px at res 150), so the
# figure renders here exactly as it is saved.
# 3.1
set.seed(11)
n_per <- 6
grid_scores <- 0:6
grid_files <- lapply(grid_scores, function(sc) {
  sample(list.files(file.path(data_dir, sc), pattern = "\\.tif$", full.names = TRUE), n_per)
})

draw_seg_grid <- function() {
  par(
    mfrow = c(length(grid_scores), n_per), mar = c(0.15, 0.15, 0.15, 0.15),
    oma = c(0, 3.2, 0, 0)
  )
  for (r in seq_along(grid_scores)) {
    for (f in grid_files[[r]]) {
      img <- load_img(f)
      plot(as.raster(outline(img, segment_cda(img)$mask)))
    }
    mtext(sprintf("Score %d", grid_scores[r]),
      side = 2, outer = TRUE, las = 0,
      line = 1, cex = 1, at = 1 - (r - 0.5) / length(grid_scores)
    )
  }
}
png(plot_output_path("10_segmentation_and_feature_extraction", "supp_plots", "seg_validation_grid.png"),
  width = 2000, height = 2000, res = 150
)
draw_seg_grid()
invisible(dev.off())
draw_seg_grid()
```

### 3.2 Otsu vs background-only leaf masks

Which pixels count as leaf when estimating per-image lighting is a separate
question from segmenting the lesion. The image-adaptive Otsu threshold is
compared against a low fixed cut (brightness at or above 0.10) that removes only
the near-black background.

```{r fig.width=10, fig.height=12, dpi=150}
#| fig-width: 10
#| fig-height: 12
#| fig-dpi: 150
# Display device matches the saved png() below (1500x1800 px at res 150), so the
# figure renders here exactly as it is saved.
# 3.2
set.seed(7)
leaf_mask_examples <- lapply(c(0, 2, 4, 6), function(sc) {
  files <- list.files(file.path(data_dir, sc), pattern = "\\.tif$", full.names = TRUE)
  f <- sample(files, 1)
  img <- readTIFF(f, info = FALSE)
  if (length(dim(img)) == 2) img <- array(img, dim = c(dim(img), 3))
  img <- img[, , 1:3, drop = FALSE]
  bright_px <- (img[, , 1] + img[, , 2] + img[, , 3]) / 3
  thr <- EBImage::otsu(bright_px, range = c(0, 1))
  otsu_mask <- bright_px > thr
  if (sum(otsu_mask) < 0.02 * length(otsu_mask)) otsu_mask <- bright_px >= 0.1
  leaf_mask <- bright_px >= 0.10
  list(
    sc = sc, img = img, otsu = otsu_mask, leaf = leaf_mask,
    otsu_frac = mean(otsu_mask), leaf_frac = mean(leaf_mask)
  )
})

mask_out <- function(img, mask) {
  out <- img
  for (k in 1:3) {
    ch <- out[, , k]
    ch[!mask] <- 0
    out[, , k] <- ch
  }
  out
}
draw_leaf_masks <- function() {
  par(mfrow = c(length(leaf_mask_examples), 3), mar = c(0.5, 2, 2.2, 0.5))
  for (e in leaf_mask_examples) {
    plot(as.raster(e$img))
    title(sprintf("score %d: crop", e$sc))
    plot(as.raster(mask_out(e$img, e$otsu)))
    title(sprintf("Otsu region (%.0f%%)", 100 * e$otsu_frac))
    plot(as.raster(mask_out(e$img, e$leaf)))
    title(sprintf("leaf region (%.0f%%)", 100 * e$leaf_frac))
  }
}
png(plot_output_path("10_segmentation_and_feature_extraction", "supp_plots", "leaf_segmentation_examples.png"),
  width = 1500, height = 1800, res = 150
)
draw_leaf_masks()
invisible(dev.off())
draw_leaf_masks()
```

- At score 6 the Otsu mask keeps only 27% of the crop against 100% for the fixed
  cut, because the bright yellow-green lesion outranks the darker red healthy
  tissue, so Otsu selects the lesion rather than the leaf.
- Where a black background exists the two agree closely (78% against 79% at score
  2), so the fixed cut is the safer lighting proxy for `09_image_entropy`.

### 3.3 Choosing the morphological cleanup step

- A median filter at radius 2b gives the roundest outline at circularity 0.95
  while retaining 98.6% of the raw mask area, so it strips vein spurs without
  eroding the lesion.
- The raw mask sits at circularity 0.68 and opening-by-reconstruction leaves it
  there unchanged, so reconstruction is no use against spurs attached to the
  bulk of the mask.

```{r fig.width=16, fig.height=12, dpi=150}
#| fig-width: 16
#| fig-height: 12
#| fig-dpi: 150
# Display device matches the saved png() below (2400x1800 px at res 150), so the
# figure renders here exactly as it is saved.
# 3.3
odd <- function(n) {
  n <- round(n)
  if (n %% 2 == 0) n <- n + 1
  max(3, n)
}
largest_cc <- function(mask) {
  lab <- EBImage::bwlabel(mask)
  if (max(lab) > 1) mask <- lab == which.max(tabulate(lab[lab > 0]))
  mask
}
# raw thresholded CDA mask, before any morphological cleanup
raw_cda_mask <- function(img, sigma = 2) {
  leaf <- leaf_mask(img)
  g <- greenness(img)
  gs <- (g - min(g)) / (max(g) - min(g) + 1e-9)
  gb <- EBImage::gblur(gs, sigma = sigma)
  thr <- EBImage::otsu(matrix(gb[leaf]), range = c(0, 1))
  leaf & (gb > thr)
}
# geodesic reconstruction by dilation of a marker under a mask
reconstruct <- function(marker, mask, brush) {
  prev <- marker
  for (i in 1:300) {
    cur <- EBImage::dilate(prev, brush) & mask
    if (all(cur == prev)) break
    prev <- cur
  }
  prev
}
cleanups <- list(
  "Raw" = function(m, b, d) largest_cc(EBImage::fillHull(m)),
  "Open x2" = function(m, b, d) {
    largest_cc(EBImage::fillHull(EBImage::opening(m, EBImage::makeBrush(odd(b * 2), "disc"))))
  },
  "Open x3" = function(m, b, d) {
    largest_cc(EBImage::fillHull(EBImage::opening(m, EBImage::makeBrush(odd(b * 3), "disc"))))
  },
  "Median r=b" = function(m, b, d) largest_cc(EBImage::fillHull(EBImage::medianFilter(m, b))),
  "Median r=2b" = function(m, b, d) largest_cc(EBImage::fillHull(EBImage::medianFilter(m, 2 * b))),
  "Blur sig=side/50" = function(m, b, d) largest_cc(EBImage::fillHull(EBImage::gblur(m * 1, sigma = d / 50) > 0.5)),
  "Blur sig=side/30" = function(m, b, d) largest_cc(EBImage::fillHull(EBImage::gblur(m * 1, sigma = d / 30) > 0.5)),
  "Reconstruction" = function(m, b, d) {
    br <- EBImage::makeBrush(odd(b), "disc")
    largest_cc(EBImage::fillHull(reconstruct(EBImage::erode(m, br), m, br)))
  }
)
circularity <- function(mask) {
  if (sum(mask) < 10) {
    return(NA_real_)
  }
  sh <- EBImage::computeFeatures.shape(EBImage::bwlabel(mask))
  if (is.null(sh) || nrow(sh) == 0 || sh[1, "s.perimeter"] <= 0) {
    return(NA_real_)
  }
  4 * pi * sh[1, "s.area"] / (sh[1, "s.perimeter"]^2)
}

set.seed(11)
clean_files <- unlist(lapply(3:6, function(sc) {
  sample(list.files(file.path(data_dir, sc), pattern = "\\.tif$", full.names = TRUE), 4)
}))
circ_m <- area_m <- matrix(NA_real_, length(clean_files), length(cleanups),
  dimnames = list(NULL, names(cleanups))
)
for (i in seq_along(clean_files)) {
  img <- load_img(clean_files[i])
  d <- mean(dim(img)[1:2])
  b <- max(1, round(5 * d / 264))
  rm <- raw_cda_mask(img)
  for (j in seq_along(cleanups)) {
    mk <- cleanups[[j]](rm, b, d)
    circ_m[i, j] <- circularity(mk)
    area_m[i, j] <- mean(mk)
  }
}
cleanup_summary <- data.frame(
  method = names(cleanups),
  circularity = round(colMeans(circ_m, na.rm = TRUE), 3),
  area_retained = round(colMeans(area_m / area_m[, "Raw"], na.rm = TRUE), 3),
  row.names = NULL
)
print(cleanup_summary)

# Visual comparison: rows = crops, columns = cleanup methods.
vis_files <- clean_files[c(1, 3, 5, 7, 9, 11)]
draw_cleanup_grid <- function() {
  par(mfrow = c(length(vis_files), length(cleanups)), mar = c(0.3, 0.3, 1.8, 0.3))
  for (f in vis_files) {
    img <- load_img(f)
    d <- mean(dim(img)[1:2])
    b <- max(1, round(5 * d / 264))
    rm <- raw_cda_mask(img)
    for (j in seq_along(cleanups)) {
      plot(as.raster(outline(img, cleanups[[j]](rm, b, d))))
      if (f == vis_files[1]) title(main = names(cleanups)[j], cex.main = 0.9, line = 0.4)
    }
  }
}
png(plot_output_path("10_segmentation_and_feature_extraction", "supp_plots", "seg_cleanup_comparison.png"),
  width = 2400, height = 1800, res = 150
)
draw_cleanup_grid()
invisible(dev.off())
draw_cleanup_grid()
```

---

## 4. Can the distribution alone detect whether a CDA is present?

Segmenting assumes a CDA exists. The natural universal test for whether a lesion
is present is the bimodality of the leaf-greenness distribution, computed below
for every image alongside the other region features.

```{r}
# 5.1
cache_path <- here(
  "analyses", "10_segmentation_and_feature_extraction", "segmentation_cache",
  "cda_segmentation_metrics.rds"
)

if (file.exists(cache_path)) {
  seg <- readRDS(cache_path)
} else {
  scores <- list.dirs(data_dir, recursive = FALSE, full.names = FALSE)
  scores <- scores[grepl("^[0-9]+$", scores)]
  rows <- list()
  for (sc in scores) {
    files <- list.files(file.path(data_dir, sc), pattern = "\\.tif$", full.names = TRUE)
    m <- t(vapply(files, function(f) {
      # All region features are read off the SAME segmentation: the leaf mask,
      # the spatial CDA mask from segment_cda() (Otsu on blurred greenness ->
      # morphological cleanup -> largest connected component), and the healthy
      # leaf (leaf minus CDA). Nothing is re-thresholded per feature.
      img <- load_img(f)
      s <- segment_cda(img)
      leaf <- s$leaf # leaf region ((R+G+B)/3 >= 0.15)
      cda <- s$mask & leaf # segmented CDA (largest greenness component)
      healthy <- leaf & !cda # healthy leaf = leaf minus CDA
      gmat <- greenness(img) # greenness G - R over the whole image
      bmat <- (img[, , 1] + img[, , 2] + img[, , 3]) / 3 # brightness (mean RGB)
      R <- img[, , 1]
      G <- img[, , 2]
      nsd <- function(x) if (length(x) > 1) sd(x) else 0
      mn <- function(x) if (length(x)) mean(x) else 0

      gv_leaf <- gmat[leaf] # greenness over the whole leaf
      o <- otsu_sep(gv_leaf) # separability (bimodality) of leaf greenness
      n_green <- sum(G[leaf] >= R[leaf])
      n_red <- sum(R[leaf] > G[leaf]) # colour counts over leaf

      c(
        sep = unname(o["sep"]), # 4  leaf-greenness bimodality
        green_gap = if (any(cda) && any(healthy)) { # 3  CDA vs healthy-leaf greenness
          mean(gmat[cda]) - mean(gmat[healthy])
        } else {
          0
        },
        cda_green = mn(gmat[cda]), # 2  CDA greenness
        frac_g_gt_r = sum(cda) / sum(leaf), # 1  lesion size = |CDA| / |leaf|
        cda_green_sd = nsd(gmat[cda]) * 255, # 7  within-CDA greenness spread
        cda_bright = mn(bmat[cda]) * 255, # 5  CDA brightness
        cda_bright_sd = nsd(bmat[cda]) * 255, # 8  within-CDA brightness spread
        leaf_green_sd = nsd(gv_leaf) * 255, # 6  whole-leaf greenness spread
        mean_green = mn(G[leaf]) * 255, # 10 mean green over leaf
        log_gr = log((n_green + 1) / (n_red + 1)), # 9  log green:red count over leaf
        leaf_bright = mn(bmat[leaf]) * 255, # 11 leaf brightness
        ent_gap = if (any(cda) && any(healthy)) { # 12 CDA minus healthy-leaf Shannon entropy
          region_entropy(img, cda) - region_entropy(img, healthy)
        } else {
          0
        }
      )
    }, numeric(12)))
    rows[[sc]] <- data.frame(
      image_id = tools::file_path_sans_ext(basename(files)),
      score = as.integer(sc), m, row.names = NULL
    )
    message(sprintf("score %s done (%d)", sc, length(files)))
  }
  seg <- dplyr::bind_rows(rows)
  saveRDS(seg, cache_path)
}

seg <- seg %>% mutate(
  score_f = factor(score, levels = 0:6),
  non0 = score > 0
)
cat(sprintf("Segmentation metrics for %d images.\n", nrow(seg)))
```

- Median separability rises from 0.65 at score 0 to 0.85 at score 6 and
  classifies 0 against non-0 at AUC 0.944 on its own, so the distribution detects
  a lesion reliably.
- The same test on segmented CDA area gives AUC 0.451, no better than chance, so
  the shape of the greenness distribution carries the presence signal where its
  area does not.

```{r}
# 5.2
# Rank-based (Mann-Whitney) AUC: probability a random non-0 image scores higher
# than a random 0 image on the given feature.
auc <- function(feature, positive) {
  r <- rank(feature)
  n_pos <- sum(positive)
  n_neg <- sum(!positive)
  (sum(r[positive]) - n_pos * (n_pos + 1) / 2) / (n_pos * n_neg)
}
cat(sprintf(
  "AUC (0 vs non-0):  separability = %.3f  |  fixed-threshold area (G>R) = %.3f\n",
  auc(seg$sep, seg$non0), auc(seg$frac_g_gt_r, seg$non0)
))
```

```{r fig.width=8, fig.height=5}
#| fig-width: 8
#| fig-height: 5
# 5.3
p_sep <- ggplot(seg, aes(x = score_f, y = sep)) +
  geom_boxplot(aes(fill = score_f), outlier.shape = NA, alpha = 0.85) +
  geom_jitter(width = 0.18, height = 0, alpha = 0.1, size = 0.4) +
  scale_fill_manual(name = "Severity Score", values = sev_pal) +
  labs(x = "Severity score", y = "Leaf-greenness separability (bimodality)") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 13), axis.title = element_text(size = 15),
    legend.title = element_text(size = 13), legend.text = element_text(size = 11)
  )
p_sep
save_plot(p_sep, "seg_separability_by_score.png", width = 8, height = 5)
```

### 4.1 The twelve features, and the exported feature table

The full pipeline, end to end, for one cropped CDA image:

1. Load the raw RGB crop. No resizing, since every feature is a mean, a spread or
   a ratio and so does not depend on the crop's pixel dimensions.
2. Leaf mask. Keep pixels whose mean RGB is at least 0.15, which drops the
   near-black off-leaf background.
3. Segment the CDA. Rescale greenness (G - R) to 0-1, Gaussian blur at sigma 2,
   take the Otsu threshold over the leaf pixels only, smooth the mask with a
   median filter at radius 2b (b = round(5 x side / 264), so smoothing is
   constant across crop resolutions), re-apply the leaf mask, keep the largest
   connected component, and fill its enclosed holes.
4. Summarise the crop. Eleven scalar features are measured here, and Shannon
   entropy is joined from `09_image_entropy` to make twelve.

No per-image threshold tuning anywhere. "Greenness" means G - R (green minus red
channel). The "CDA" is the segmented region from step 3, the "leaf" is every
pixel kept by step 2, and the "healthy leaf" is the leaf minus the CDA.

Segmentation features, measured in this analysis:

| Variable | Feature | What it measures | With severity |
|---|---|---|---|
| `ent_gap` | CDA-leaf entropy difference | Shannon entropy (bits) of the segmented CDA minus that of the healthy leaf | rises |
| `cda_green` | CDA greenness | mean greenness (G − R) inside the segmented CDA | rises |
| `green_gap` | CDA–leaf contrast | CDA greenness minus healthy-leaf (leaf minus CDA) greenness | rises |
| `sep` | greenness bimodality | Otsu between-class ÷ total variance of leaf greenness (0–1): how cleanly the leaf splits into CDA + healthy | rises (also detects presence) |
| `cda_bright` | CDA brightness | mean brightness (mean of R,G,B, 0–255) inside the CDA | rises |
| `leaf_green_sd` | whole-leaf greenness spread | SD of greenness over all leaf pixels (0–255) | rises |
| `cda_green_sd` | within-CDA greenness spread | SD of greenness among CDA pixels: how mottled the lesion's colour is | ~flat |
| `cda_bright_sd` | within-CDA brightness spread | SD of brightness among CDA pixels | ~flat |

Colour, brightness and entropy features, taken from `09_image_entropy`:

| Variable | Feature | What it measures | With severity |
|---|---|---|---|
| `mean_green` | mean green | mean green-channel intensity over leaf pixels (0–255) | rises |
| `log_gr` | log green/red ratio | log of the green-pixel to red-pixel count ratio over leaf pixels | rises |
| `leaf_bright` | leaf brightness | mean brightness over non-black leaf pixels (0–255) | rises |
| `entropy_raw` | Shannon entropy | information content of the raw crop (bits) | weakly rises |

All twelve are written out for every image, along with the source leaf each crop
came from and the lesion-area fraction `frac_g_gt_r`. `source` is the crop
filename with the trailing `_Row_Col_Pos` stripped, and is the grouping unit for
any leakage-safe cross validation, since crops from one leaf share lighting,
plant and tissue. The table is the input to `10b_score_prediction.qmd` and
`11_project_feature_differences`.

```{r}
# 5.4  Export the full (unbalanced) feature table for every image.
# Region features 1-11 come from `seg` above; Shannon entropy is the twelfth,
# taken from the entropy analysis.
features_all <- seg %>%
  left_join(
    readRDS(here("analyses", "09_image_entropy", "entropy_cache", "cda_metrics.rds")) %>%
      filter(size == "Raw") %>%
      transmute(image_id, entropy_raw = entropy),
    by = "image_id"
  ) %>%
  mutate(source = sub("_[0-9]+_[0-9]+_[0-9]+$", "", image_id)) %>%
  select(
    image_id, source, score,
    ent_gap, green_gap, cda_green, sep, cda_green_sd, cda_bright,
    cda_bright_sd, leaf_green_sd, mean_green, log_gr, leaf_bright,
    entropy_raw, frac_g_gt_r
  )

features_path <- here("data", "features", "cda_image_features.csv")
dir.create(dirname(features_path), showWarnings = FALSE, recursive = TRUE)
write.csv(features_all, features_path, row.names = FALSE)
cat(sprintf(
  "Wrote %d images x %d features to data/features/%s (%d source leaves).\n",
  nrow(features_all), length(setdiff(names(features_all), c("image_id", "source", "score"))),
  basename(features_path), length(unique(features_all$source))
))
```

---

## 5. How each feature relates to severity

- Median CDA greenness climbs from -0.19 at score 0 to +0.02 at score 6, crossing
  zero near score 4, which is where the lesion becomes greener than the leaf
  around it.
- Interquartile ranges of neighbouring scores overlap for all twelve features, so
  the boxplots already show that no single feature will separate adjacent
  classes.

```{r fig.width=11, fig.height=12}
#| fig-width: 11
#| fig-height: 12
# 6.1
# All twelve features are in the table exported in section 4.1.
feat6 <- features_all %>% mutate(score_f = factor(score, levels = 0:6))

# Order groups the features by region (lesion -> whole-leaf -> global), filled
# row by row into the 3-wide facet. Readable labels replace the variable names.
# Ordered by region so related features sit together: CDA (lesion) measures,
# then whole-leaf measures, then the CDA-vs-leaf contrast + global measures.
# The two entropy features (CDA-leaf entropy difference and Shannon entropy) are
# placed adjacent in the final group.
feat_order <- c(
  "cda_green", "cda_bright", "cda_green_sd", "cda_bright_sd",
  "mean_green", "leaf_bright", "leaf_green_sd", "sep", "log_gr",
  "green_gap", "ent_gap", "entropy_raw"
)
feat_labels <- c(
  "CDA greenness", "CDA brightness", "Within-CDA greenness variation", "Within-CDA brightness variation",
  "Whole-leaf mean green level", "Whole-leaf brightness", "Whole-leaf greenness spread",
  "Whole-leaf greenness bimodality", "Log green-to-red ratio",
  "CDA-leaf greenness contrast", "CDA-leaf entropy difference", "Shannon entropy"
)

long6 <- feat6 %>%
  select(image_id, score_f, all_of(feat_order)) %>%
  pivot_longer(all_of(feat_order), names_to = "feature", values_to = "value") %>%
  mutate(feature = factor(feature, levels = feat_order, labels = feat_labels))

p_features <- ggplot(long6, aes(x = score_f, y = value, fill = score_f)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.15) +
  facet_wrap(~feature, ncol = 3, scales = "free_y") +
  scale_fill_manual(values = sev_pal, guide = "none") +
  labs(x = "Severity score", y = NULL) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 10), axis.title = element_text(size = 13),
    strip.text = element_text(size = 11, face = "bold")
  )
p_features
save_plot(p_features, "seg_features_by_score.png", width = 11, height = 12)
```

### 5.1 Feature exemplars: low, medium and high

- Within score class 3 alone, CDA greenness spans -0.16 to +0.04 between the 2nd
  and 98th percentile, over half the full-dataset spread of -0.32 to +0.05.
- That within-class spread is why crops sharing a score look so different, and why
  a single feature threshold cannot recover the score.

```{r fig.width=14, fig.height=14.667, dpi=150}
#| fig-width: 14
#| fig-height: 14.667
#| fig-dpi: 150
# Display device matches the saved png() below (2100x2200 px at res 150), so the
# figure renders here exactly as it is saved.
# 6.1b  Low / median / high example crop for each feature.
feat6_ex <- features_all

qs <- c(Low = 0.02, Medium = 0.5, High = 0.98)
pct_lab <- c("2nd percentile", "50th percentile", "98th percentile")
# All examples are drawn from a single median-score class, so the low/med/high
# comparison isolates within-feature variation from between-class differences.
target_class <- 3
# CDA-region features: the blue CDA outline is drawn only on these rows, where
# the feature is measured over the segmented CDA region.
cda_feats <- c(
  "ent_gap", "cda_green", "cda_bright",
  "cda_green_sd", "cda_bright_sd", "green_gap", "sep"
)

pick_examples <- function(fname) {
  d <- feat6_ex[, c("image_id", "score", fname)]
  names(d)[3] <- "v"
  d <- d[d$score == target_class & is.finite(d$v), ]
  targ <- quantile(d$v, qs, names = FALSE)
  do.call(rbind, lapply(seq_along(targ), function(j) {
    i <- which.min(abs(d$v - targ[j]))
    data.frame(image_id = d$image_id[i], score = d$score[i], value = d$v[i])
  }))
}

draw_feat_examples <- function() {
  half <- length(feat_order) / 2
  # Two side-by-side groups of 6 features: each a label column + Low/Med/High
  # crops, with a header row on top. Fully cellular layout so the middle label
  # column sits between the two groups.
  layout(matrix(seq_len((half + 1) * 8), nrow = half + 1, byrow = TRUE),
    widths = c(1.2, 1, 1, 1, 1.2, 1, 1, 1),
    heights = c(0.75, rep(1, half))
  )
  header <- function(cc) {
    par(mar = c(0, 0, 0, 0))
    plot.new()
    text(0.5, 0.5, names(qs)[cc], font = 2, cex = 2.4)
    text(0.5, 0.28, paste0("(", pct_lab[cc], ")"), cex = 1.85)
  }
  blank <- function() {
    par(mar = c(0, 0, 0, 0))
    plot.new()
  }
  # Feature name right-aligned so it sits against its crops, trimming the gap.
  lab_cell <- function(idx) {
    par(mar = c(0, 0, 0, 0))
    plot.new()
    text(0.97, 0.5, paste(strwrap(feat_labels[idx], width = 12), collapse = "\n"),
      adj = c(1, 0.5), cex = 1.95
    )
  }
  # xaxs/yaxs = "i" + rasterImage fills the plot region exactly (no axis
  # padding); the region is square, so the crop is not stretched. The value at
  # line 0.3 sits just above its own image, while the large top margin (4.2)
  # keeps the separating gap above the value, tying it to the image below.
  crop_cell <- function(idx) {
    ex <- pick_examples(feat_order[idx])
    do_outline <- feat_order[idx] %in% cda_feats
    for (j in 1:3) {
      par(mar = c(0.2, 0.15, 4.2, 0.15))
      img <- load_img(file.path(data_dir, ex$score[j], paste0(ex$image_id[j], ".tif")))
      if (do_outline) img <- outline(img, segment_cda(img)$mask)
      plot(c(0, 1), c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "", xaxs = "i", yaxs = "i")
      rasterImage(as.raster(img), 0, 0, 1, 1)
      mtext(format(signif(ex$value[j], 3), scientific = FALSE, drop0trailing = TRUE, trim = TRUE),
        side = 3, line = 0.3, cex = 1.5
      )
    }
  }
  # Header row: blank | Low Med High | blank | Low Med High
  blank()
  for (cc in 1:3) header(cc)
  blank()
  for (cc in 1:3) header(cc)
  # Feature rows: left group (1..half) beside right group (half+1..2*half)
  for (r in seq_len(half)) {
    lab_cell(r)
    crop_cell(r)
    lab_cell(r + half)
    crop_cell(r + half)
  }
}
png(plot_output_path("10_segmentation_and_feature_extraction", "thesis_plots", "seg_feature_exemplars.png"),
  width = 2100, height = 2200, res = 150
)
draw_feat_examples()
invisible(dev.off())
draw_feat_examples()
```

### 5.2 How each feature relates to score, and to the others

Both summaries use Spearman (rank) correlation on the full unbalanced dataset, so
a feature counts as informative if it is monotonically related to score.

- CDA greenness, the log green-to-red ratio and greenness bimodality lead at
  rho = 0.84, 0.84 and 0.83, so lesion colour carries most of the severity signal.
- Within-CDA greenness variation is flat at rho = 0.03, so how mottled a lesion is
  says nothing about how severe it is.

```{r fig.width=9, fig.height=4.5}
#| fig-width: 9
#| fig-height: 4.5
# 6.2  Correlation of each feature with severity score (Spearman, full dataset).
# Shown standalone here and reused as panel B of the features-by-score figure.
score_cor <- data.frame(
  feature = feat_labels,
  rho = vapply(feat_order, function(f) cor(feat6[[f]], feat6$score, method = "spearman"), numeric(1))
)
p_scorecor <- ggplot(score_cor, aes(x = reorder(feature, -rho), y = rho, fill = rho > 0)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.2f", rho), vjust = ifelse(rho > 0, -0.4, 1.3)), size = 3) +
  scale_fill_manual(values = c(`TRUE` = "firebrick", `FALSE` = "steelblue"), guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0.12, 0.12))) +
  labs(x = NULL, y = "Spearman correlation with score") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 40, hjust = 1, size = 10),
    axis.text.y = element_text(size = 11), axis.title = element_text(size = 12)
  )
p_scorecor
```

- On neighbouring score pairs even the best features reach only AUC 0.68, so the
  strongest feature is wrong about a third of the time on exactly the comparisons
  raters confuse.
- Within-CDA greenness variation sits at AUC 0.49, which is chance, so a feature's
  marginal correlation overstates how well it separates adjacent classes.
- The segmented area fraction was dropped as a model feature because Otsu splits
  every leaf at roughly half its area whatever the score (rho = -0.1), and the
  CDA-leaf entropy difference replaced it at rho = 0.32.

```{r fig.width=9, fig.height=4.5}
#| fig-width: 9
#| fig-height: 4.5
# 6.2b  Per-feature class separability.
# Shown standalone here and reused as panel C of the features-by-score figure.
#
# Separability is summarised as the mean adjacent-class AUC: for each pair of
# neighbouring scores (0-1, 1-2, ... 5-6), the probability that a CDA from the
# higher score has the larger feature value (ties count half), averaged over the
# six pairs. Adjacent pairs are the ones raters actually confuse (off-by-one
# disagreement), so this targets the separability that matters rather than being
# flattered by easily-separated far-apart scores. The general principle of
# averaging pairwise AUC over multiple classes follows Hand & Till (2001); 0.5
# marks no separability and 1 fully separated neighbours.
auc_pair <- function(hi, lo) {
  r <- rank(c(hi, lo))
  (sum(r[seq_along(hi)]) - length(hi) * (length(hi) + 1) / 2) /
    (length(hi) * length(lo))
}
adjacent_auc <- function(x, score) {
  ok <- is.finite(x)
  x <- x[ok]
  score <- score[ok]
  cls <- sort(unique(score))
  mean(vapply(seq_len(length(cls) - 1), function(k) {
    auc_pair(x[score == cls[k + 1]], x[score == cls[k]])
  }, numeric(1)))
}
score_auc <- data.frame(
  feature = feat_labels,
  auc = vapply(feat_order, function(f) adjacent_auc(feat6[[f]], feat6$score), numeric(1))
)
# Panel C shares panel B's feature order (descending Spearman rho) so the two bar
# rows line up under one shared set of x-axis labels.
feat_by_rho <- as.character(score_cor$feature[order(-score_cor$rho)])
score_auc$feature <- factor(score_auc$feature, levels = feat_by_rho)
p_auc <- ggplot(score_auc, aes(x = feature, y = auc)) +
  geom_col(fill = "#35978F") +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey40") +
  geom_text(aes(label = sprintf("%.2f", auc)), vjust = -0.4, size = 3) +
  scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Mean adjacent-class AUC (separability)") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 40, hjust = 1, size = 10),
    axis.text.y = element_text(size = 11), axis.title = element_text(size = 12)
  )
p_auc
```

```{r}
#| output: false
# 6.2c  Combined thesis figure: boxplots (panel A) over the Spearman correlations
# (panel B, x labels dropped since panel C carries them) over the separability
# AUC (panel C). Saved to thesis_plots.
p_scorecor_stk <- p_scorecor + theme(axis.text.x = element_blank())
p_features_abc <- (p_features / p_scorecor_stk / p_auc) +
  plot_layout(heights = c(3.4, 1, 1.25)) +
  plot_annotation(tag_levels = "A") &
  # size 30 so the tags render at ~16 pt on the thesis page (the 11 in figure is
  # scaled to ~149 mm there, a factor of ~0.53); Arial Bold matches the template.
  theme(plot.tag = element_text(size = 30, face = "bold", family = "Arial"))
save_plot(p_features_abc, "features_by_score.png", width = 11, height = 16, dir = "thesis_plots")
```

- The CDA-leaf greenness contrast and whole-leaf greenness spread correlate at
  rho = 0.97, and CDA greenness with the log green-to-red ratio at rho = 0.92, so
  twelve columns carry far fewer than twelve independent measurements.
- The CDA-leaf entropy difference and within-CDA greenness variation join the
  dendrogram last, so they are the only two features adding information the
  colour and brightness measures miss.

```{r fig.width=9.5, fig.height=10}
#| fig-width: 9.5
#| fig-height: 10
# 6.3  Feature-feature correlation matrix with a similarity dendrogram. Spearman,
# all-vs-all; rows/columns are ordered by hierarchical clustering (average linkage
# on 1 - rho) so redundant features group together, and the dendrogram above shows
# the redundancy structure.
cormat <- cor(as.matrix(feat6[, feat_order]), method = "spearman")
dimnames(cormat) <- list(feat_labels, feat_labels)
hc <- hclust(as.dist(1 - cormat), method = "average")
ord_labels <- feat_labels[hc$order]

# Dendrogram line segments straight from the hclust object (ggdendro is not
# installed): each merge draws two verticals (up from each child to the merge
# height) and one horizontal joining them.
dendro_segments <- function(hc) {
  n <- length(hc$order)
  xpos <- match(seq_len(n), hc$order) # leaf i sits at this x position
  cx <- cy <- numeric(nrow(hc$merge))
  node <- function(i) if (i < 0) list(x = xpos[-i], y = 0) else list(x = cx[i], y = cy[i])
  segs <- vector("list", nrow(hc$merge))
  for (k in seq_len(nrow(hc$merge))) {
    a <- node(hc$merge[k, 1])
    b <- node(hc$merge[k, 2])
    h <- hc$height[k]
    segs[[k]] <- data.frame(
      x = c(a$x, b$x, a$x), xend = c(a$x, b$x, b$x),
      y = c(a$y, b$y, h), yend = c(h, h, h)
    )
    cx[k] <- (a$x + b$x) / 2
    cy[k] <- h
  }
  do.call(rbind, segs)
}
dseg <- dendro_segments(hc)

cor_long <- as.data.frame(as.table(cormat))
names(cor_long) <- c("x", "y", "rho")
cor_long$x <- factor(cor_long$x, levels = ord_labels)
cor_long$y <- factor(cor_long$y, levels = rev(ord_labels))

p_heat <- ggplot(cor_long, aes(x, y, fill = rho)) +
  geom_tile(colour = "grey92") +
  geom_text(aes(label = sprintf("%.2f", rho), colour = abs(rho) > 0.6), size = 3) +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B",
    midpoint = 0, limits = c(-1, 1), name = "Spearman rho"
  ) +
  scale_colour_manual(values = c(`FALSE` = "grey15", `TRUE` = "white"), guide = "none") +
  scale_x_discrete(expand = expansion(add = 0.5)) +
  scale_y_discrete(expand = expansion(add = 0.5)) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    panel.grid = element_blank(),
    legend.position = "bottom", legend.key.width = unit(2, "cm")
  )
# Dendrogram shares the heatmap's x range (leaves at 1..n, expand add 0.5) so
# patchwork aligns the two panels column-for-column.
p_dendro <- ggplot(dseg) +
  geom_segment(aes(x = x, xend = xend, y = y, yend = yend), linewidth = 0.4) +
  scale_x_continuous(expand = expansion(add = 0.5)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = NULL, y = NULL) +
  theme_void()

p_cormat <- p_dendro / p_heat + plot_layout(heights = c(1, 6))
p_cormat
save_plot(p_cormat, "feature_correlation_matrix.png", width = 9.5, height = 10, dir = "thesis_plots")
```

---

## 6. PCA of the image content, with and without the background

The same PCA as `07_visual_complexity`, repeated on images with everything
outside the segmented CDA set to black. Both runs use the same class-balanced
sample and the same 64 x 64 normalised inputs, so the only difference is whether
the background is masked.

```{r}
# 8.1  Build original and CDA-masked 64x64 inputs and run PCA (cached: only the
# small PC scores are stored, not the image matrices).
pca_cache <- here("analyses", "10_segmentation_and_feature_extraction", "segmentation_cache", "masked_pca.rds")
if (file.exists(pca_cache)) {
  pca_res <- readRDS(pca_cache)
} else {
  set.seed(42)
  n_per <- min(vapply(0:6, function(s) {
    length(list.files(file.path(data_dir, s), pattern = "\\.tif$"))
  }, integer(1)))
  pca_files <- unlist(lapply(0:6, function(s) {
    sample(list.files(file.path(data_dir, s), pattern = "\\.tif$", full.names = TRUE), n_per)
  }))
  y_pca <- rep(0:6, each = n_per)
  Xo <- matrix(0, length(pca_files), 64 * 64 * 3)
  Xm <- Xo
  for (i in seq_along(pca_files)) {
    img <- load_img(pca_files[i])
    Xo[i, ] <- as.vector(EBImage::resize(img, 64, 64))
    mask <- segment_cda(img)$mask
    mm <- img
    for (k in 1:3) {
      ch <- mm[, , k]
      ch[!mask] <- 0
      mm[, , k] <- ch
    }
    Xm[i, ] <- as.vector(EBImage::resize(mm, 64, 64))
  }
  Xo <- pmin(pmax(Xo, 0), 1)
  Xm <- pmin(pmax(Xm, 0), 1)
  po <- prcomp(Xo, center = TRUE, rank. = 3)
  pm <- prcomp(Xm, center = TRUE, rank. = 3)
  eta_by_pc <- function(x) {
    sapply(1:2, function(j) {
      v <- x[, j]
      gr <- mean(v)
      sum(tapply(v, y_pca, function(z) length(z) * (mean(z) - gr)^2)) / sum((v - gr)^2)
    })
  }
  pca_res <- list(
    score = y_pca,
    orig = po$x[, 1:2], masked = pm$x[, 1:2],
    orig_var = summary(po)$importance[2, 1:2], masked_var = summary(pm)$importance[2, 1:2],
    orig_eta = eta_by_pc(po$x), masked_eta = eta_by_pc(pm$x)
  )
  saveRDS(pca_res, pca_cache)
}
cat(sprintf(
  "PC1/PC2 variance explained: original %.1f%%/%.1f%%, masked %.1f%%/%.1f%%\n",
  100 * pca_res$orig_var[1], 100 * pca_res$orig_var[2],
  100 * pca_res$masked_var[1], 100 * pca_res$masked_var[2]
))
cat(sprintf(
  "eta^2 (PC ~ score; higher = more ordered): PC1 %.3f -> %.3f, PC2 %.3f -> %.3f  (original -> masked)\n",
  pca_res$orig_eta[1], pca_res$masked_eta[1], pca_res$orig_eta[2], pca_res$masked_eta[2]
))
```

```{r fig.width=11, fig.height=5}
#| fig-width: 11
#| fig-height: 5
# 8.2
pca_panel <- function(pc, var, title) {
  data.frame(
    PC1 = pc[, 1], PC2 = pc[, 2],
    score = factor(pca_res$score, levels = 0:6)
  ) %>%
    ggplot(aes(PC1, PC2, colour = score)) +
    geom_point(alpha = 0.5, size = 0.8) +
    scale_colour_manual(name = "Severity Score", values = sev_pal) +
    guides(colour = guide_legend(override.aes = list(alpha = 1, size = 2))) +
    labs(
      title = title, x = sprintf("PC1 (%.1f%%)", 100 * var[1]),
      y = sprintf("PC2 (%.1f%%)", 100 * var[2])
    ) +
    theme_minimal() +
    theme(
      axis.text = element_text(size = 11), axis.title = element_text(size = 12),
      plot.title = element_text(size = 12), legend.title = element_text(size = 12)
    )
}
pca_left <- pca_panel(pca_res$orig, pca_res$orig_var, "Original (background included)")
pca_right <- pca_panel(pca_res$masked, pca_res$masked_var, "Masked to the segmented CDA")
p_pca <- (pca_left | pca_right) + plot_layout(guides = "collect")
p_pca
save_plot(p_pca, "seg_pca_masked_vs_original.png", width = 11, height = 5)
```

- Masking the background raises PC1's alignment with score from eta-squared 0.27
  to 0.37, so the severity signal concentrates onto one component once the leaf
  and imaging artefacts are gone.
- PC1's share of variance falls from 31.0% to 21.7%, so the masked components
  explain less of the image while explaining more of the score.

---

## 7. Interpretation and Discussion

- Greenness solves the universal-threshold problem. Otsu's brightness cut moves
  from 66 to 137 of 255 between two crops while its greenness cut holds at -0.16
  to -0.15, so changing channel is what removes the per-image tuning that defeated
  the earlier brightness watershed.
- Presence detection works, so segmentation should be conditional. Leaf-greenness
  bimodality separates 0 from non-0 at AUC 0.944, yet the spatial segmentation
  still carves 49% of a score-0 leaf into a spurious lesion, so it should run only
  after the non-zero decision.
- The features track severity without separating neighbouring scores. Three
  features reach rho = 0.84 against score, but the best adjacent-class AUC is
  0.68, so the ordering is recoverable while the exact class is not.
- The feature set is redundant. Near-duplicate pairs reach rho = 0.97, so the
  twelve columns carry far fewer than twelve independent measurements.
- What the features can predict is answered in `10b_score_prediction.qmd`, where
  they place a CDA within one point of its score almost always but pin the exact
  score about 60% of the time.
- Low scores remain hard. Median CDA greenness at scores 1-2 (-0.13 and -0.08) is
  barely above the score-0 value of -0.19, so mild cell death is genuinely close
  to healthy tissue in these images.
