---
title: "Segmenting cell death area from leaf background"
format:
  html:
    code-fold: true
---

Date: 13/7/2026

## GOHREP

**Goal:**

Test whether the cell death area (CDA) of a leaf can be separated from the
healthy leaf background *universally* — with one method and no per-image manual
threshold tuning — using the distribution of pixel values.

**Hypothesis:**

An earlier attempt (`PhD_DL_Models/analyses/0009_Non_DL_Green_Channel_Analysis`)
watershedded on image **brightness/grayscale** and found the threshold had to be
retuned per image. The reason is colour: under this UV-fluorescence imaging the
healthy leaf is **red** (high R, low G) and cell death is **yellow-green**
(G ≈ or > R), and red leaf and green CDA have *similar brightness*. Segmenting
on a **greenness** channel (G − R) instead should separate CDA from leaf with a
threshold that Otsu can set automatically per image, because the two tissues
form distinct modes in greenness even when they do not in brightness.

**Rationale:**

Otsu's method finds the threshold that best splits a bimodal distribution. If
CDA and leaf are two modes in greenness, Otsu adapts to each image with no
manual input. Where there is no CDA (median-0 leaves), there is only one mode,
so segmentation is only meaningful once an image is known to be non-zero — hence
0 and non-0 are treated separately here.

---

## 1. Setup

### 1.1 Load required packages

```{r}
#| output: false
# 1.1
library(tiff)
suppressMessages(library(EBImage))
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
library(here)
library(RColorBrewer)
# randomForest and MASS are called namespaced (randomForest::randomForest,
# MASS::polr) rather than attached, so they do not mask dplyr::select etc.
requireNamespace("randomForest", quietly = TRUE)
requireNamespace("MASS", quietly = TRUE)
```

### 1.2 Helper functions and colour palettes

```{r}
# 1.2
source(here("src", "plotting_helpers.R"))
save_plot <- make_save_plot("10_segmentation_and_feature_extraction")
data_dir <- here("data", "cropped_images_medoid")
```

### 1.3 Segmentation functions

```{r}
# 1.3
load_img <- function(path) {
  img <- readTIFF(path, info = FALSE)
  if (length(dim(img)) == 2) img <- array(img, dim = c(dim(img), 3))
  img[, , 1:3, drop = FALSE]
}

# Leaf mask: non-black pixels (mean RGB >= 0.15), dropping the near-black
# background and holes. Analysis 09_image_entropy (§5.5) uses a comparable
# non-dark threshold on peak brightness.
leaf_mask <- function(img) (img[, , 1] + img[, , 2] + img[, , 3]) / 3 >= 0.15

# Greenness channel: green minus red. Healthy (red) leaf is strongly negative;
# yellow-green CDA is near zero or positive.
greenness <- function(img) img[, , 2] - img[, , 1]

# Otsu threshold on a value vector, plus the "separability" = between-class
# variance at the threshold / total variance (a 2-class eta^2 in [0,1]). High
# separability => the values are well split into two modes (bimodal).
otsu_sep <- function(x, nbins = 64) {
  rng <- range(x)
  if (diff(rng) < 1e-6) {
    return(c(thr = rng[1], sep = 0, frac_hi = 0))
  }
  br <- seq(rng[1], rng[2], length.out = nbins + 1)
  mids <- (br[-1] + br[-length(br)]) / 2
  h <- tabulate(cut(x, br, include.lowest = TRUE), nbins)
  p <- h / sum(h)
  mu <- sum(p * mids)
  tot <- sum(p * (mids - mu)^2)
  w1 <- cumsum(p)
  m1 <- cumsum(p * mids)
  bcv <- (mu * w1 - m1)^2 / (w1 * (1 - w1) + 1e-12)
  k <- which.max(bcv)
  c(thr = br[k + 1], sep = max(bcv) / (tot + 1e-12), frac_hi = mean(x > br[k + 1]))
}

# Full spatial segmentation of one image: Otsu on the (blurred) greenness within
# the leaf, followed by morphological cleanup.
segment_cda <- function(img, sigma = 2) {
  leaf <- leaf_mask(img)
  g <- greenness(img)
  gs <- (g - min(g)) / (max(g) - min(g) + 1e-9) # rescale to 0-1 for EBImage
  gb <- EBImage::gblur(gs, sigma = sigma)
  thr <- EBImage::otsu(matrix(gb[leaf]), range = c(0, 1))
  mask <- leaf & (gb > thr)
  # Clean up with a median filter, whose radius scales with image size so the
  # smoothing is constant across crop resolutions: base 5 px at the 264-px median
  # crop, applied at radius 2*base. A median filter removes thin protrusions
  # (veins, spurs) far more effectively than a morphological opening of the same
  # scale while barely shrinking the bulk; see the cleanup comparison in section
  # 3.3 for the circularity/area justification of this choice.
  b <- max(1, round(5 * mean(dim(img)[1:2]) / 264))
  mask <- EBImage::fillHull(EBImage::medianFilter(mask, 2 * b))
  # The median filter and fillHull ignore the leaf boundary, so they can bridge
  # across background notches and pull off-leaf (near-black) pixels into the mask.
  # Re-apply the leaf mask so the CDA can never extend onto the background.
  mask <- mask & leaf
  # Keep only the largest connected component, dropping small floating islands.
  lab <- EBImage::bwlabel(mask)
  if (max(lab) > 1) mask <- lab == which.max(tabulate(lab[lab > 0]))
  # Re-applying the leaf mask above punches holes back into the CDA wherever the
  # leaf is dark inside the lesion (veins, specks below the 0.15 brightness cut).
  # Fill those enclosed holes so the CDA is a single solid region with no interior
  # dots in its outline. fillHull only fills fully-enclosed holes, so it cannot
  # re-extend the mask back onto the background.
  mask <- EBImage::fillHull(mask)
  list(leaf = leaf, greenness = gb, mask = mask, cda_frac = mean(mask))
}

# Render a single-channel raster (e.g. the greenness map) as RGB, recolouring
# pixels outside `keep` (e.g. off-leaf background) a flat colour. Used so the
# Otsu-excluded background reads visually distinct from the leaf pixels that
# actually feed the threshold, rather than looking like more greenness data.
recolour_outside <- function(chan, keep, colour = c(0.12, 0.56, 1)) {
  out <- array(chan, dim = c(dim(chan), 3))
  for (k in 1:3) {
    ch <- out[, , k]
    ch[!keep] <- colour[k]
    out[, , k] <- ch
  }
  out
}

# Overlay a mask on an image by turning masked pixels yellow-magenta.
overlay <- function(img, mask, colour = c(1, 0, 1)) {
  out <- img
  for (k in 1:3) {
    ch <- out[, , k]
    ch[mask] <- colour[k]
    out[, , k] <- ch
  }
  out
}

# Bold column headers (top) and rotated row labels (left) for the base-graphics
# example grids, drawn once into the outer margins instead of per-panel titles.
colhead <- function(labels, ncol, cex = 1) {
  for (c in seq_along(labels)) {
    mtext(labels[c],
      side = 3, outer = TRUE, line = 0.3, at = (c - 0.5) / ncol,
      font = 2, cex = cex
    )
  }
}
rowlab <- function(label, r, nrow, cex = 1) {
  mtext(label,
    side = 2, outer = TRUE, las = 0, line = 1,
    at = 1 - (r - 0.5) / nrow, cex = cex
  )
}

# Draw a coloured outline of a mask onto an image (a boundary band of the given
# half-width), leaving the interior visible - clearer than a filled overlay.
outline <- function(img, mask, colour = c(0.12, 0.56, 1), width = 2) {
  br <- EBImage::makeBrush(2 * width + 1, "disc")
  edge <- EBImage::dilate(mask, br) & !EBImage::erode(mask, br)
  out <- img
  for (k in 1:3) {
    ch <- out[, , k]
    ch[edge] <- colour[k]
    out[, , k] <- ch
  }
  out
}

# Shannon entropy of pixel values in [0, 1], matching 09_image_entropy's
# shannon_entropy(): quantise to 8-bit intensities and take H = -sum(p*log2 p).
# region_entropy() applies it over the pixels of a mask, flattened across the
# three RGB channels, exactly as the whole-image version does.
shannon_entropy_vec <- function(v) {
  q <- as.integer(round(pmin(pmax(v, 0), 1) * 255))
  counts <- tabulate(q + 1L, nbins = 256L)
  p <- counts[counts > 0] / length(q)
  -sum(p * log2(p))
}
region_entropy <- function(img, mask) {
  if (!any(mask)) {
    return(NA_real_)
  }
  shannon_entropy_vec(c(img[, , 1][mask], img[, , 2][mask], img[, , 3][mask]))
}
```

---

## 2. Why brightness fails and greenness works

For one low- and one high-score CDA, the brightness of the leaf and the CDA
overlap (a single smeared mode), while in greenness the CDA and leaf separate
into two modes. This is why an Otsu/watershed threshold on brightness cannot be
set once and reused, but one on greenness can.

```{r fig.width=10, fig.height=5.4}
#| fig-width: 10
#| fig-height: 5.4
# 2.1
set.seed(2)
demo <- c(
  sample(list.files(file.path(data_dir, 2), pattern = "\\.tif$", full.names = TRUE), 1),
  sample(list.files(file.path(data_dir, 6), pattern = "\\.tif$", full.names = TRUE), 1)
)
demo_scores <- c(2, 6)
draw_channels <- function() {
  par(mfrow = c(2, 3), mar = c(3.2, 3.4, 1, 1), mgp = c(2, 0.6, 0), oma = c(0, 3, 2.4, 0))
  for (r in seq_along(demo)) {
    img <- load_img(demo[r])
    leaf <- leaf_mask(img)
    bright <- ((img[, , 1] + img[, , 2] + img[, , 3]) / 3)[leaf]
    g <- greenness(img)[leaf]
    plot(as.raster(img))
    hist(bright * 255,
      breaks = 40, col = "grey70", main = "",
      xlab = "Brightness (0-255)", ylab = "Pixels"
    )
    hist(g,
      breaks = 40, col = "darkolivegreen3", main = "",
      xlab = "Greenness (G - R)", ylab = "Pixels"
    )
    abline(v = otsu_sep(g)["thr"], col = "red", lwd = 2)
    rowlab(sprintf("Severity Score %d", demo_scores[r]), r, 2)
  }
  colhead(c("Crop", "Whole-leaf brightness", "Whole-leaf greenness (G - R)"), 3, 0.95)
}
png(plot_output_path("10_segmentation_and_feature_extraction", "supp_plots", "seg_brightness_vs_greenness.png"),
  width = 1500, height = 800, res = 130
)
draw_channels()
invisible(dev.off())
draw_channels()
```

---

## 3. Per-image segmentation via Otsu on greenness

One CDA per score (including a median-0 leaf). Columns: crop, blurred greenness,
Otsu CDA overlay, and the leaf-greenness histogram with the Otsu threshold. For
non-zero scores the histogram is bimodal and Otsu lands in the valley, cleanly
isolating the CDA. For score 0 there is no true CDA, yet Otsu still forces a
split — a spurious "CDA" on healthy tissue (see §4).

```{r fig.width=10, fig.height=10}
#| fig-width: 10
#| fig-height: 10
# 3.1
set.seed(7)
seg_scores <- c(0, 2, 4, 6)
seg_files <- sapply(seg_scores, function(sc) {
  sample(list.files(file.path(data_dir, sc), pattern = "\\.tif$", full.names = TRUE), 1)
})

draw_seg <- function() {
  par(
    mfrow = c(length(seg_files), 4), mar = c(2.4, 2, 0.6, 0.6), mgp = c(1.3, 0.4, 0),
    oma = c(0, 2.6, 2.4, 0)
  )
  for (i in seq_along(seg_files)) {
    img <- load_img(seg_files[i])
    s <- segment_cda(img)
    g <- greenness(img)[s$leaf]
    plot(as.raster(img))
    plot(as.raster(recolour_outside(s$greenness, s$leaf)))
    hist(g, breaks = 40, col = "darkolivegreen3", main = "", xlab = "Greenness (G - R)", ylab = "Frequency")
    abline(v = otsu_sep(g)["thr"], col = "red", lwd = 2)
    plot(as.raster(outline(img, s$mask)))
    rowlab(sprintf("Score %d", seg_scores[i]), i, length(seg_scores))
  }
  colhead(c("Crop", "Greenness (G - R)", "Greenness histogram", "Segmented CDA"), 4, 0.9)
}
png(plot_output_path("10_segmentation_and_feature_extraction", "supp_plots", "seg_otsu_examples.png"),
  width = 1500, height = 1500, res = 140
)
draw_seg()
invisible(dev.off())
draw_seg()
```

### 3.1 Segmentation of six CDAs per severity score

To check that the cell-death-area selection is correct across many images, six
randomly sampled CDAs per severity score (0–6) are shown below. The segmented CDA
boundary (from the Otsu-on-greenness spatial segmentation of §3, after
morphological cleanup) is drawn in blue on the original crop, so both the leaf
and the selected region are visible. Rows are severity scores; the six crops in a
row are independent random samples.

```{r fig.width=11, fig.height=12}
#| fig-width: 11
#| fig-height: 12
# 3.1
set.seed(11)
n_per <- 6
grid_scores <- 0:6
grid_files <- lapply(grid_scores, function(sc) {
  sample(list.files(file.path(data_dir, sc), pattern = "\\.tif$", full.names = TRUE), n_per)
})

draw_seg_grid <- function() {
  par(
    mfrow = c(length(grid_scores), n_per), mar = c(0.15, 0.15, 0.15, 0.15),
    oma = c(0, 3.2, 0, 0)
  )
  for (r in seq_along(grid_scores)) {
    for (f in grid_files[[r]]) {
      img <- load_img(f)
      plot(as.raster(outline(img, segment_cda(img)$mask)))
    }
    mtext(sprintf("Score %d", grid_scores[r]),
      side = 2, outer = TRUE, las = 0,
      line = 1, cex = 1, at = 1 - (r - 0.5) / length(grid_scores)
    )
  }
}
png(plot_output_path("10_segmentation_and_feature_extraction", "supp_plots", "seg_validation_grid.png"),
  width = 2000, height = 2000, res = 150
)
draw_seg_grid()
invisible(dev.off())
draw_seg_grid()
```

### 3.2 Otsu vs background-only leaf masks

The spatial segmentation above isolates the cell-death area. A related but
distinct question is which pixels count as *leaf* when estimating per-image
lighting (used in the entropy analysis, `09_image_entropy`, §5.5). Two thresholds
on each CDA's own brightness histogram are compared: the image-adaptive **Otsu**
threshold, and a low fixed **leaf** cut (brightness >= 0.1) that removes only the
near-black background and keeps all tissue. For one CDA per score, the original
crop, the Otsu mask, and the background-only leaf mask are shown.

```{r fig.width=9, fig.height=11}
#| fig-width: 9
#| fig-height: 11
# 3.2
set.seed(7)
leaf_mask_examples <- lapply(c(0, 2, 4, 6), function(sc) {
  files <- list.files(file.path(data_dir, sc), pattern = "\\.tif$", full.names = TRUE)
  f <- sample(files, 1)
  img <- readTIFF(f, info = FALSE)
  if (length(dim(img)) == 2) img <- array(img, dim = c(dim(img), 3))
  img <- img[, , 1:3, drop = FALSE]
  bright_px <- (img[, , 1] + img[, , 2] + img[, , 3]) / 3
  thr <- EBImage::otsu(bright_px, range = c(0, 1))
  otsu_mask <- bright_px > thr
  if (sum(otsu_mask) < 0.02 * length(otsu_mask)) otsu_mask <- bright_px >= 0.1
  leaf_mask <- bright_px >= 0.10
  list(
    sc = sc, img = img, otsu = otsu_mask, leaf = leaf_mask,
    otsu_frac = mean(otsu_mask), leaf_frac = mean(leaf_mask)
  )
})

mask_out <- function(img, mask) {
  out <- img
  for (k in 1:3) {
    ch <- out[, , k]
    ch[!mask] <- 0
    out[, , k] <- ch
  }
  out
}
draw_leaf_masks <- function() {
  par(mfrow = c(length(leaf_mask_examples), 3), mar = c(0.5, 2, 2.2, 0.5))
  for (e in leaf_mask_examples) {
    plot(as.raster(e$img))
    title(sprintf("score %d: crop", e$sc))
    plot(as.raster(mask_out(e$img, e$otsu)))
    title(sprintf("Otsu region (%.0f%%)", 100 * e$otsu_frac))
    plot(as.raster(mask_out(e$img, e$leaf)))
    title(sprintf("leaf region (%.0f%%)", 100 * e$leaf_frac))
  }
}
png(plot_output_path("10_segmentation_and_feature_extraction", "supp_plots", "leaf_segmentation_examples.png"),
  width = 1500, height = 1800, res = 150
)
draw_leaf_masks()
invisible(dev.off())
draw_leaf_masks()
```

The Otsu mask cleanly removes the black background where there is one (scores 2,
4), but because healthy tissue fluoresces red (darker) while lesions are bright
yellow-green, in higher-score CDAs Otsu selects the **lesion** rather than the
whole leaf; the background-only leaf mask keeps all tissue regardless of score.
This is why the entropy analysis (§5.5) uses background-only leaf brightness,
not Otsu, as its lighting proxy.

### 3.3 Choosing the morphological cleanup step

The raw Otsu mask is ragged: it carries thin vein spurs and a jagged boundary that
the near-circular cell death phenotype does not have, so a smoothing step is applied
before the region features are measured. Several scale-aware smoothers were compared
to choose it, each scored by **circularity** (4πA/P², where 1 is a perfect circle)
and **area retained** relative to the raw mask (1 = no shrinkage), on a sample of
higher-severity crops where spurs are most visible.

A **median filter at radius 2b** was selected: it gives the roundest outlines
(circularity 0.94) while shrinking the CDA the least (98.7% of the area retained). A
morphological opening of comparable scale rounds less for more area loss, because it
erodes the whole boundary rather than only the thin protrusions; a strong boundary
blur over-rounds genuine lobes; and opening-by-reconstruction leaves spurs that are
attached to the bulk untouched (its outline is identical to the raw mask), so it does
not help here. All cleanup radii scale with image size (base b = round(5 · side /
264)), so the smoothing is constant across crop resolutions.

```{r fig.width=12, fig.height=9}
#| fig-width: 12
#| fig-height: 9
# 3.3
odd <- function(n) {
  n <- round(n)
  if (n %% 2 == 0) n <- n + 1
  max(3, n)
}
largest_cc <- function(mask) {
  lab <- EBImage::bwlabel(mask)
  if (max(lab) > 1) mask <- lab == which.max(tabulate(lab[lab > 0]))
  mask
}
# raw thresholded CDA mask, before any morphological cleanup
raw_cda_mask <- function(img, sigma = 2) {
  leaf <- leaf_mask(img)
  g <- greenness(img)
  gs <- (g - min(g)) / (max(g) - min(g) + 1e-9)
  gb <- EBImage::gblur(gs, sigma = sigma)
  thr <- EBImage::otsu(matrix(gb[leaf]), range = c(0, 1))
  leaf & (gb > thr)
}
# geodesic reconstruction by dilation of a marker under a mask
reconstruct <- function(marker, mask, brush) {
  prev <- marker
  for (i in 1:300) {
    cur <- EBImage::dilate(prev, brush) & mask
    if (all(cur == prev)) break
    prev <- cur
  }
  prev
}
cleanups <- list(
  "Raw" = function(m, b, d) largest_cc(EBImage::fillHull(m)),
  "Open x2" = function(m, b, d) {
    largest_cc(EBImage::fillHull(EBImage::opening(m, EBImage::makeBrush(odd(b * 2), "disc"))))
  },
  "Open x3" = function(m, b, d) {
    largest_cc(EBImage::fillHull(EBImage::opening(m, EBImage::makeBrush(odd(b * 3), "disc"))))
  },
  "Median r=b" = function(m, b, d) largest_cc(EBImage::fillHull(EBImage::medianFilter(m, b))),
  "Median r=2b" = function(m, b, d) largest_cc(EBImage::fillHull(EBImage::medianFilter(m, 2 * b))),
  "Blur sig=side/50" = function(m, b, d) largest_cc(EBImage::fillHull(EBImage::gblur(m * 1, sigma = d / 50) > 0.5)),
  "Blur sig=side/30" = function(m, b, d) largest_cc(EBImage::fillHull(EBImage::gblur(m * 1, sigma = d / 30) > 0.5)),
  "Reconstruction" = function(m, b, d) {
    br <- EBImage::makeBrush(odd(b), "disc")
    largest_cc(EBImage::fillHull(reconstruct(EBImage::erode(m, br), m, br)))
  }
)
circularity <- function(mask) {
  if (sum(mask) < 10) {
    return(NA_real_)
  }
  sh <- EBImage::computeFeatures.shape(EBImage::bwlabel(mask))
  if (is.null(sh) || nrow(sh) == 0 || sh[1, "s.perimeter"] <= 0) {
    return(NA_real_)
  }
  4 * pi * sh[1, "s.area"] / (sh[1, "s.perimeter"]^2)
}

set.seed(11)
clean_files <- unlist(lapply(3:6, function(sc) {
  sample(list.files(file.path(data_dir, sc), pattern = "\\.tif$", full.names = TRUE), 4)
}))
circ_m <- area_m <- matrix(NA_real_, length(clean_files), length(cleanups),
  dimnames = list(NULL, names(cleanups))
)
for (i in seq_along(clean_files)) {
  img <- load_img(clean_files[i])
  d <- mean(dim(img)[1:2])
  b <- max(1, round(5 * d / 264))
  rm <- raw_cda_mask(img)
  for (j in seq_along(cleanups)) {
    mk <- cleanups[[j]](rm, b, d)
    circ_m[i, j] <- circularity(mk)
    area_m[i, j] <- mean(mk)
  }
}
cleanup_summary <- data.frame(
  method = names(cleanups),
  circularity = round(colMeans(circ_m, na.rm = TRUE), 3),
  area_retained = round(colMeans(area_m / area_m[, "Raw"], na.rm = TRUE), 3),
  row.names = NULL
)
print(cleanup_summary)

# Visual comparison: rows = crops, columns = cleanup methods.
vis_files <- clean_files[c(1, 3, 5, 7, 9, 11)]
draw_cleanup_grid <- function() {
  par(mfrow = c(length(vis_files), length(cleanups)), mar = c(0.3, 0.3, 1.8, 0.3))
  for (f in vis_files) {
    img <- load_img(f)
    d <- mean(dim(img)[1:2])
    b <- max(1, round(5 * d / 264))
    rm <- raw_cda_mask(img)
    for (j in seq_along(cleanups)) {
      plot(as.raster(outline(img, cleanups[[j]](rm, b, d))))
      if (f == vis_files[1]) title(main = names(cleanups)[j], cex.main = 0.9, line = 0.4)
    }
  }
}
png(plot_output_path("10_segmentation_and_feature_extraction", "supp_plots", "seg_cleanup_comparison.png"),
  width = 2400, height = 1800, res = 150
)
draw_cleanup_grid()
invisible(dev.off())
draw_cleanup_grid()
```

---

## 4. Can the distribution alone detect whether a CDA is present?

Segmenting assumes a CDA exists. The natural universal test for "is there a CDA"
is the bimodality (separability) of the leaf-greenness distribution. The
following computes, for every image, the eleven region features summarised from
the segmentation (§6.1), including the leaf-greenness separability. Computed once
and cached.

```{r}
# 5.1
cache_path <- here(
  "analyses", "10_segmentation_and_feature_extraction", "segmentation_cache",
  "cda_segmentation_metrics.rds"
)

if (file.exists(cache_path)) {
  seg <- readRDS(cache_path)
} else {
  scores <- list.dirs(data_dir, recursive = FALSE, full.names = FALSE)
  scores <- scores[grepl("^[0-9]+$", scores)]
  rows <- list()
  for (sc in scores) {
    files <- list.files(file.path(data_dir, sc), pattern = "\\.tif$", full.names = TRUE)
    m <- t(vapply(files, function(f) {
      # All region features are read off the SAME segmentation: the leaf mask,
      # the spatial CDA mask from segment_cda() (Otsu on blurred greenness ->
      # morphological cleanup -> largest connected component), and the healthy
      # leaf (leaf minus CDA). Nothing is re-thresholded per feature.
      img <- load_img(f)
      s <- segment_cda(img)
      leaf <- s$leaf # leaf region ((R+G+B)/3 >= 0.10)
      cda <- s$mask & leaf # segmented CDA (largest greenness component)
      healthy <- leaf & !cda # healthy leaf = leaf minus CDA
      gmat <- greenness(img) # greenness G - R over the whole image
      bmat <- (img[, , 1] + img[, , 2] + img[, , 3]) / 3 # brightness (mean RGB)
      R <- img[, , 1]
      G <- img[, , 2]
      nsd <- function(x) if (length(x) > 1) sd(x) else 0
      mn <- function(x) if (length(x)) mean(x) else 0

      gv_leaf <- gmat[leaf] # greenness over the whole leaf
      o <- otsu_sep(gv_leaf) # separability (bimodality) of leaf greenness
      n_green <- sum(G[leaf] >= R[leaf])
      n_red <- sum(R[leaf] > G[leaf]) # colour counts over leaf

      c(
        sep = unname(o["sep"]), # 4  leaf-greenness bimodality
        green_gap = if (any(cda) && any(healthy)) { # 3  CDA vs healthy-leaf greenness
          mean(gmat[cda]) - mean(gmat[healthy])
        } else {
          0
        },
        cda_green = mn(gmat[cda]), # 2  CDA greenness
        frac_g_gt_r = sum(cda) / sum(leaf), # 1  lesion size = |CDA| / |leaf|
        cda_green_sd = nsd(gmat[cda]) * 255, # 7  within-CDA greenness spread
        cda_bright = mn(bmat[cda]) * 255, # 5  CDA brightness
        cda_bright_sd = nsd(bmat[cda]) * 255, # 8  within-CDA brightness spread
        leaf_green_sd = nsd(gv_leaf) * 255, # 6  whole-leaf greenness spread
        mean_green = mn(G[leaf]) * 255, # 10 mean green over leaf
        log_gr = log((n_green + 1) / (n_red + 1)), # 9  log green:red count over leaf
        leaf_bright = mn(bmat[leaf]) * 255, # 11 leaf brightness
        ent_gap = if (any(cda) && any(healthy)) { # 12 CDA minus healthy-leaf Shannon entropy
          region_entropy(img, cda) - region_entropy(img, healthy)
        } else {
          0
        }
      )
    }, numeric(12)))
    rows[[sc]] <- data.frame(
      image_id = tools::file_path_sans_ext(basename(files)),
      score = as.integer(sc), m, row.names = NULL
    )
    message(sprintf("score %s done (%d)", sc, length(files)))
  }
  seg <- dplyr::bind_rows(rows)
  saveRDS(seg, cache_path)
}

seg <- seg %>% mutate(
  score_f = factor(score, levels = 0:6),
  non0 = score > 0
)
cat(sprintf("Segmentation metrics for %d images.\n", nrow(seg)))
```

Separability rises with score. Median-0 leaves score lower (their greenness is
closer to unimodal), though not zero — veins, brightness gradients and the
infiltration-site ring still give Otsu something to split. Even so, the gap from
non-0 images is large: the area under the ROC curve for classifying 0 vs non-0
from separability alone is high (§4.2), so the distribution *does* detect CDA
presence well — consistent with 0-vs-non-0 being the easy part of the problem.

```{r}
# 5.2
# Rank-based (Mann-Whitney) AUC: probability a random non-0 image scores higher
# than a random 0 image on the given feature.
auc <- function(feature, positive) {
  r <- rank(feature)
  n_pos <- sum(positive)
  n_neg <- sum(!positive)
  (sum(r[positive]) - n_pos * (n_pos + 1) / 2) / (n_pos * n_neg)
}
cat(sprintf(
  "AUC (0 vs non-0):  separability = %.3f  |  fixed-threshold area (G>R) = %.3f\n",
  auc(seg$sep, seg$non0), auc(seg$frac_g_gt_r, seg$non0)
))
```

```{r fig.width=8, fig.height=5}
#| fig-width: 8
#| fig-height: 5
# 5.3
p_sep <- ggplot(seg, aes(x = score_f, y = sep)) +
  geom_boxplot(aes(fill = score_f), outlier.shape = NA, alpha = 0.85) +
  geom_jitter(width = 0.18, height = 0, alpha = 0.1, size = 0.4) +
  scale_fill_manual(name = "Severity Score", values = sev_pal) +
  labs(x = "Severity score", y = "Leaf-greenness separability (bimodality)") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 13), axis.title = element_text(size = 15),
    legend.title = element_text(size = 13), legend.text = element_text(size = 11)
  )
p_sep
save_plot(p_sep, "seg_separability_by_score.png", width = 8, height = 5)
```

---

## 5. How each feature relates to severity

Each of the twelve features used for prediction (§6) is shown as a boxplot
against median severity score. The eight segmentation features and the four
colour/brightness/entropy features (from `09_image_entropy`) are faceted together
(3 wide × 4 high, with free y-axes since the features are on different scales).
Most rise with severity but overlap heavily between adjacent scores; the
within-CDA variation features (`cda_green_sd`, `cda_bright_sd`) are close to flat.

```{r fig.width=11, fig.height=12}
#| fig-width: 11
#| fig-height: 12
# 6.1
# All region features (1-11) are already in `seg`; only Shannon entropy (12) is
# joined from the entropy analysis so all twelve can be faceted.
feat6 <- seg %>%
  left_join(
    readRDS(here("analyses", "09_image_entropy", "entropy_cache", "cda_metrics.rds")) %>%
      filter(size == "Raw") %>%
      transmute(image_id, entropy_raw = entropy),
    by = "image_id"
  )

# Order groups the features by region (lesion -> whole-leaf -> global), filled
# row by row into the 3-wide facet. Readable labels replace the variable names.
# Ordered by region so related features sit together: CDA (lesion) measures,
# then whole-leaf measures, then the CDA-vs-leaf contrast + global measures.
# The two entropy features (CDA-leaf entropy difference and Shannon entropy) are
# placed adjacent in the final group.
feat_order <- c(
  "cda_green", "cda_bright", "cda_green_sd", "cda_bright_sd",
  "mean_green", "leaf_bright", "leaf_green_sd", "sep", "log_gr",
  "green_gap", "ent_gap", "entropy_raw"
)
feat_labels <- c(
  "CDA greenness", "CDA brightness", "Within-CDA greenness variation", "Within-CDA brightness variation",
  "Whole-leaf mean green level", "Whole-leaf brightness", "Whole-leaf greenness spread",
  "Whole-leaf greenness bimodality", "Log green-to-red ratio",
  "CDA-leaf greenness contrast", "CDA-leaf entropy difference", "Shannon entropy"
)

long6 <- feat6 %>%
  select(image_id, score_f, all_of(feat_order)) %>%
  pivot_longer(all_of(feat_order), names_to = "feature", values_to = "value") %>%
  mutate(feature = factor(feature, levels = feat_order, labels = feat_labels))

p_features <- ggplot(long6, aes(x = score_f, y = value, fill = score_f)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.15) +
  facet_wrap(~feature, ncol = 3, scales = "free_y") +
  scale_fill_manual(values = sev_pal, guide = "none") +
  labs(x = "Severity score", y = NULL) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 10), axis.title = element_text(size = 13),
    strip.text = element_text(size = 11, face = "bold")
  )
p_features
save_plot(p_features, "seg_features_by_score.png", width = 11, height = 12)
```

### 5.1 Feature exemplars: low, medium and high

For each of the twelve features, three median-score-3 CDA crops are shown: at a
low (2nd percentile), median and high (98th percentile) value of that feature.
Restricting to a single severity class isolates each feature's own variation
from differences between score classes, so the low-to-high change reflects the
feature alone. The blue outline marks the segmented CDA on the rows where the
feature is measured over that region; the whole-leaf and global features
(measured over the whole leaf) are drawn without an outline. Each crop is
annotated with its feature value.

```{r fig.width=8, fig.height=8.4}
#| fig-width: 8
#| fig-height: 8.4
# 6.1b  Low / median / high example crop for each feature.
feat6_ex <- seg %>%
  left_join(
    readRDS(here("analyses", "09_image_entropy", "entropy_cache", "cda_metrics.rds")) %>%
      filter(size == "Raw") %>%
      transmute(image_id, entropy_raw = entropy),
    by = "image_id"
  )

qs <- c(Low = 0.02, Medium = 0.5, High = 0.98)
pct_lab <- c("2nd percentile", "50th percentile", "98th percentile")
# All examples are drawn from a single median-score class, so the low/med/high
# comparison isolates within-feature variation from between-class differences.
target_class <- 3
# CDA-region features: the blue CDA outline is drawn only on these rows, where
# the feature is measured over the segmented CDA region.
cda_feats <- c(
  "ent_gap", "cda_green", "cda_bright",
  "cda_green_sd", "cda_bright_sd", "green_gap", "sep"
)

pick_examples <- function(fname) {
  d <- feat6_ex[, c("image_id", "score", fname)]
  names(d)[3] <- "v"
  d <- d[d$score == target_class & is.finite(d$v), ]
  targ <- quantile(d$v, qs, names = FALSE)
  do.call(rbind, lapply(seq_along(targ), function(j) {
    i <- which.min(abs(d$v - targ[j]))
    data.frame(image_id = d$image_id[i], score = d$score[i], value = d$v[i])
  }))
}

draw_feat_examples <- function() {
  half <- length(feat_order) / 2
  # Two side-by-side groups of 6 features: each a label column + Low/Med/High
  # crops, with a header row on top. Fully cellular layout so the middle label
  # column sits between the two groups.
  layout(matrix(seq_len((half + 1) * 8), nrow = half + 1, byrow = TRUE),
    widths = c(1.2, 1, 1, 1, 1.2, 1, 1, 1),
    heights = c(0.75, rep(1, half))
  )
  header <- function(cc) {
    par(mar = c(0, 0, 0, 0))
    plot.new()
    text(0.5, 0.5, names(qs)[cc], font = 2, cex = 2.4)
    text(0.5, 0.28, paste0("(", pct_lab[cc], ")"), cex = 1.85)
  }
  blank <- function() {
    par(mar = c(0, 0, 0, 0))
    plot.new()
  }
  # Feature name right-aligned so it sits against its crops, trimming the gap.
  lab_cell <- function(idx) {
    par(mar = c(0, 0, 0, 0))
    plot.new()
    text(0.97, 0.5, paste(strwrap(feat_labels[idx], width = 12), collapse = "\n"),
      adj = c(1, 0.5), cex = 1.95
    )
  }
  # xaxs/yaxs = "i" + rasterImage fills the plot region exactly (no axis
  # padding); the region is square, so the crop is not stretched. The value at
  # line 0.3 sits just above its own image, while the large top margin (4.2)
  # keeps the separating gap above the value, tying it to the image below.
  crop_cell <- function(idx) {
    ex <- pick_examples(feat_order[idx])
    do_outline <- feat_order[idx] %in% cda_feats
    for (j in 1:3) {
      par(mar = c(0.2, 0.15, 4.2, 0.15))
      img <- load_img(file.path(data_dir, ex$score[j], paste0(ex$image_id[j], ".tif")))
      if (do_outline) img <- outline(img, segment_cda(img)$mask)
      plot(c(0, 1), c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "", xaxs = "i", yaxs = "i")
      rasterImage(as.raster(img), 0, 0, 1, 1)
      mtext(format(signif(ex$value[j], 3), scientific = FALSE, drop0trailing = TRUE, trim = TRUE),
        side = 3, line = 0.3, cex = 1.5
      )
    }
  }
  # Header row: blank | Low Med High | blank | Low Med High
  blank()
  for (cc in 1:3) header(cc)
  blank()
  for (cc in 1:3) header(cc)
  # Feature rows: left group (1..half) beside right group (half+1..2*half)
  for (r in seq_len(half)) {
    lab_cell(r)
    crop_cell(r)
    lab_cell(r + half)
    crop_cell(r + half)
  }
}
png(plot_output_path("10_segmentation_and_feature_extraction", "thesis_plots", "seg_feature_exemplars.png"),
  width = 2100, height = 2200, res = 150
)
draw_feat_examples()
invisible(dev.off())
draw_feat_examples()
```

### 5.2 How each feature relates to score, and to the others

Two summaries of the twelve features, both on the full (unbalanced) dataset and
using Spearman (rank) correlation so a feature counts as informative if it is
monotonically related to score, not necessarily linearly.

First, each feature's correlation with severity score (panel B of the
features-by-score figure above). CDA greenness, the log green-to-red ratio and
whole-leaf greenness bimodality lead (ρ ≈ 0.83-0.84); Shannon entropy and
whole-leaf brightness sit mid-pack (ρ ≈ 0.4-0.45); the CDA-leaf entropy
difference is a moderate, positive signal (ρ ≈ 0.3, and a clean monotone trend in
the class medians — the segmented CDA is more texturally complex, i.e. has higher
Shannon entropy, than the surrounding healthy leaf, and increasingly so with
severity); and within-CDA greenness variation carries almost no marginal signal
(ρ ≈ 0.02).

Panel C reframes the same features as adjacent-class separability: the mean, over
the six neighbouring score pairs (0-1, 1-2, ... 5-6), of the AUC — the probability
that a CDA from the higher score has the larger feature value. Adjacent pairs are
where raters actually disagree (off-by-one), so this isolates the separability
that matters, and the picture is stark: even the strongest features (CDA greenness,
the log green-to-red ratio) reach only AUC ≈ 0.68, the mid-pack features sit at
AUC ≈ 0.56-0.58, and within-CDA greenness variation is at chance (AUC ≈ 0.49). No
single hand-crafted feature comes close to cleanly separating neighbouring severity
scores, consistent with the heavy boxplot overlap in panel A and the off-by-one
disagreement seen among human raters.

The CDA-leaf entropy difference (segmented CDA entropy minus healthy-leaf entropy,
so it parallels the CDA-leaf greenness contrast) replaces the CDA area fraction
that earlier versions of this feature set used. Area was effectively an Otsu
artefact: Otsu splits every leaf into roughly two greenness classes whether or not
a real lesion is present, so the segmented area sat near half the leaf at every
score (ρ ≈ -0.1 with score) and contributed only through interactions. The entropy
difference is both monotonically related to score and statistically independent of
the other eleven features (see the matrix below), and swapping it in leaves
accuracy unchanged (§6.5). Note that some features with a high *marginal* correlation here,
such as within-CDA brightness variation (ρ ≈ 0.7), are conditionally redundant —
see the matrix below and the near-zero SHAP values in §6.8.

```{r fig.width=9, fig.height=4.5}
#| fig-width: 9
#| fig-height: 4.5
# 6.2  Correlation of each feature with severity score (Spearman, full dataset).
# Shown standalone here and reused as panel B of the features-by-score figure.
score_cor <- data.frame(
  feature = feat_labels,
  rho = vapply(feat_order, function(f) cor(feat6[[f]], feat6$score, method = "spearman"), numeric(1))
)
p_scorecor <- ggplot(score_cor, aes(x = reorder(feature, -rho), y = rho, fill = rho > 0)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.2f", rho), vjust = ifelse(rho > 0, -0.4, 1.3)), size = 3) +
  scale_fill_manual(values = c(`TRUE` = "#B2182B", `FALSE` = "#2166AC"), guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0.12, 0.12))) +
  labs(x = NULL, y = "Spearman correlation with score") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 40, hjust = 1, size = 10),
    axis.text.y = element_text(size = 11), axis.title = element_text(size = 12)
  )
p_scorecor
```

```{r}
#| output: false
# 6.2b  Per-feature class separability, and assembly of the thesis
# features-by-score figure.
#
# Separability is summarised as the mean adjacent-class AUC: for each pair of
# neighbouring scores (0-1, 1-2, ... 5-6), the probability that a CDA from the
# higher score has the larger feature value (ties count half), averaged over the
# six pairs. Adjacent pairs are the ones raters actually confuse (off-by-one
# disagreement), so this targets the separability that matters rather than being
# flattered by easily-separated far-apart scores. The general principle of
# averaging pairwise AUC over multiple classes follows Hand & Till (2001); 0.5
# marks no separability and 1 fully separated neighbours.
auc_pair <- function(hi, lo) {
  r <- rank(c(hi, lo))
  (sum(r[seq_along(hi)]) - length(hi) * (length(hi) + 1) / 2) /
    (length(hi) * length(lo))
}
adjacent_auc <- function(x, score) {
  ok <- is.finite(x)
  x <- x[ok]
  score <- score[ok]
  cls <- sort(unique(score))
  mean(vapply(seq_len(length(cls) - 1), function(k) {
    auc_pair(x[score == cls[k + 1]], x[score == cls[k]])
  }, numeric(1)))
}
score_auc <- data.frame(
  feature = feat_labels,
  auc = vapply(feat_order, function(f) adjacent_auc(feat6[[f]], feat6$score), numeric(1))
)
# Panel C shares panel B's feature order (descending Spearman rho) so the two bar
# rows line up under one shared set of x-axis labels.
feat_by_rho <- as.character(score_cor$feature[order(-score_cor$rho)])
score_auc$feature <- factor(score_auc$feature, levels = feat_by_rho)
p_auc <- ggplot(score_auc, aes(x = feature, y = auc)) +
  geom_col(fill = "#35978F") +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey40") +
  geom_text(aes(label = sprintf("%.2f", auc)), vjust = -0.4, size = 3) +
  scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Mean adjacent-class AUC (separability)") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 40, hjust = 1, size = 10),
    axis.text.y = element_text(size = 11), axis.title = element_text(size = 12)
  )
# Combined thesis figure: boxplots (panel A) over the Spearman correlations
# (panel B, x labels dropped since panel C carries them) over the separability
# AUC (panel C). Saved to thesis_plots.
p_scorecor_stk <- p_scorecor + theme(axis.text.x = element_blank())
p_features_abc <- (p_features / p_scorecor_stk / p_auc) +
  plot_layout(heights = c(3.4, 1, 1.25)) +
  plot_annotation(tag_levels = "A") &
  # size 30 so the tags render at ~16 pt on the thesis page (the 11 in figure is
  # scaled to ~149 mm there, a factor of ~0.53); Arial Bold matches the template.
  theme(plot.tag = element_text(size = 30, face = "bold", family = "Arial"))
save_plot(p_features_abc, "features_by_score.png", width = 11, height = 16, dir = "thesis_plots")
```

Second, the feature-by-feature correlation matrix, with its rows and columns
ordered by hierarchical clustering (average linkage on 1 − ρ) and the resulting
dendrogram drawn above, so that similar features sit next to each other. Several
features are near duplicates and cluster into tight, low branches: the CDA-leaf
greenness contrast and whole-leaf greenness spread are almost the same
measurement (ρ = 0.97); CDA brightness, whole-leaf mean green and whole-leaf
brightness form a brightness block (ρ = 0.87-0.92); and CDA greenness clusters
with the log green-to-red ratio (ρ = 0.92). The two least-redundant features — the
CDA-leaf entropy difference and within-CDA greenness variation — join the tree
last, at the greatest height, confirming they carry information the colour and
brightness features do not. The redundancy does not hurt the tree/ordinal models
but does mean the SHAP and Gini rankings (§6.7-6.8) split credit across
interchangeable features and should be read as blocks rather than as a strict
per-feature ordering.

```{r fig.width=9.5, fig.height=10}
#| fig-width: 9.5
#| fig-height: 10
# 6.3  Feature-feature correlation matrix with a similarity dendrogram. Spearman,
# all-vs-all; rows/columns are ordered by hierarchical clustering (average linkage
# on 1 - rho) so redundant features group together, and the dendrogram above shows
# the redundancy structure.
cormat <- cor(as.matrix(feat6[, feat_order]), method = "spearman")
dimnames(cormat) <- list(feat_labels, feat_labels)
hc <- hclust(as.dist(1 - cormat), method = "average")
ord_labels <- feat_labels[hc$order]

# Dendrogram line segments straight from the hclust object (ggdendro is not
# installed): each merge draws two verticals (up from each child to the merge
# height) and one horizontal joining them.
dendro_segments <- function(hc) {
  n <- length(hc$order)
  xpos <- match(seq_len(n), hc$order) # leaf i sits at this x position
  cx <- cy <- numeric(nrow(hc$merge))
  node <- function(i) if (i < 0) list(x = xpos[-i], y = 0) else list(x = cx[i], y = cy[i])
  segs <- vector("list", nrow(hc$merge))
  for (k in seq_len(nrow(hc$merge))) {
    a <- node(hc$merge[k, 1])
    b <- node(hc$merge[k, 2])
    h <- hc$height[k]
    segs[[k]] <- data.frame(
      x = c(a$x, b$x, a$x), xend = c(a$x, b$x, b$x),
      y = c(a$y, b$y, h), yend = c(h, h, h)
    )
    cx[k] <- (a$x + b$x) / 2
    cy[k] <- h
  }
  do.call(rbind, segs)
}
dseg <- dendro_segments(hc)

cor_long <- as.data.frame(as.table(cormat))
names(cor_long) <- c("x", "y", "rho")
cor_long$x <- factor(cor_long$x, levels = ord_labels)
cor_long$y <- factor(cor_long$y, levels = rev(ord_labels))

p_heat <- ggplot(cor_long, aes(x, y, fill = rho)) +
  geom_tile(colour = "grey92") +
  geom_text(aes(label = sprintf("%.2f", rho), colour = abs(rho) > 0.6), size = 3) +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B",
    midpoint = 0, limits = c(-1, 1), name = "Spearman rho"
  ) +
  scale_colour_manual(values = c(`FALSE` = "grey15", `TRUE` = "white"), guide = "none") +
  scale_x_discrete(expand = expansion(add = 0.5)) +
  scale_y_discrete(expand = expansion(add = 0.5)) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    panel.grid = element_blank(),
    legend.position = "bottom", legend.key.width = unit(2, "cm")
  )
# Dendrogram shares the heatmap's x range (leaves at 1..n, expand add 0.5) so
# patchwork aligns the two panels column-for-column.
p_dendro <- ggplot(dseg) +
  geom_segment(aes(x = x, xend = xend, y = y, yend = yend), linewidth = 0.4) +
  scale_x_continuous(expand = expansion(add = 0.5)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = NULL, y = NULL) +
  theme_void()

p_cormat <- p_dendro / p_heat + plot_layout(heights = c(1, 6))
p_cormat
save_plot(p_cormat, "feature_correlation_matrix.png", width = 9.5, height = 10, dir = "thesis_plots")
```

---

## 6. Predicting score from the hand-crafted features

Can these hand-crafted features predict a CDA's score? All modelling here uses a
**class-balanced** dataset — every score downsampled to the smallest class — so
the abundant score-0 class cannot dominate (the same balancing used in
`07_visual_complexity` and the AutoCDAScorer ensemble analysis). Because the
classes are equal, a majority-class guess scores only 1/7 and exact accuracy
equals balanced accuracy. Accuracy is reported as **exact** (predicted = true),
**near-miss** (within ±1, as in `07_visual_complexity`), and **balanced** (mean
of the per-class exact accuracies).

**The full pipeline, end to end**, for one cropped CDA image:

1. **Load** the raw RGB crop (no resizing — the features are scale-free).
2. **Leaf mask** — keep pixels with mean brightness ≥ 0.1 (§1.3).
3. **Segment the CDA** — Otsu threshold on the blurred leaf greenness (G − R),
   morphological cleanup, then keep the largest connected component (§3).
4. **Summarise** the crop into 12 scalar features (§6.1).
5. **Predict** the 0–6 score from those features (random forest, k-NN or ordinal
   regression). No per-image threshold tuning anywhere.

**The twelve features** — all per-image summaries of the raw crop, so they can be
read directly off the importance (§6.7) and SHAP (§6.8) plots. "Greenness" means
G − R (green minus red channel); the "CDA" is the segmented cell-death region
(the largest connected greenness component after Otsu thresholding and
morphological cleanup, §3); the "leaf" is all non-black pixels (§1.3); the
"healthy leaf" is the leaf minus the CDA.

_Segmentation features (this analysis, §3 & §4.1):_

| Variable | Feature | What it measures | With severity |
|---|---|---|---|
| `ent_gap` | CDA-leaf entropy difference | Shannon entropy (bits) of the segmented CDA minus that of the healthy leaf | rises |
| `cda_green` | CDA greenness | mean greenness (G − R) inside the segmented CDA | rises |
| `green_gap` | CDA–leaf contrast | CDA greenness minus healthy-leaf (leaf minus CDA) greenness | rises |
| `sep` | greenness bimodality | Otsu between-class ÷ total variance of leaf greenness (0–1): how cleanly the leaf splits into CDA + healthy | rises (also detects presence) |
| `cda_bright` | CDA brightness | mean brightness (mean of R,G,B, 0–255) inside the CDA | rises |
| `leaf_green_sd` | whole-leaf greenness spread | SD of greenness over all leaf pixels (0–255) | rises |
| `cda_green_sd` | within-CDA greenness spread | SD of greenness among CDA pixels: how mottled the lesion's colour is | ~flat |
| `cda_bright_sd` | within-CDA brightness spread | SD of brightness among CDA pixels | ~flat |

_Colour / brightness / entropy features (from `09_image_entropy`, §4 & §5):_

| Variable | Feature | What it measures | With severity |
|---|---|---|---|
| `mean_green` | mean green | mean green-channel intensity over leaf pixels (0–255) | rises |
| `log_gr` | log green/red ratio | log of the green-pixel to red-pixel count ratio over leaf pixels | rises |
| `leaf_bright` | leaf brightness | mean brightness over non-black leaf pixels (0–255) | rises |
| `entropy_raw` | Shannon entropy | information content of the raw crop (bits) | weakly rises |

### 6.1 Assemble features, and a leakage-safe cross-validation

Each cropped CDA comes from a source leaf image (~29 crops per leaf). Crops from
the same leaf share lighting, plant and tissue, so they are correlated; splitting
them across training and held-out folds would leak information and inflate
accuracy. All cross-validation here is therefore **grouped by source leaf** —
whole leaves are assigned to folds, so no leaf ever appears in both the training
and held-out folds.
(Ungrouped image-level cross validation gives ~0.58 exact; grouping removes the
leak and gives the honest ~0.55.) A shuffled-label check (§6.1) confirms the
pipeline has no other leakage.

```{r}
# 7.1.1  Features and class balancing
# All region features (1-11) are computed in `seg`; only Shannon entropy (12) is
# taken from the entropy analysis.
main_feat <- readRDS(here(
  "analyses", "09_image_entropy", "entropy_cache",
  "cda_metrics.rds"
)) %>%
  filter(size == "Raw") %>%
  transmute(image_id, entropy_raw = entropy)

seg_features <- c(
  "ent_gap", "green_gap", "cda_green", "sep",
  "cda_green_sd", "cda_bright", "cda_bright_sd", "leaf_green_sd",
  "mean_green", "log_gr", "leaf_bright"
)
other_features <- c("entropy_raw")
all_features <- c(seg_features, other_features)

# frac_g_gt_r is no longer a model feature (its Otsu-artefact behaviour is
# discussed in section 5.2); it is kept in the frame only for the single-feature
# "Threshold: CDA area" baseline in section 6.5.
feat_all <- seg %>%
  select(image_id, score, frac_g_gt_r, all_of(seg_features)) %>%
  inner_join(main_feat, by = "image_id")

set.seed(42)
min_n <- min(table(feat_all$score))
bal_idx <- unlist(lapply(0:6, function(s) {
  idx <- which(feat_all$score == s)
  if (length(idx) > min_n) sample(idx, min_n) else idx
}))
feat <- feat_all[bal_idx, ]
X <- as.matrix(feat[, all_features])
y <- feat$score
cat(sprintf(
  "Balanced set: %d images per class, %d total, %d features\n",
  min_n, nrow(feat), length(all_features)
))
```

```{r}
# 7.1.2  Source-grouped 5-fold assignment (a leaf's crops never split across folds)
# The crop filename is <SourceBasename>_<Row>_<Col>_<Pos>, so stripping the
# trailing _Row_Col_Pos recovers the source leaf image.
feat$source <- sub("_[0-9]+_[0-9]+_[0-9]+$", "", feat$image_id)
K <- 5
set.seed(42)
usrc <- unique(feat$source)
src_fold <- setNames(sample(rep(1:K, length.out = length(usrc))), usrc)
fold <- unname(src_fold[feat$source])
cat(sprintf(
  "%d source leaves assigned across %d folds; fold sizes: %s\n",
  length(usrc), K, paste(as.integer(table(fold)), collapse = ", ")
))

metric_row <- function(pred, name, tuning = "") {
  per <- vapply(0:6, function(s) mean(pred[y == s] == s), numeric(1))
  data.frame(
    model = name, tuning = tuning,
    exact = mean(pred == y), near_miss = mean(abs(pred - y) <= 1),
    balanced = mean(per)
  )
}
# Per-fold accuracy, so the cross-validation is fully transparent (reported
# alongside the mean, not just pooled).
per_fold <- function(pred, name) {
  do.call(rbind, lapply(1:K, function(f) {
    m <- fold == f
    data.frame(
      model = name, fold = as.character(f), n = sum(m),
      exact = mean(pred[m] == y[m]),
      near_miss = mean(abs(pred[m] - y[m]) <= 1)
    )
  }))
}
```

```{r}
# 7.1.3  Random-forest regressor under the grouped cross validation, plus a leakage sanity check.
rf_reg_cv <- function(ntree, target = y) {
  pred <- numeric(length(target))
  for (f in 1:K) {
    tr <- fold != f
    te <- fold == f
    m <- randomForest::randomForest(x = X[tr, , drop = FALSE], y = target[tr], ntree = ntree)
    pred[te] <- predict(m, X[te, , drop = FALSE])
  }
  pmin(pmax(round(pred), 0), 6)
}
# Shuffle the labels: a correctly-built cross validation with no leakage MUST collapse to chance.
set.seed(7)
y_shuffled <- sample(y)
shuffle_acc <- mean(rf_reg_cv(200, y_shuffled) == y_shuffled)
cat(sprintf(
  "Leakage check - shuffled-label cross validation exact accuracy = %.3f (chance = %.3f)\n",
  shuffle_acc, 1 / 7
))
```

### 6.2 Choosing k for k-NN

k is swept over a range and chosen by grouped cross-validated exact accuracy (the
supervised analogue of the k-selection in `07_visual_complexity`). Held-out
accuracy is essentially flat beyond k ≈ 5, so the precise value barely matters;
it is taken as the cross validation maximum for definiteness. (Only held-out accuracy is shown
— training-set accuracy is meaningless for k-NN, since at k = 1 each point is its
own neighbour and scores 100% by construction.)

```{r}
# 7.2.1
knn_cv <- function(ks) {
  P <- matrix(0L, nrow(X), length(ks), dimnames = list(NULL, paste0("k", ks)))
  for (f in 1:K) {
    tr <- fold != f
    te <- fold == f
    mu <- colMeans(X[tr, ])
    sdv <- apply(X[tr, ], 2, sd)
    sdv[sdv == 0] <- 1
    Xtr <- scale(X[tr, , drop = FALSE], mu, sdv)
    Xte <- scale(X[te, , drop = FALSE], mu, sdv)
    d <- outer(rowSums(Xte^2), rowSums(Xtr^2), "+") - 2 * Xte %*% t(Xtr)
    ytr <- y[tr]
    ord <- t(apply(d, 1, order))
    for (j in seq_along(ks)) {
      P[te, j] <- apply(ord[, 1:ks[j], drop = FALSE], 1, function(ix) {
        tb <- table(ytr[ix])
        as.integer(names(tb)[which.max(tb)])
      })
    }
  }
  P
}
kseq <- seq(1, 51, by = 2)
knn_preds <- knn_cv(kseq)
knn_curve <- data.frame(
  k = kseq,
  exact = apply(knn_preds, 2, function(p) mean(p == y)),
  near_miss = apply(knn_preds, 2, function(p) mean(abs(p - y) <= 1))
)
best_k <- kseq[which.max(knn_curve$exact)]
cat(sprintf(
  "Best k = %d  (held-out exact = %.3f, near-miss = %.3f)\n",
  best_k, max(knn_curve$exact), knn_curve$near_miss[knn_curve$k == best_k]
))
```

```{r fig.width=8, fig.height=4.5}
#| fig-width: 8
#| fig-height: 4.5
# 7.2.2
p_ksweep <- ggplot(knn_curve, aes(k, exact)) +
  geom_line(colour = "#2166AC") +
  geom_point(colour = "#2166AC", size = 1.4) +
  geom_vline(xintercept = best_k, linetype = "dashed", colour = "grey40") +
  coord_cartesian(ylim = c(min(knn_curve$exact) - 0.02, max(knn_curve$exact) + 0.02)) +
  labs(
    x = "k (number of neighbours)", y = "Held-out exact accuracy (grouped 5-fold cross validation)",
    title = sprintf("k-NN: held-out accuracy is flat beyond k ~ 5; chosen k = %d", best_k)
  ) +
  theme_minimal() +
  theme(axis.text = element_text(size = 12), axis.title = element_text(size = 13))
p_ksweep
save_plot(p_ksweep, "seg_knn_k_sweep.png", width = 8, height = 4.5)
```

### 6.3 Random forest: regularising to control overfitting

A fully-grown random forest memorises the training set — its training accuracy is
~0.91 against ~0.55 held-out, a ~36-point gap — so by the rule of discarding any
model whose train–cross-validation gap exceeds 15%, it must be regularised. The
forest is fixed at 200 trees (accuracy plateaus well before that, and more trees
never hurt a random forest) and the tree size is capped with `maxnodes`,
increasing it until the train–cross-validation gap reaches 15%; the largest
held-out accuracy within that bound is chosen. Training accuracy is fit and scored
on all the data (resubstitution); held-out is the grouped 5-fold cross validation.

```{r}
# 7.3.1
NTREE <- 200
rf_fit_cv <- function(maxnodes) { # grouped cross validation (held-out) predictions
  pred <- numeric(length(y))
  for (f in 1:K) {
    tr <- fold != f
    te <- fold == f
    m <- randomForest::randomForest(
      x = X[tr, , drop = FALSE], y = y[tr],
      ntree = NTREE, maxnodes = maxnodes
    )
    pred[te] <- predict(m, X[te, , drop = FALSE])
  }
  pmin(pmax(round(pred), 0), 6)
}
rf_fit_train <- function(maxnodes) { # resubstitution (training) predictions
  set.seed(1)
  m <- randomForest::randomForest(x = X, y = y, ntree = NTREE, maxnodes = maxnodes)
  pmin(pmax(round(predict(m, X)), 0), 6)
}
maxnodes_grid <- c(4, 8, 16, 32, 64, 128, 256)
set.seed(42)
reg_curve <- data.frame(
  maxnodes = maxnodes_grid,
  train = sapply(maxnodes_grid, function(mn) mean(rf_fit_train(mn) == y)),
  cross_validation = sapply(maxnodes_grid, function(mn) mean(rf_fit_cv(mn) == y))
)
reg_curve$gap <- reg_curve$train - reg_curve$cross_validation
best_maxnodes <- max(reg_curve$maxnodes[reg_curve$gap <= 0.15])
cat(sprintf(
  "Chosen maxnodes = %d: train = %.3f, cross validation = %.3f, gap = %.3f (<= 0.15)\n",
  best_maxnodes,
  reg_curve$train[reg_curve$maxnodes == best_maxnodes],
  reg_curve$cross_validation[reg_curve$maxnodes == best_maxnodes],
  reg_curve$gap[reg_curve$maxnodes == best_maxnodes]
))
rf_best <- rf_fit_cv(best_maxnodes) # held-out predictions of the chosen model
```

```{r fig.width=8, fig.height=4.5}
#| fig-width: 8
#| fig-height: 4.5
# 7.3.2
p_reg <- reg_curve %>%
  pivot_longer(c(train, cross_validation), names_to = "set", values_to = "accuracy") %>%
  ggplot(aes(maxnodes, accuracy, colour = set)) +
  geom_line() +
  geom_point(size = 1.6) +
  geom_vline(xintercept = best_maxnodes, linetype = "dashed", colour = "grey40") +
  scale_colour_manual(
    values = c(train = "#B2182B", cross_validation = "#2166AC"),
    labels = c(cross_validation = "held out (grouped 5-fold cross validation)", train = "training"),
    name = NULL
  ) +
  scale_x_log10(breaks = maxnodes_grid) +
  labs(
    x = "max leaf nodes per tree (log scale)", y = "Exact accuracy",
    title = sprintf("RF regularisation: train-cross-validation gap kept <= 15%%; maxnodes = %d", best_maxnodes)
  ) +
  theme_minimal() +
  theme(axis.text = element_text(size = 12), axis.title = element_text(size = 13))
p_reg
save_plot(p_reg, "seg_rf_regularisation.png", width = 8, height = 4.5)
```

### 6.4 Ordinal regression

Proportional-odds ordinal logistic regression models the ordered 0–6 scale directly:
a single linear predictor is fit to the 12 features, and six ordered cut-points split
that predictor into the seven score classes, so the model respects the ordering of
the scale rather than treating the classes as unrelated labels. It has no
hyperparameter to tune, so unlike k-NN (§6.2) and the random forest (§6.3) there is no
sweep — just the fit below, scored under the same source-grouped cross validation.
With only twelve coefficients plus the cut-points it has little capacity to memorise
the training set, so its training and held-out accuracy are almost identical.

```{r}
# 7.4.0  Ordinal logistic regression (proportional odds, MASS::polr), scored under
# the same source-grouped cross validation. Features are standardised with
# training-fold statistics inside each fold, so nothing leaks into the scaling.
ord_cv <- function() {
  pred <- integer(length(y))
  for (f in 1:K) {
    tr <- fold != f
    te <- fold == f
    mu <- colMeans(X[tr, ])
    sdv <- apply(X[tr, ], 2, sd)
    sdv[sdv == 0] <- 1
    dtr <- data.frame(scale(X[tr, , drop = FALSE], mu, sdv),
      y = factor(y[tr], levels = 0:6, ordered = TRUE)
    )
    dte <- data.frame(scale(X[te, , drop = FALSE], mu, sdv))
    fit <- suppressWarnings(MASS::polr(y ~ ., data = dtr, Hess = TRUE))
    pred[te] <- as.integer(as.character(predict(fit, dte, type = "class")))
  }
  pred
}
ord_train <- function() {
  mu <- colMeans(X)
  sdv <- apply(X, 2, sd)
  sdv[sdv == 0] <- 1
  d <- data.frame(scale(X, mu, sdv), y = factor(y, levels = 0:6, ordered = TRUE))
  fit <- suppressWarnings(MASS::polr(y ~ ., data = d, Hess = TRUE))
  as.integer(as.character(predict(fit, data.frame(scale(X, mu, sdv)), type = "class")))
}
ord_best <- ord_cv()
ord_tr <- ord_train()
ord_fmt <- paste0(
  "Ordinal regression: held-out (grouped cross validation) exact = %.3f, near-miss = %.3f; ",
  "training = %.3f, gap = %.3f\n"
)
cat(sprintf(
  ord_fmt,
  mean(ord_best == y), mean(abs(ord_best - y) <= 1),
  mean(ord_tr == y), mean(ord_tr == y) - mean(ord_best == y)
))
```

### 6.5 Model comparison, with per-fold results

The tuned models are compared with a majority-class baseline and a single-feature
threshold rule, all under the same grouped 5-fold cross validation. Both
**training** (fit and scored on all the data) and **held-out** (grouped cross
validation) accuracy are reported for every model, with the gap between them; any
model whose gap exceeds 15% is flagged for discarding as overfit (only the
*fully-grown* forest of §6.3 fails — every tuned model here passes).

```{r}
# 7.4.1
# Held-out (grouped cross validation) predictors. The ordinal-regression
# predictors (ord_best, ord_tr) are defined in section 6.4 above.
thresh_cv <- function(x) {
  pred <- integer(length(y))
  for (f in 1:K) {
    tr <- fold != f
    te <- fold == f
    med <- tapply(x[tr], y[tr], median)
    sc <- as.integer(names(med))
    o <- order(sc)
    med <- med[o]
    sc <- sc[o]
    cuts <- cummax((head(med, -1) + tail(med, -1)) / 2)
    pred[te] <- sc[findInterval(x[te], cuts) + 1]
  }
  pred
}
# Training (resubstitution) predictors, for the train-cross-validation gap
thresh_train <- function(x) {
  med <- tapply(x, y, median)
  sc <- as.integer(names(med))
  o <- order(sc)
  med <- med[o]
  sc <- sc[o]
  cuts <- cummax((head(med, -1) + tail(med, -1)) / 2)
  sc[findInterval(x, cuts) + 1]
}
knn_train_at <- function(k) { # self-inclusive; meaningful for k > 1
  mu <- colMeans(X)
  sdv <- apply(X, 2, sd)
  sdv[sdv == 0] <- 1
  Xs <- scale(X, mu, sdv)
  d <- outer(rowSums(Xs^2), rowSums(Xs^2), "+") - 2 * Xs %*% t(Xs)
  ord <- t(apply(d, 1, order))
  apply(ord[, 1:k, drop = FALSE], 1, function(ix) {
    tb <- table(y[ix])
    as.integer(names(tb)[which.max(tb)])
  })
}

knn_best <- knn_preds[, which(kseq == best_k)]
maj <- rep(as.integer(names(which.max(table(y)))), length(y))

cmp_row <- function(train_pred, test_pred, name, tuning) {
  data.frame(
    model = name, tuning = tuning,
    train = mean(train_pred == y), cross_validation = mean(test_pred == y),
    gap = mean(train_pred == y) - mean(test_pred == y),
    near_miss = mean(abs(test_pred - y) <= 1)
  )
}
results <- rbind(
  cmp_row(maj, maj, "Majority class", ""),
  cmp_row(thresh_train(feat$frac_g_gt_r), thresh_cv(feat$frac_g_gt_r), "Threshold: CDA area", "1 feature"),
  cmp_row(ord_tr, ord_best, "Ordinal regression", "12 features"),
  cmp_row(knn_train_at(best_k), knn_best, "k-NN", sprintf("k = %d", best_k)),
  cmp_row(rf_fit_train(best_maxnodes), rf_best, "Random forest", sprintf("%d nodes", best_maxnodes))
)

results %>%
  mutate(across(c(train, cross_validation, gap, near_miss), ~ round(., 3)),
    overfit = ifelse(gap > 0.15, "DISCARD", "ok")
  ) %>%
  arrange(desc(cross_validation)) %>%
  as.data.frame()
```

```{r}
# 7.4.2  Per-fold breakdown of the best model (chosen by held-out accuracy).
cv_preds <- list("Random forest" = rf_best, "k-NN" = knn_best, "Ordinal regression" = ord_best)
best_cv_name <- names(cv_preds)[which.max(vapply(cv_preds, function(p) mean(p == y), numeric(1)))]
pf <- per_fold(cv_preds[[best_cv_name]], best_cv_name)
print(transform(pf, exact = round(exact, 3), near_miss = round(near_miss, 3)))
cat(sprintf(
  "MEAN across folds: exact = %.3f (fold range %.3f-%.3f); near-miss = %.3f\n",
  mean(pf$exact), min(pf$exact), max(pf$exact), mean(pf$near_miss)
))
```

For reference (not a like-for-like comparison — different data split and class
distribution), the project's deep-learning models reach ~70% training and 60–67%
held-out test accuracy: a small train–test gap, and test accuracy broadly
comparable to these hand-crafted features
(~`r sprintf("%.0f%%", 100 * max(results$cross_validation))`).

### 6.6 Confusion matrix of the best model

```{r fig.width=7, fig.height=6}
#| fig-width: 7
#| fig-height: 6
# 7.5
model_preds <- list(
  "Random forest" = rf_best, "k-NN" = knn_best,
  "Ordinal regression" = ord_best
)
best_name <- names(model_preds)[which.max(vapply(
  model_preds,
  function(p) mean(p == y), numeric(1)
))]
best_pred <- model_preds[[best_name]]

cm <- as.data.frame(table(true = factor(y, 0:6), pred = factor(best_pred, 0:6)))
p_cm <- ggplot(cm, aes(x = pred, y = true, fill = Freq)) +
  geom_tile(colour = "grey90") +
  geom_text(aes(label = Freq), size = 3.5) +
  scale_fill_gradient(low = "white", high = "#B2182B", guide = "none") +
  scale_y_discrete(limits = rev) +
  labs(
    x = "Predicted score", y = "True score",
    title = sprintf(
      "%s (grouped cross validation): exact %.1f%%, near-miss %.1f%%",
      best_name, 100 * mean(best_pred == y),
      100 * mean(abs(best_pred - y) <= 1)
    )
  ) +
  coord_equal() +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 12), axis.title = element_text(size = 14),
    plot.title = element_text(size = 12), panel.grid = element_blank()
  )
p_cm
save_plot(p_cm, "seg_confusion_matrix.png", width = 7, height = 6)
```

### 6.7 Gini importance

A random forest classifier on the balanced data ranks the features by mean
decrease in Gini impurity.

```{r fig.width=7, fig.height=5}
#| fig-width: 7
#| fig-height: 5
# 7.6
set.seed(1)
rf_clf <- randomForest::randomForest(
  x = X, y = factor(y, levels = 0:6),
  ntree = NTREE, maxnodes = best_maxnodes,
  importance = TRUE
)
imp_df <- data.frame(
  feature = rownames(randomForest::importance(rf_clf)),
  importance = randomForest::importance(rf_clf)[, "MeanDecreaseGini"]
)
p_imp <- ggplot(imp_df, aes(x = reorder(feature, importance), y = importance)) +
  geom_col(fill = "#2166AC") +
  coord_flip() +
  labs(x = NULL, y = "Random-forest importance (mean decrease in Gini)") +
  theme_minimal() +
  theme(axis.text = element_text(size = 12), axis.title = element_text(size = 13))
p_imp
save_plot(p_imp, "seg_feature_importance.png", width = 7, height = 5)
```

### 6.8 SHAP analysis

Gini importance is global and directionless. SHAP (SHapley Additive
exPlanations) attributes each prediction to its features, showing both magnitude
and direction — replicating the SHAP series from the AutoCDAScorer ensemble
analysis, here in R via `kernelshap` + `shapviz`. SHAP is computed on the random
forest **regressor** (single output) with an interventional, class-balanced
background set.

```{r}
#| output: false
# 7.7.1
library(kernelshap)
library(shapviz)

set.seed(1)
rf_reg <- randomForest::randomForest(x = X, y = y, ntree = NTREE, maxnodes = best_maxnodes)

# Class-balanced background (~14/class) and a class-balanced explanation set.
Xdf <- as.data.frame(X)
bg_idx <- unlist(lapply(0:6, function(s) sample(which(y == s), 14)))
ex_idx <- unlist(lapply(0:6, function(s) sample(which(y == s), 20)))
shp <- shapviz(suppressWarnings(
  permshap(rf_reg, X = Xdf[ex_idx, ], bg_X = Xdf[bg_idx, ])
))
y_ex <- y[ex_idx]
```

```{r fig.width=8, fig.height=6}
#| fig-width: 8
#| fig-height: 6
# 7.7.2 Beeswarm: global importance + direction of each feature's effect.
p_bee <- sv_importance(shp, kind = "beeswarm", max_display = length(all_features)) +
  theme_minimal() + theme(axis.text = element_text(size = 11))
p_bee
save_plot(p_bee, "seg_shap_beeswarm.png", width = 8, height = 6)
```

```{r fig.width=11, fig.height=4}
#| fig-width: 11
#| fig-height: 4
# 7.7.3 Dependence plots for the three most important features (by mean |SHAP|).
mean_abs <- sort(colMeans(abs(get_shap_values(shp))), decreasing = TRUE)
top3 <- names(mean_abs)[1:3]
p_dep <- patchwork::wrap_plots(lapply(top3, function(v) {
  sv_dependence(shp, v = v, color_var = NULL) + theme_minimal()
}), nrow = 1)
p_dep
save_plot(p_dep, "seg_shap_dependence.png", width = 11, height = 4)
```

```{r}
# 7.7.4 Does the feature ranking reshuffle between low (0-1) and high (5-6)
# severity? (mean |SHAP| per feature, matching the AutoCDAScorer MSHAP table.)
S <- abs(get_shap_values(shp))
mshap <- data.frame(
  feature = colnames(S),
  low_severity = colMeans(S[y_ex %in% 0:1, , drop = FALSE]),
  high_severity = colMeans(S[y_ex %in% 5:6, , drop = FALSE])
)
mshap$rank_low <- rank(-mshap$low_severity)
mshap$rank_high <- rank(-mshap$high_severity)
mshap %>%
  mutate(across(c(low_severity, high_severity), ~ round(., 3))) %>%
  arrange(desc(high_severity)) %>%
  as.data.frame()
```

```{r fig.width=8, fig.height=5}
#| fig-width: 8
#| fig-height: 5
# 7.7.5 Waterfall: how the features add up to the prediction for one CDA
# (a mid-severity example, where the models struggle most).
wf_row <- which(y_ex == 3)[1]
p_wf <- sv_waterfall(shp, row_id = wf_row) +
  labs(title = sprintf("SHAP waterfall for one score-%d CDA", y_ex[wf_row]))
p_wf
save_plot(p_wf, "seg_shap_waterfall.png", width = 8, height = 5)
```

---

## 7. PCA of the image content, with and without the background

`07_visual_complexity` runs a principal component analysis (PCA) on the CDA
images and colours the first two components by severity score. Here the same PCA
is repeated on images in which everything outside the segmented CDA is set to
black, to test whether removing the leaf background and imaging artefacts makes
the score ordering clearer. Both PCAs use the same class-balanced sample (the
smallest class's size per score), the same 64 × 64 pixel normalised inputs, and
the same severity-score colouring; the only difference is whether the background
is masked out.

```{r}
# 8.1  Build original and CDA-masked 64x64 inputs and run PCA (cached: only the
# small PC scores are stored, not the image matrices).
pca_cache <- here("analyses", "10_segmentation_and_feature_extraction", "segmentation_cache", "masked_pca.rds")
if (file.exists(pca_cache)) {
  pca_res <- readRDS(pca_cache)
} else {
  set.seed(42)
  n_per <- min(vapply(0:6, function(s) {
    length(list.files(file.path(data_dir, s), pattern = "\\.tif$"))
  }, integer(1)))
  pca_files <- unlist(lapply(0:6, function(s) {
    sample(list.files(file.path(data_dir, s), pattern = "\\.tif$", full.names = TRUE), n_per)
  }))
  y_pca <- rep(0:6, each = n_per)
  Xo <- matrix(0, length(pca_files), 64 * 64 * 3)
  Xm <- Xo
  for (i in seq_along(pca_files)) {
    img <- load_img(pca_files[i])
    Xo[i, ] <- as.vector(EBImage::resize(img, 64, 64))
    mask <- segment_cda(img)$mask
    mm <- img
    for (k in 1:3) {
      ch <- mm[, , k]
      ch[!mask] <- 0
      mm[, , k] <- ch
    }
    Xm[i, ] <- as.vector(EBImage::resize(mm, 64, 64))
  }
  Xo <- pmin(pmax(Xo, 0), 1)
  Xm <- pmin(pmax(Xm, 0), 1)
  po <- prcomp(Xo, center = TRUE, rank. = 3)
  pm <- prcomp(Xm, center = TRUE, rank. = 3)
  eta_by_pc <- function(x) {
    sapply(1:2, function(j) {
      v <- x[, j]
      gr <- mean(v)
      sum(tapply(v, y_pca, function(z) length(z) * (mean(z) - gr)^2)) / sum((v - gr)^2)
    })
  }
  pca_res <- list(
    score = y_pca,
    orig = po$x[, 1:2], masked = pm$x[, 1:2],
    orig_var = summary(po)$importance[2, 1:2], masked_var = summary(pm)$importance[2, 1:2],
    orig_eta = eta_by_pc(po$x), masked_eta = eta_by_pc(pm$x)
  )
  saveRDS(pca_res, pca_cache)
}
cat(sprintf(
  "PC1/PC2 variance explained: original %.1f%%/%.1f%%, masked %.1f%%/%.1f%%\n",
  100 * pca_res$orig_var[1], 100 * pca_res$orig_var[2],
  100 * pca_res$masked_var[1], 100 * pca_res$masked_var[2]
))
cat(sprintf(
  "eta^2 (PC ~ score; higher = more ordered): PC1 %.3f -> %.3f, PC2 %.3f -> %.3f  (original -> masked)\n",
  pca_res$orig_eta[1], pca_res$masked_eta[1], pca_res$orig_eta[2], pca_res$masked_eta[2]
))
```

```{r fig.width=11, fig.height=5}
#| fig-width: 11
#| fig-height: 5
# 8.2
pca_panel <- function(pc, var, title) {
  data.frame(
    PC1 = pc[, 1], PC2 = pc[, 2],
    score = factor(pca_res$score, levels = 0:6)
  ) %>%
    ggplot(aes(PC1, PC2, colour = score)) +
    geom_point(alpha = 0.5, size = 0.8) +
    scale_colour_manual(name = "Severity Score", values = sev_pal) +
    guides(colour = guide_legend(override.aes = list(alpha = 1, size = 2))) +
    labs(
      title = title, x = sprintf("PC1 (%.1f%%)", 100 * var[1]),
      y = sprintf("PC2 (%.1f%%)", 100 * var[2])
    ) +
    theme_minimal() +
    theme(
      axis.text = element_text(size = 11), axis.title = element_text(size = 12),
      plot.title = element_text(size = 12), legend.title = element_text(size = 12)
    )
}
pca_left <- pca_panel(pca_res$orig, pca_res$orig_var, "Original (background included)")
pca_right <- pca_panel(pca_res$masked, pca_res$masked_var, "Masked to the segmented CDA")
p_pca <- (pca_left | pca_right) + plot_layout(guides = "collect")
p_pca
save_plot(p_pca, "seg_pca_masked_vs_original.png", width = 11, height = 5)
```

Masking the background concentrates the severity signal onto the first component:
its alignment with score (η²) increases, and the masked scatter shows a clearer
low-to-high gradient along PC1, whereas in the original the gradient runs
diagonally across PC1 and PC2. Adjacent scores still overlap substantially,
consistent with the rest of this analysis, but removing the background does make
the ordering somewhat cleaner.

---

## 8. Interpretation and Discussion

- **Greenness solves the universal-threshold problem.** Switching the
  segmentation channel from brightness to greenness (G − R) makes CDA and leaf
  distinct modes, so **Otsu sets the threshold automatically per image** with no
  manual tuning (§2–§3). This is the fix for the failure in
  `0009_Non_DL_Green_Channel_Analysis`, where brightness watershedding needed a
  different threshold for every image.
- **Presence detection works, but segment conditionally.** The bimodality
  (separability) of the leaf-greenness distribution classifies 0 vs non-0 well on
  its own (AUC ≈ 0.94, §4.2), so the distribution *does* detect whether a CDA is
  present — confirming this is the easy part of the problem. But if the spatial
  Otsu segmentation is run unconditionally it still outlines a spurious region on
  a median-0 leaf (§3), so segmentation should be gated on the non-0 decision,
  as hypothesised.
- **The features predict score to ±1 reliably, but the exact score only about
  half the time (§6).** All modelling is on the class-balanced data (436/class),
  so there is no score-0 free ride and exact = balanced accuracy, and all
  accuracy is measured with **source-grouped** 5-fold cross validation (a leaf's
  crops are never split across folds; an image-level split leaks ~2–3 points,
  §6.1). A shuffled-label check collapses to chance, confirming no other leakage.
  The best model is a linear **ordinal regression** at
  **`r sprintf("%.0f%%", 100 * results$cross_validation[results$model == "Ordinal regression"])` held-out exact**
  and
  **`r sprintf("%.0f%%", 100 * results$near_miss[results$model == "Ordinal regression"])` near-miss**,
  with a train–cross-validation gap of essentially zero; a regularised random forest
  (tree size capped so the train–cross-validation gap stays under 15%, §6.3) and a
  k-NN (k = `r best_k`, chosen by cross validation) trail a few points behind. A
  single-feature threshold rule reaches
  **`r sprintf("%.0f%%", 100 * results$cross_validation[results$model == "Threshold: CDA area"])`**.
  The confusion matrix is a tight band-diagonal (errors almost
  entirely between adjacent scores; score 1 mostly confused with 0 and 2), so the
  features locate a CDA to within ±1 of its score almost always but cannot pin the
  exact neighbouring severity. An unregularised forest reaches ~91% *training*
  accuracy but the same ~55% held-out — pure memorisation (a ~36-point gap), so it
  is discarded; regularising costs almost nothing in held-out accuracy. For
  reference (not a like-for-like comparison), the project's deep-learning models
  reach ~70% training and 60–67% held-out test — a small gap and higher test
  accuracy than these hand-crafted features.
- **Gini importance and SHAP agree on what matters (§6.7–6.8).** CDA greenness,
  the green/red ratio, greenness bimodality and CDA brightness drive the
  predictions. The SHAP beeswarm shows high values
  of these features push the prediction toward higher scores, consistently across
  images rather than as narrow high- or low-severity specialists; the dependence
  plots show mostly monotonic (area saturating) effects, and the within-CDA
  texture features (`cda_green_sd`, `cda_bright_sd`) sit near zero — how mottled a
  lesion is matters far less than how large, green and bright it is.
- **Low scores remain hard.** At scores 1–2 the CDA is barely greener than the
  leaf, so the modes overlap and the segmentation
  features overlap the neighbouring classes (§5, §6). This is the same
  faint-signal regime that limits every hand-crafted feature in `09_image_entropy`
  — mild cell death is genuinely close to healthy tissue in these images, which
  is why a learned model is needed to resolve the fine-grained scores.
