---
title: "Inter-Rater Agreement"
format:
  html:
    code-fold: true
---

Date: 15/5/2024 – 10/6/2024

## GOHREP

**Goal:**

Investigate changes to average pairwise agreement when each rater is removed. Investigate differences in agreement across source agroinfiltration images. Quantify the overall inter-rater reliability of the human rater panel using Krippendorff's alpha.

**Hypothesis:**

Excluding scores from raters with low pairwise agreement will increase the overall average pairwise agreement. Some images will be scored more consistently than others due to differences in clarity or familiarity. Krippendorff's alpha on raw scores will be high (>0.8), reflecting that raters broadly agree on severity ordering even when they differ by ±1.

**Rationale:**

Differences in lighting or leaf clarity could cause some images to be scored less consistently. Identifying which raters most reduce overall agreement informs future data collection decisions. Krippendorff's alpha is the standard metric for ordinal inter-rater reliability and handles missing data (not all raters rated all CDAs), giving a single overall reliability estimate that pairwise % agreement alone does not.

---

## 1. Setup

### 1.1 Load required packages

```{r}
#| output: false
# 1.1
library(readr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(forcats)
library(patchwork)
library(scales)
library(here)
library(irr)
```

### 1.2 Load data

```{r}
# 1.2
path_data <- here("analyses", "01_generate_raw_data", "04_median_and_centre_coords", "combined_cda_data_median.csv")

raw_data <- read_csv(path_data, show_col_types = FALSE)

score_data <- raw_data %>%
  select(Basename, Row, Col, Pos, Scorer1, Score1, Scorer2, Score2, Scorer3, Score3, Median_Score) %>%
  mutate(
    across(c(Scorer1, Scorer2, Scorer3), as.character),
    across(c(Score1, Score2, Score3), as.character)
  )

score_data_no_0 <- score_data %>%
  filter(Median_Score != 0)

n_cdas <- nrow(score_data)
```

### 1.3 Helper functions

```{r}
# 1.3.1
source(here("src", "data_helpers.R"))
source(here("src", "agreement_metrics.R"))
source(here("src", "plotting_helpers.R"))

save_plot <- make_save_plot("05_inter_rater_agreement")

theme_ira <- function() {
  theme_minimal() +
    theme(
      axis.text = element_text(size = 13),
      axis.title = element_text(size = 15),
      legend.title = element_text(size = 13),
      legend.text = element_text(size = 11),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_blank(),
      legend.key.size = unit(1.5, "lines")
    )
}
```

```{r}
# 1.3.2
score_data_long <- long_score_table(score_data)
score_data_long_no_0 <- long_score_table(score_data_no_0)
```

### 1.4 Build rater × CDA matrix (for Krippendorff's alpha)

```{r}
# 1.4
# kripp.alpha() needs a rater x item matrix rather than the long format used
# above, so scores are kept numeric here (missing scores, stored as "[]",
# become NA via as.numeric() rather than being filtered out as a string).
score_data_numeric <- raw_data %>%
  select(Basename, Row, Col, Pos, Scorer1, Score1, Scorer2, Score2, Scorer3, Score3, Median_Score) %>%
  mutate(across(c(Scorer1, Scorer2, Scorer3), as.character),
    across(c(Score1, Score2, Score3), as.numeric),
    Median_Score = as.numeric(Median_Score)
  )

scorer_matrix <- build_scorer_matrix(score_data_numeric)
cat("Rater matrix dimensions:", nrow(scorer_matrix), "raters ×", ncol(scorer_matrix), "CDAs\n")
cat("Non-NA cells:", sum(!is.na(scorer_matrix)), "/", prod(dim(scorer_matrix)), "\n")
```

---

## 2. Agreement by median score

```{r}
# 2
# One row per CDA. Missing scores are stored as "[]"; treat them as NA.
# Agreement is defined by how many of the three scorers match the median score:
#   3 -> all three agreed, 2 -> two of three agreed, 1 -> all three differed.
conf_cda <- raw_data %>%
  select(Basename, Row, Col, Pos, Score1, Score2, Score3, Median_Score) %>%
  mutate(across(c(Score1, Score2, Score3), ~ na_if(as.character(.), "[]")),
    med = as.character(as.integer(Median_Score)),
    n_agree_median = (!is.na(Score1) & Score1 == med) +
      (!is.na(Score2) & Score2 == med) +
      (!is.na(Score3) & Score3 == med),
    Agreement = factor(
      case_when(
        n_agree_median == 3 ~ "All 3 agreed",
        n_agree_median == 2 ~ "2 of 3 agreed",
        TRUE ~ "All 3 differed"
      ),
      levels = c("All 3 agreed", "2 of 3 agreed", "All 3 differed")
    )
  )
```

### 2.1 Summary table

```{r}
# 2.1
confidence_table <- conf_cda %>%
  group_by(`Median Score` = as.integer(Median_Score)) %>%
  summarise(
    Frequency = n(),
    `All-3-Agree` = sum(Agreement == "All 3 agreed"),
    `All-3-Agree %` = round(mean(Agreement == "All 3 agreed") * 100, 1),
    `2-of-3-Agree` = sum(Agreement == "2 of 3 agreed"),
    `2-of-3-Agree %` = round(mean(Agreement == "2 of 3 agreed") * 100, 1),
    `All-3-Disagree` = sum(Agreement == "All 3 differed"),
    `All-3-Disagree %` = round(mean(Agreement == "All 3 differed") * 100, 1),
    .groups = "drop"
  )
# 148 CDAs were scored by only two raters; in every case the two present
# scores match the median, so they fall in the 2-of-3-Agree group.

knitr::kable(confidence_table, align = "c")
```

### 2.2 Stacked bar

```{r fig.width=8, fig.height=6}
#| fig-width: 8
#| fig-height: 6
# 2.2
conf_counts <- conf_cda %>%
  count(`Median Score` = as.integer(Median_Score), Agreement)

agree_pal <- c(
  "All 3 agreed" = "#1a9850",
  "2 of 3 agreed" = "#fdae61",
  "All 3 differed" = "#d73027"
)

p <- ggplot(conf_counts, aes(x = factor(`Median Score`), y = n, fill = Agreement)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    values = agree_pal, name = "Rater Agreement",
    guide = guide_legend(reverse = TRUE)
  ) +
  labs(x = "Median Severity Score", y = "Frequency") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 15),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 11)
  )
p
save_plot(p, "stackedbar.png", width = 8, height = 6, dir = "thesis_plots")
```

---

## 3. Per-Rater Agreement with Median

```{r fig.width=8, fig.height=6}
#| fig-width: 8
#| fig-height: 6
# 3
# Stacked bar chart showing each rater's exact agreement and within-+/-1
# agreement with the CDA median score, computed across all CDAs scored by
# that individual. Adapted from
# AutoCDAScorer_Models/analyses/0014_Scoring_Consistency/scoring_consistency.qmd
# (human-only, no model results).
#
# Corrected for the median-of-3 tautology: for an odd number of raters, the
# median is always exactly equal to one rater's own score, so that rater
# trivially "agrees" by construction whenever all 3 raters actually
# disagree - not genuine consensus. corrected_median_agreement() (see
# src/agreement_metrics.R) zeroes out that specific case, leaving real 2+
# agreement untouched, so a rater's own score is never just compared
# against itself.
scorer_accuracy <- score_data_long %>%
  filter(!is.na(Score), Score != "[]") %>%
  mutate(Score = as.integer(Score)) %>%
  corrected_median_agreement() %>%
  group_by(Scorer) %>%
  summarise(
    exact = mean(Exact_Match),
    within_one = mean(NearMiss_Match),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(extra_within_one = within_one - exact)

avg_row <- scorer_accuracy %>%
  summarise(
    Scorer = factor("Avg"),
    exact = mean(exact),
    within_one = mean(within_one),
    extra_within_one = mean(extra_within_one),
    n = NA_integer_
  )

# Axis tick labels annotated with each rater's CDA count (n), matching the
# "(n=...)" convention used on the score-class-by-project heatmap axis in
# analyses/04_project_level_analysis.
rater_axis_labels <- scorer_accuracy %>%
  transmute(Scorer = as.character(Scorer), label = paste0(Scorer, "\n(n=", n, ")"))
x_labels <- setNames(rater_axis_labels$label, rater_axis_labels$Scorer)

# Panel-wide averages, drawn as neutral dashed reference lines (they replace the
# old stacked "Avg" bar): the lower line is mean exact agreement, the upper line
# is mean near-miss agreement (within +/-1). They are labelled on a secondary
# right-hand y-axis, so they add no extra legend entries and their colour stays
# distinct from the bars.
mean_exact <- mean(scorer_accuracy$exact)
mean_within_one <- mean(scorer_accuracy$within_one)

plot_acc <- scorer_accuracy %>%
  mutate(Scorer = factor(Scorer,
    levels = as.character(sort(as.numeric(
      unique(as.character(scorer_accuracy$Scorer))
    )))
  )) %>%
  pivot_longer(
    cols = c(exact, extra_within_one),
    names_to = "metric", values_to = "value"
  ) %>%
  mutate(metric = factor(metric,
    levels = c("extra_within_one", "exact"),
    labels = c("Near-miss agreement (±1)", "Exact agreement")
  ))

p <- ggplot(plot_acc, aes(x = Scorer, y = value, fill = metric)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = mean_exact, linetype = "dashed", colour = "grey20", linewidth = 0.6) +
  geom_hline(yintercept = mean_within_one, linetype = "dashed", colour = "grey20", linewidth = 0.6) +
  scale_fill_manual(
    values = c(
      "Exact agreement" = "steelblue",
      "Near-miss agreement (±1)" = "#E69F00"
    ),
    name = NULL,
    breaks = c("Exact agreement", "Near-miss agreement (±1)")
  ) +
  scale_x_discrete(labels = x_labels) +
  scale_y_continuous(
    labels = scales::percent, limits = c(0, 1),
    sec.axis = sec_axis(~.,
      breaks = c(mean_exact, mean_within_one),
      labels = c(
        sprintf("Mean exact\n(%.1f%%)", 100 * mean_exact),
        sprintf("Mean near-miss\n(%.1f%%)", 100 * mean_within_one)
      )
    )
  ) +
  labs(x = "Rater", y = "Corrected Agreement with Median Severity Score") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 13),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y.right = element_text(size = 11, hjust = 0),
    axis.title = element_text(size = 15),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 11),
    legend.position = "top"
  )
p
save_plot(p, "scorer_accuracy.png", width = 8, height = 6, dir = "thesis_plots")
```

The same per-rater values shown in the figure above, as a table (Exact agreement,
and total agreement within ±1 of the median; n = CDAs scored by that rater):

```{r}
# 3 (table)
scorer_accuracy_values <- bind_rows(scorer_accuracy, avg_row) %>%
  mutate(Scorer = factor(Scorer,
    levels = c(as.character(sort(as.numeric(
      unique(as.character(scorer_accuracy$Scorer))
    ))), "Avg")
  )) %>%
  arrange(Scorer) %>%
  transmute(
    Rater = as.character(Scorer),
    `CDAs Scored (n)` = ifelse(is.na(n), "", as.character(n)),
    `Exact Agreement` = scales::percent(exact, accuracy = 0.1),
    `Within ±1` = scales::percent(within_one, accuracy = 0.1)
  )

scorer_accuracy_values %>%
  knitr::kable(caption = "Per-rater corrected agreement with the median severity score")
```

### 3.1 Direction of disagreement (rater score minus median)

Ported from PhD_DL_Models/analyses/0024_EDA_Plots/EDA.qmd (anonymised raters,
same corrected-median dataset). For every CDA a rater scored, how far their score
sat from that CDA's median severity (Score - median): 0 means they matched the
consensus, positive means they scored higher than it, negative lower. Each cell is
the proportion of that rater's scores at that distance (columns sum to 100%), so
it shows the *direction* in which a rater tends to disagree, not just how often.

```{r fig.width=10, fig.height=10.4}
#| fig-width: 10
#| fig-height: 10.4
# 3.1
score_distance <- score_data_long %>%
  mutate(
    Score = as.integer(Score),
    Distance = Score - as.integer(Median_Score)
  ) %>%
  count(Scorer, Distance, name = "Count") %>%
  group_by(Scorer) %>%
  mutate(Proportion = Count / sum(Count)) %>%
  ungroup()

p <- ggplot(score_distance, aes(x = Scorer, y = Distance, fill = Proportion * 100)) +
  geom_tile(colour = "grey40") +
  geom_text(
    aes(label = sprintf("%.1f%%", Proportion * 100), colour = Proportion * 100 > 70),
    size = 4
  ) +
  scale_fill_gradient(name = "% of\nrater's\nscores", low = "white", high = "firebrick") +
  scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "black"), guide = "none") +
  scale_y_continuous(breaks = sort(unique(score_distance$Distance))) +
  labs(x = "Rater", y = "Score − median severity") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 15),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11),
    panel.grid.minor = element_blank()
  )
p
save_plot(p, "score_distance_by_rater_heatmap.png", width = 10, height = 10.4, dir = "thesis_plots")
```

Summarising the same data as the direction of each rater's bias: of the times a
rater disagreed with the median (distance != 0), what share was scored higher
versus lower. Disagreement rate is the share of all of that rater's scores that
differed from the median.

```{r}
# 3.1 (table)
rater_bias_table <- score_distance %>%
  group_by(Scorer) %>%
  summarise(
    Total = sum(Count),
    Disagreements = sum(Count[Distance != 0]),
    Higher = sum(Count[Distance > 0]),
    Lower = sum(Count[Distance < 0]),
    .groups = "drop"
  ) %>%
  arrange(Scorer) %>%
  transmute(
    Rater = as.character(Scorer),
    `Disagreements (n)` = Disagreements,
    `Disagreement rate` = scales::percent(Disagreements / Total, accuracy = 0.1),
    `Scored higher` = scales::percent(Higher / Disagreements, accuracy = 0.1),
    `Scored lower` = scales::percent(Lower / Disagreements, accuracy = 0.1)
  )

rater_bias_table %>%
  knitr::kable(caption = "Direction of each rater's disagreements with the median severity score")
```

---

## 4. Pairwise % Agreement Between Raters

```{r}
# 4
merged_df <- pairwise_score_data(score_data_long)
merged_df_no_0 <- pairwise_score_data(score_data_long_no_0)
average_agreement <- average_pairwise_agreement(merged_df)
average_agreement_no_0 <- average_pairwise_agreement(merged_df_no_0)
```

```{r fig.width=12, fig.height=10}
#| fig-width: 12
#| fig-height: 10
# 4
agree_lims <- range(c(merged_df$Percent_Agree, merged_df_no_0$Percent_Agree), na.rm = TRUE)

plot_agree <- function(df, title) {
  ggplot(df, aes(
    x = Scorer.x,
    y = factor(Scorer.y, levels = as.character(sort(unique(as.integer(Scorer.y))))),
    fill = Percent_Agree
  )) +
    geom_tile(color = "black", linewidth = 1, na.rm = TRUE) +
    geom_text(aes(label = ifelse(is.na(Percent_Agree), "", sprintf("%.1f%%", Percent_Agree))),
      color = "white", size = 6, vjust = 0.5, fontface = "bold", na.rm = TRUE
    ) +
    scale_fill_gradient(
      low = "#3182bd", high = "firebrick", name = "% Agreement",
      na.value = "white", limits = agree_lims
    ) +
    scale_x_discrete(limits = as.character(1:10)) +
    scale_y_discrete(limits = as.character(1:10)) +
    coord_cartesian(xlim = c(1, 10), ylim = c(1, 10)) +
    labs(x = "Rater X", y = "Rater Y", title = title) +
    theme_ira() +
    theme(plot.title = element_text(size = 15, face = "bold"))
}

# Saved as two separate images (rather than one patchwork-combined image) so
# the thesis figure can compose them itself.
p_agree_all <- plot_agree(merged_df, "All CDAs")
p_agree_all
save_plot(p_agree_all, "pairwise_agreement_all.png", width = 12, height = 10, dir = "thesis_plots")

p_agree_excl_zero <- plot_agree(merged_df_no_0, "Excluding median = 0")
p_agree_excl_zero
save_plot(p_agree_excl_zero, "pairwise_agreement_excl_zero.png", width = 12, height = 10, dir = "thesis_plots")
```

---

## 5. Effect of Removing Each Rater on Metric 3 Agreement

```{r}
# 5.1
# For each rater, drop their own scores and recompute Metric 3 (see
# 03_human_performance/04_project_level_analysis) over the rest - no
# "kept vs cleaned" ambiguity here (unlike the pairwise version below):
# Exact_Match is a per-CDA-rater indicator against a median already fixed
# from the full original panel, so removing one rater's rows never changes
# what any remaining row's Exact_Match value is.
median_corrected_long <- score_data_long %>%
  filter(!is.na(Score), Score != "[]") %>%
  mutate(Score = as.integer(Score)) %>%
  corrected_median_agreement()

baseline_metric3 <- mean(median_corrected_long$Exact_Match) * 100

unique_scorers_m3 <- as.character(sort(as.numeric(unique(median_corrected_long$Scorer))))

metric3_removal <- lapply(unique_scorers_m3, function(scorer) {
  remaining <- median_corrected_long %>% filter(Scorer != scorer)
  data.frame(
    Rater_Removed = scorer,
    Metric_3_Agreement = mean(remaining$Exact_Match) * 100
  )
})

metric3_result_df <- bind_rows(metric3_removal) %>%
  mutate(Delta = Metric_3_Agreement - baseline_metric3)

metric3_result_df %>% arrange(desc(Metric_3_Agreement))
```

```{r fig.width=10, fig.height=6}
#| fig-width: 10
#| fig-height: 6
# 5.2
bar_df3 <- metric3_result_df %>%
  mutate(Rater_Removed = reorder(Rater_Removed, -Delta))
bar_df3$xpos <- as.numeric(bar_df3$Rater_Removed)

p <- ggplot(bar_df3) +
  geom_rect(aes(
    xmin = xpos - 0.4, xmax = xpos + 0.4,
    ymin = baseline_metric3, ymax = Metric_3_Agreement,
    fill = Delta > 0
  )) +
  geom_hline(yintercept = baseline_metric3, linewidth = 0.6) +
  geom_text(aes(
    x = xpos, y = Metric_3_Agreement, label = sprintf("%+.2f", Delta),
    vjust = ifelse(Delta > 0, -0.4, 1.3)
  ), size = 4) +
  scale_x_continuous(breaks = sort(bar_df3$xpos), labels = levels(bar_df3$Rater_Removed)) +
  scale_y_continuous(
    expand = expansion(mult = c(0.1, 0.1)),
    sec.axis = sec_axis(~.,
      breaks = baseline_metric3,
      labels = sprintf("Original (%.1f%%)", baseline_metric3)
    )
  ) +
  scale_fill_manual(values = c("TRUE" = "firebrick", "FALSE" = "steelblue"), guide = "none") +
  labs(x = "Rater removed", y = "Metric 3 agreement (%)") +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 15),
    axis.text.y.right = element_text(face = "italic", colour = "grey35"),
    axis.ticks.y.right = element_line(colour = "grey35"),
    plot.margin = margin(12, 12, 5.5, 12)
  )
p
save_plot(p, "rater_removal_metric3_sensitivity.png", width = 10, height = 6, dir = "thesis_plots")
```

---

## 6. Agreement Patterns Between Rater Pairs

```{r fig.width=16, fig.height=16}
#| fig-width: 16
#| fig-height: 16
# 6
gen_pairwise_counts <- function(score_data_long) {
  data <- score_data_long %>%
    mutate(Score = as.numeric(Score)) %>%
    group_by(Basename, Row, Col, Pos) %>%
    mutate(id = row_number()) %>%
    ungroup()

  pairwise <- data %>%
    full_join(data, by = c("Basename", "Row", "Col", "Pos"), relationship = "many-to-many") %>%
    filter(id.x < id.y) %>%
    select(Scorer.x, Score.x, Scorer.y, Score.y) %>%
    group_by(Scorer.x, Score.x, Scorer.y, Score.y) %>%
    summarise(Count = n(), .groups = "drop") %>%
    filter(!is.na(Score.x) & !is.na(Score.y))

  # id.x < id.y reflects column position (Scorer1/2/3), not rater ID value,
  # so a given rater pair can appear with either rater in the Scorer.x slot
  # depending on which column they happened to occupy for a given CDA.
  # Folding in the mirrored copy and then keeping only Scorer.x < Scorer.y
  # (by value) consolidates both into one consistent, value-ordered
  # direction per pair - shown once, not twice.
  sym_copy <- pairwise
  names(sym_copy)[names(sym_copy) %in% c("Scorer.x", "Scorer.y")] <- c("Scorer.y", "Scorer.x")
  names(sym_copy)[names(sym_copy) %in% c("Score.x", "Score.y")] <- c("Score.y", "Score.x")
  rbind(pairwise, sym_copy) %>%
    filter(as.integer(Scorer.x) < as.integer(Scorer.y))
}

# facet_grid always draws the full cross product of its two faceting
# variables' declared factor levels, regardless of whether a given
# combination has any data - so even with only the 45 upper-triangle rater
# pairs above, it would still draw all 100 panels (self-pairs + the
# now-empty lower-triangle mirror). This blanks those 55 non-pairs out
# entirely (rather than leaving them as visible empty bordered boxes), by
# replacing their panel grobs with nullGrob() after the plot is built.
hide_non_pairs <- function(p) {
  panel_info <- ggplot_build(p)$layout$layout
  hide_panels <- panel_info$PANEL[
    as.integer(as.character(panel_info$Scorer.x)) >= as.integer(as.character(panel_info$Scorer.y))
  ]

  g <- ggplotGrob(p)
  panel_idx <- grep("^panel", g$layout$name)
  stopifnot(length(panel_idx) == nrow(panel_info))
  for (pnum in as.integer(as.character(hide_panels))) {
    g$grobs[[panel_idx[pnum]]] <- grid::nullGrob()
  }
  g
}

make_patterns_plot <- function(sdl) {
  levels_xy <- as.character(1:10)
  score_levels <- as.character(0:6)

  # Self-pairs (a rater against themselves) carry no data and are excluded
  # outright by gen_pairwise_counts's Scorer.x < Scorer.y filter above.
  counts <- gen_pairwise_counts(sdl) %>%
    mutate(
      Scorer.x = factor(Scorer.x, levels = levels_xy),
      Scorer.y = factor(Scorer.y, levels = levels_xy),
      Score.x = factor(Score.x, levels = score_levels),
      Score.y = factor(Score.y, levels = score_levels)
    )

  # Complete the severity-score grid (0-6 x 0-6) *within* each rater pair
  # that actually shares data (nesting), rather than crossing all 10x10
  # rater combinations. The latter used to fabricate entirely-empty panels
  # (self-pairs, or raters who never scored together) whose Score.x/Score.y
  # were then all NA, which made ggplot's per-panel range() warn "no
  # non-missing arguments to min/max" for every such panel.
  complete_counts <- counts %>%
    complete(
      nesting(Scorer.x, Scorer.y),
      Score.x = factor(score_levels, levels = score_levels),
      Score.y = factor(score_levels, levels = score_levels),
      fill = list(Count = 0)
    ) %>%
    mutate(
      Count_Binned = cut(Count,
        breaks = c(0, 1, 10, 100, 250, Inf),
        labels = c("1", "2-10", "11-100", "101-250", "251-1000"),
        include.lowest = TRUE
      ),
      Scorer.y = fct_rev(Scorer.y)
    )

  p <- ggplot(complete_counts, aes(x = Score.x, y = Score.y)) +
    geom_tile(aes(fill = Count_Binned), color = "black", linewidth = 0.5) +
    # Score.x/Score.y are discrete factors (see above), so the exact-agreement
    # diagonal is drawn as a segment between the first and last factor
    # positions rather than geom_abline (which needs a continuous scale).
    geom_segment(
      x = 1, xend = length(score_levels), y = 1, yend = length(score_levels),
      linetype = "dashed", color = "grey40", inherit.aes = FALSE
    ) +
    scale_fill_manual(
      values = c(
        "1" = "#eff3ff",
        "2-10" = "#9ecae1",
        "11-100" = "#3182bd",
        "101-250" = "#fdae61",
        "251-1000" = "#d73027"
      ),
      name = "Frequency", na.value = "white", drop = FALSE
    ) +
    facet_grid(Scorer.y ~ Scorer.x, switch = "both") +
    theme_bw() +
    labs(x = "Rater X", y = "Rater Y") +
    theme(
      panel.spacing = unit(0.25, "lines"),
      strip.placement = "outside",
      panel.grid = element_blank(),
      strip.text = element_text(size = 11),
      axis.text = element_text(size = 10),
      axis.title = element_text(size = 13),
      legend.title = element_text(size = 13),
      legend.text = element_text(size = 11)
    )

  hide_non_pairs(p)
}

p_patterns <- make_patterns_plot(score_data_long)
grid::grid.newpage()
grid::grid.draw(p_patterns)
# Matches the chunk's fig.width/fig.height (16 x 16) - see
# 06_intra_rater_reliability's equivalent plots for the repo's convention of
# always matching the two.
save_plot(p_patterns, "agreement_patterns.png", width = 16, height = 16)
```

---

## 7. Krippendorff's Alpha — Raw Scores

```{r}
# Shared plotting helper for 7.1-7.3.
# The mean and threshold lines are geom_segments bounded to the bar heights
# (not full-height geom_vlines), so they physically cannot reach the label
# row above - the dashed/dotted line can't cover the text because it stops
# below it. Both labels sit at the same height, left-justified from their
# own line's x-position, growing rightward. The one exception: when the
# threshold (0.8) sits close enough to the mean that its left-justified
# label would run into the mean's label - as happens in 7.2's non-zero panel,
# where they differ by ~0.005 - the threshold label is right-justified
# instead, so it grows leftward, away from the mean label.
make_alpha_hist_panel <- function(boot_vec, xlab) {
  mean_val <- mean(boot_vec)
  threshold <- 0.8

  base <- ggplot(data.frame(alpha = boot_vec), aes(x = alpha)) +
    geom_histogram(binwidth = 0.005, fill = "cornflowerblue", color = "white")
  max_count <- max(ggplot_build(base)$data[[1]]$count)
  line_top <- max_count * 1.06
  label_y <- max_count * 1.15

  threshold_hjust <- if (abs(mean_val - threshold) < 0.02) 1 else 0

  base +
    geom_segment(
      x = mean_val, xend = mean_val, y = 0, yend = line_top,
      color = "firebrick", linetype = "dashed", linewidth = 1
    ) +
    geom_segment(
      x = threshold, xend = threshold, y = 0, yend = line_top,
      color = "black", linetype = "dotted", linewidth = 0.8
    ) +
    annotate("text",
      x = mean_val, y = label_y, hjust = 0, size = 4, color = "firebrick",
      label = sprintf("Mean=%.3f", mean_val)
    ) +
    annotate("text",
      x = threshold, y = label_y, hjust = threshold_hjust, size = 4, color = "black",
      label = "α=0.8"
    ) +
    scale_y_continuous(limits = c(0, max_count * 1.3), expand = expansion(mult = c(0.02, 0))) +
    coord_cartesian(clip = "off") +
    labs(x = xlab, y = "Count") +
    theme_minimal() +
    theme(
      axis.text = element_text(size = 13),
      axis.title = element_text(size = 15)
    )
}
```

### 7.1 Ordinal metric

```{r}
# 7.1.1
# Computed on the full rater x CDA matrix directly (no median involved).
# Two distance metrics are compared here and in 7.3 as a robustness check on
# the assumed measurement level: ordinal (treats 0-6 as ordered categories)
# and interval (treats 0-6 as equally spaced). If the two estimates agree,
# the reliability conclusion doesn't hinge on which scale assumption is
# correct. Missing values are handled automatically: expected agreement is
# estimated from all rated pairs.
alpha_raw <- kripp.alpha(scorer_matrix, method = "ordinal")
cat("Krippendorff's alpha (ordinal, raw scores):", round(alpha_raw$value, 4), "\n")
```

```{r}
# 7.1.2
# Subsample distribution: 1,000 iterations, n = 500 CDAs per draw, without
# replacement (see bootstrap_kripp_alpha() docs - "bootstrap" in the function
# name refers to the general repeated-resampling technique, but this call
# draws fixed-size subsamples rather than classic with-replacement bootstrap
# resamples).
alpha_raw_boot <- bootstrap_kripp_alpha(
  scorer_matrix,
  method = "ordinal", sample_size = 500,
  cache_path = here("analyses", "05_inter_rater_agreement", "bootstrap_cache", "alpha_raw_ordinal.rds")
)

cat("Subsample mean α:", round(mean(alpha_raw_boot), 4), "\n")
cat(
  "95% CI: [", round(quantile(alpha_raw_boot, 0.025), 4), ",",
  round(quantile(alpha_raw_boot, 0.975), 4), "]\n"
)
```

```{r fig.width=7, fig.height=3.2}
#| fig-width: 7
#| fig-height: 3.2
# 7.1.3
p <- make_alpha_hist_panel(alpha_raw_boot, "Krippendorff's α (ordinal, all CDAs)")
p
save_plot(p, "krippendorff_alpha_raw.png", width = 7, height = 3.2, dir = "thesis_plots")
```

### 7.2 Ordinal metric — non-zero-median CDAs

```{r}
# 7.2.1
# Same rater x CDA matrix as 7.1, restricted to CDAs whose aggregated median
# score is non-zero (same restriction already applied to pairwise agreement
# in section 3 via score_data_no_0). Checks whether 7.1's headline ordinal
# alpha is inflated by the abundant, near-unanimous median-0 CDAs, where
# raters trivially agree "no cell death".
score_data_numeric_no_0 <- score_data_numeric %>%
  filter(Median_Score > 0)

scorer_matrix_no_0 <- build_scorer_matrix(score_data_numeric_no_0)
cat("Rater matrix dimensions:", nrow(scorer_matrix_no_0), "raters ×", ncol(scorer_matrix_no_0), "CDAs\n")

alpha_raw_no0 <- kripp.alpha(scorer_matrix_no_0, method = "ordinal")
cat("Krippendorff's alpha (ordinal, non-zero-median CDAs):", round(alpha_raw_no0$value, 4), "\n")
```

```{r}
# 7.2.2
# Subsample distribution: 1,000 iterations, n = 500 CDAs per draw (see 7.1.2
# on "subsample" vs. classic bootstrap terminology).
alpha_raw_no0_boot <- bootstrap_kripp_alpha(
  scorer_matrix_no_0,
  method = "ordinal", sample_size = 500,
  cache_path = here("analyses", "05_inter_rater_agreement", "bootstrap_cache", "alpha_raw_ordinal_no0.rds")
)

cat("Subsample mean α:", round(mean(alpha_raw_no0_boot), 4), "\n")
cat(
  "95% CI: [", round(quantile(alpha_raw_no0_boot, 0.025), 4), ",",
  round(quantile(alpha_raw_no0_boot, 0.975), 4), "]\n"
)
```

```{r fig.width=7, fig.height=3.2}
#| fig-width: 7
#| fig-height: 3.2
# 7.2.3
p <- make_alpha_hist_panel(alpha_raw_no0_boot, "Krippendorff's α (ordinal, non-zero-median CDAs)")
p
save_plot(p, "krippendorff_alpha_raw_no0.png", width = 7, height = 3.2, dir = "thesis_plots")
```

### 7.3 Interval metric

```{r}
# 7.3.1
# Same rater x CDA matrix as 7.1, but with method = "interval" instead of
# "ordinal" - no residuals or medians involved, just a different distance
# metric applied to the same raw scores.
alpha_raw_interval <- kripp.alpha(scorer_matrix, method = "interval")
cat("Krippendorff's alpha (interval, raw scores):", round(alpha_raw_interval$value, 4), "\n")
```

```{r}
# 7.3.2
# Subsample distribution: 1,000 iterations, n = 500 CDAs per draw (see 7.1.2
# on "subsample" vs. classic bootstrap terminology).
alpha_raw_interval_boot <- bootstrap_kripp_alpha(
  scorer_matrix,
  method = "interval", sample_size = 500,
  cache_path = here("analyses", "05_inter_rater_agreement", "bootstrap_cache", "alpha_raw_interval.rds")
)

cat("Subsample mean α:", round(mean(alpha_raw_interval_boot), 4), "\n")
cat(
  "95% CI: [", round(quantile(alpha_raw_interval_boot, 0.025), 4), ",",
  round(quantile(alpha_raw_interval_boot, 0.975), 4), "]\n"
)
```

```{r fig.width=7, fig.height=3.2}
#| fig-width: 7
#| fig-height: 3.2
# 7.3.3
p <- make_alpha_hist_panel(alpha_raw_interval_boot, "Krippendorff's α (interval, all CDAs)")
p
save_plot(p, "krippendorff_alpha_raw_interval.png", width = 7, height = 3.2, dir = "thesis_plots")
```

---

## 8. Interpretation and Discussion

_Add notes here._

---

## 9. Additional Visualisation

### 9.1 Shared CDA Count per Rater Pair

```{r fig.width=11, fig.height=10}
#| fig-width: 11
#| fig-height: 10
# 9.1
p <- ggplot(
  merged_df,
  aes(
    x = Scorer.x,
    y = factor(Scorer.y, levels = as.character(sort(unique(as.integer(Scorer.y))))),
    fill = Total
  )
) +
  geom_tile(color = "black", linewidth = 1) +
  geom_text(aes(label = ifelse(is.na(Total), "", Total)), color = "black", size = 5, vjust = 0.5) +
  scale_fill_gradient(low = "white", high = "firebrick", name = "Shared CDAs", na.value = "white") +
  labs(x = "Rater X", y = "Rater Y") +
  theme_ira()
p
save_plot(p, "shared_cda_count.png", width = 11, height = 10)
```
