---
title: "Confident Scores"
format:
  html:
    code-fold: true
---

Date: 2/6/2024

## GOHREP

**Goal:**

Quantify, for each median (ground-truth) severity score, how often the three raters agreed. Each CDA falls into one of three mutually exclusive groups: all three raters agreed (unanimous), two of three agreed with the median (majority), or all three gave a different score (no agreement).

**Hypothesis:**

Extreme scores (0 and 6) will have a higher proportion of unanimous annotations than intermediate scores, as these cases are most visually clear.

**Rationale:**

Understanding which score classes have reliable ground truth labels informs model training and evaluation strategies.

---

## 1. Setup

```{r}
# 1.1
library(readr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(here)
```

```{r}
# 1.2
source(here("src", "plotting_helpers.R"))
save_plot <- make_save_plot("05_inter_rater_agreement")
```

```{r}
# 1.3
path_data <- here("analyses", "01_generate_raw_data", "04_median_and_centre_coords", "combined_cda_data_median.csv")

raw_data <- read_csv(path_data, show_col_types = FALSE)

# One row per CDA. Missing scores are stored as "[]"; treat them as NA.
# Agreement is defined by how many of the three scorers match the median score:
#   3 -> all three agreed, 2 -> two of three agreed, 1 -> all three differed.
conf_cda <- raw_data %>%
  select(Basename, Row, Col, Pos, Score1, Score2, Score3, Median_Score) %>%
  mutate(across(c(Score1, Score2, Score3), ~ na_if(as.character(.), "[]")),
    med = as.character(as.integer(Median_Score)),
    n_agree_median = (!is.na(Score1) & Score1 == med) +
      (!is.na(Score2) & Score2 == med) +
      (!is.na(Score3) & Score3 == med),
    Agreement = factor(
      case_when(
        n_agree_median == 3 ~ "All 3 agreed",
        n_agree_median == 2 ~ "2 of 3 agreed",
        TRUE ~ "All 3 differed"
      ),
      levels = c("All 3 agreed", "2 of 3 agreed", "All 3 differed")
    )
  )
```

---

## 2. Agreement breakdown by median score

```{r}
# 2
confidence_table <- conf_cda %>%
  group_by(`Median Score` = as.integer(Median_Score)) %>%
  summarise(
    Frequency = n(),
    `All-3-Agree` = sum(Agreement == "All 3 agreed"),
    `All-3-Agree %` = round(mean(Agreement == "All 3 agreed") * 100, 1),
    `2-of-3-Agree` = sum(Agreement == "2 of 3 agreed"),
    `2-of-3-Agree %` = round(mean(Agreement == "2 of 3 agreed") * 100, 1),
    `All-3-Disagree` = sum(Agreement == "All 3 differed"),
    `All-3-Disagree %` = round(mean(Agreement == "All 3 differed") * 100, 1),
    .groups = "drop"
  )

knitr::kable(confidence_table, align = "c")
```

Each CDA is categorised by how many of its three raters agreed with the median: *All-3-Agree* (unanimous), *2-of-3-Agree* (a majority match the median, one differs), and *All-3-Disagree* (all three scores distinct). Percentages are out of the total number of CDAs at each median score. Note that 148 CDAs were scored by only two raters; in every case the two present scores match the median, so they fall in the *2-of-3-Agree* group.

---

## 3. Stacked bar of agreement by median score

```{r fig.width=8, fig.height=6}
#| fig-width: 8
#| fig-height: 6
# 3
conf_counts <- conf_cda %>%
  count(`Median Score` = as.integer(Median_Score), Agreement)

agree_pal <- c(
  "All 3 agreed" = "#1a9850",
  "2 of 3 agreed" = "#fdae61",
  "All 3 differed" = "#d73027"
)

p <- ggplot(conf_counts, aes(x = factor(`Median Score`), y = n, fill = Agreement)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    values = agree_pal, name = "Rater Agreement",
    guide = guide_legend(reverse = TRUE)
  ) +
  labs(x = "Median Severity Score", y = "Frequency") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 15),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 11)
  )
p
save_plot(p, "stackedbar.png", width = 8, height = 6)
```

**What this shows:**

- Median-0 CDAs are overwhelmingly unanimous, reflecting the dominance of negative controls and their visual clarity.
- Median-6 is the next most unanimous; both extremes are easier to identify consistently.
- Intermediate scores (1–5) are dominated by partial or full disagreement, highlighting the difficulty of distinguishing adjacent severity levels.
- All-three-differed is only possible for intermediate medians, since an extreme median (0 or 6) requires at least two raters to have given that extreme value.

---

## 4. Discussion

### 4.1 Agreement breakdown by median score

_Add notes here._

### 4.2 Stacked bar of agreement by median score

_Add notes here._

### 4.3 Overall narrative

_Add notes here._
