---
title: "Inter-Rater Reliability: Krippendorff's Alpha and Residual Correlation"
format:
  html:
    code-fold: true
---

Date: 2026-06-03

## GOHREP

**Goal:**

Quantify the overall inter-rater reliability of the human scorer panel using Krippendorff's alpha, and characterise the structure of scorer disagreement via Pearson correlation of score residuals.

**Hypothesis:**

1. Krippendorff's alpha on raw scores will be high (>0.8), reflecting that scorers broadly agree on severity ordering even when they differ by ±1.
2. Alpha on residuals (score − median) will be substantially lower, indicating that scorers do not share structured error patterns.
3. Residual correlations between scorer pairs will be positive but moderate — scorers make similar errors but not identical ones.

**Rationale:**

Pairwise % agreement (see `pairwise_agreement.qmd`) captures agreement on exact score matches but does not give a single overall reliability estimate. Krippendorff's alpha is the standard metric for ordinal inter-rater reliability and handles missing data (not all scorers rated all CDAs). Residual correlation is ported from the model ensemble analysis in `AutoCDAScorer_Models/analyses/10_model_scoring_patterns/model_error_independence.Rmd` where it was used to assess whether model errors were structured or independent.

**Experimental plan:**

Compute Krippendorff's alpha on raw scores (ordinal metric) and on score residuals (interval metric), bootstrap each to obtain 95% CIs, then compute and visualise pairwise Pearson correlation of residuals.

---

## 1. Setup

### 1.1 Load required packages

```{r}
# 1.1
library(readr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(here)
library(irr)
```

### 1.2 Load data

```{r}
# 1.2
path_data <- here("analyses", "01_generate_raw_data", "04_median_and_centre_coords", "combined_cda_data_median.csv")

raw_data <- read_csv(path_data, show_col_types = FALSE)

score_data <- raw_data %>%
  select(Basename, Row, Col, Pos, Scorer1, Score1, Scorer2, Score2, Scorer3, Score3, Median_Score) %>%
  mutate(across(c(Scorer1, Scorer2, Scorer3), as.character),
    across(c(Score1, Score2, Score3), as.numeric),
    Median_Score = as.numeric(Median_Score)
  )

head(score_data)
```

### 1.3 Helper functions

```{r}
# 1.3
source(here("src", "agreement_metrics.R"))
source(here("src", "plotting_helpers.R"))

save_plot <- make_save_plot("05_inter_rater_agreement")

scorer_matrix <- build_scorer_matrix(score_data)
cat("Scorer matrix dimensions:", nrow(scorer_matrix), "scorers ×", ncol(scorer_matrix), "CDAs\n")
cat("Non-NA cells:", sum(!is.na(scorer_matrix)), "/", prod(dim(scorer_matrix)), "\n")
```

---

## 2. Krippendorff's Alpha — Raw Scores

Krippendorff's alpha is computed on the full scorer × CDA matrix directly (no median involved). Two distance metrics are compared as a robustness check on the assumed measurement level: **ordinal** (treats 0–6 as ordered categories) and **interval** (treats 0–6 as equally spaced). If the two estimates agree, the reliability conclusion doesn't hinge on which scale assumption is correct. Missing values are handled automatically: expected agreement is estimated from all rated pairs.

### 2.1 Ordinal metric

```{r}
# 2.1.1
alpha_raw <- kripp.alpha(scorer_matrix, method = "ordinal")
cat("Krippendorff's alpha (ordinal, raw scores):", round(alpha_raw$value, 4), "\n")
```

Bootstrap distribution (1,000 iterations, n = 500 CDAs per draw):

```{r}
# 2.1.2
alpha_raw_boot <- bootstrap_kripp_alpha(
  scorer_matrix,
  method = "ordinal", sample_size = 500,
  cache_path = here("analyses", "05_inter_rater_agreement", "bootstrap_cache", "alpha_raw_ordinal.rds")
)

cat("Bootstrap mean α:", round(mean(alpha_raw_boot), 4), "\n")
cat(
  "95% CI: [", round(quantile(alpha_raw_boot, 0.025), 4), ",",
  round(quantile(alpha_raw_boot, 0.975), 4), "]\n"
)
```

```{r fig.width=7, fig.height=5}
#| fig-width: 7
#| fig-height: 5
# 2.1.3
boot_df <- data.frame(alpha = alpha_raw_boot)

p <- ggplot(boot_df, aes(x = alpha)) +
  geom_histogram(binwidth = 0.005, fill = "cornflowerblue", color = "white") +
  geom_vline(aes(
    xintercept = mean(alpha_raw_boot), color = "Bootstrap mean",
    linetype = "Bootstrap mean"
  ), linewidth = 1) +
  geom_vline(aes(
    xintercept = 0.8, color = "α = 0.8 reliability threshold",
    linetype = "α = 0.8 reliability threshold"
  ), linewidth = 0.8) +
  scale_color_manual(
    name = NULL,
    values = c(
      "Bootstrap mean" = "firebrick",
      "α = 0.8 reliability threshold" = "black"
    )
  ) +
  scale_linetype_manual(
    name = NULL,
    values = c(
      "Bootstrap mean" = "dashed",
      "α = 0.8 reliability threshold" = "dotted"
    )
  ) +
  labs(x = "Krippendorff's α (ordinal)", y = "Count") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 15),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 11),
    legend.position = "bottom"
  )
p
save_plot(p, "krippendorff_alpha_raw.png", width = 7, height = 5)
```

**What this shows:**

- The bootstrap distribution of ordinal Krippendorff's alpha on raw scores, centred on the bootstrap mean.
- The dotted line marks the Hayes & Krippendorff reliability threshold of α = 0.8.
- The width of the distribution reflects sampling variability across random draws of 500 CDAs.

**Discussion:**
_Add notes here._

### 2.2 Interval metric

Same scorer × CDA matrix as §2.1, but with `method = "interval"` instead of `"ordinal"` — no residuals or medians involved, just a different distance metric applied to the same raw scores.

```{r}
# 2.2.1
alpha_raw_interval <- kripp.alpha(scorer_matrix, method = "interval")
cat("Krippendorff's alpha (interval, raw scores):", round(alpha_raw_interval$value, 4), "\n")
```

Bootstrap distribution (1,000 iterations, n = 500 CDAs per draw):

```{r}
# 2.2.2
alpha_raw_interval_boot <- bootstrap_kripp_alpha(
  scorer_matrix,
  method = "interval", sample_size = 500,
  cache_path = here("analyses", "05_inter_rater_agreement", "bootstrap_cache", "alpha_raw_interval.rds")
)

cat("Bootstrap mean α:", round(mean(alpha_raw_interval_boot), 4), "\n")
cat(
  "95% CI: [", round(quantile(alpha_raw_interval_boot, 0.025), 4), ",",
  round(quantile(alpha_raw_interval_boot, 0.975), 4), "]\n"
)
```

```{r fig.width=7, fig.height=5}
#| fig-width: 7
#| fig-height: 5
# 2.2.3
boot_interval_df <- data.frame(alpha = alpha_raw_interval_boot)

p <- ggplot(boot_interval_df, aes(x = alpha)) +
  geom_histogram(binwidth = 0.005, fill = "cornflowerblue", color = "white") +
  geom_vline(aes(
    xintercept = mean(alpha_raw_interval_boot), color = "Bootstrap mean",
    linetype = "Bootstrap mean"
  ), linewidth = 1) +
  geom_vline(aes(
    xintercept = 0.8, color = "α = 0.8 reliability threshold",
    linetype = "α = 0.8 reliability threshold"
  ), linewidth = 0.8) +
  scale_color_manual(
    name = NULL,
    values = c(
      "Bootstrap mean" = "firebrick",
      "α = 0.8 reliability threshold" = "black"
    )
  ) +
  scale_linetype_manual(
    name = NULL,
    values = c(
      "Bootstrap mean" = "dashed",
      "α = 0.8 reliability threshold" = "dotted"
    )
  ) +
  labs(x = "Krippendorff's α (interval)", y = "Count") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 15),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 11),
    legend.position = "bottom"
  )
p
save_plot(p, "krippendorff_alpha_raw_interval.png", width = 7, height = 5)
```

**What this shows:**

- The bootstrap distribution of interval Krippendorff's alpha on the same raw scores as §2.1.
- Near-identical to the ordinal estimate (§2.1) — reliability here is robust to whether the 0–6 scale is treated as ordered categories or as equally-spaced values.

**Discussion:**
_Add notes here._

---

## 3. Krippendorff's Alpha — Score Residuals

Residuals are computed as (individual score − CDA median score), then Krippendorff's alpha is recomputed on the residual matrix. A high residual alpha would indicate that scorers share structured error patterns (they make the same mistakes); a low residual alpha indicates independent errors.

```{r}
# 3
# Align median scores to scorer_matrix column order by CDA_ID
cda_medians <- raw_data %>%
  mutate(CDA_ID = paste(Basename, Row, Col, Pos, sep = "_")) %>%
  select(CDA_ID, Median_Score) %>%
  distinct() %>%
  {
    setNames(.$Median_Score, .$CDA_ID)
  }

median_vec <- as.numeric(cda_medians[colnames(scorer_matrix)])

residual_matrix <- scorer_matrix - matrix(
  rep(median_vec, each = nrow(scorer_matrix)),
  nrow = nrow(scorer_matrix),
  dimnames = dimnames(scorer_matrix)
)

alpha_resid <- kripp.alpha(residual_matrix, method = "interval")
cat("Krippendorff's alpha (interval, residuals):", round(alpha_resid$value, 4), "\n")
```

### 3.1 Bootstrap Distribution of Alpha on Residuals

```{r}
# 3.1.1
alpha_resid_boot <- bootstrap_kripp_alpha(
  residual_matrix,
  method = "interval", sample_size = 500,
  cache_path = here("analyses", "05_inter_rater_agreement", "bootstrap_cache", "alpha_resid_interval.rds")
)

cat("Bootstrap mean α (residuals):", round(mean(alpha_resid_boot), 4), "\n")
cat(
  "95% CI: [", round(quantile(alpha_resid_boot, 0.025), 4), ",",
  round(quantile(alpha_resid_boot, 0.975), 4), "]\n"
)
```

```{r fig.width=7, fig.height=5}
#| fig-width: 7
#| fig-height: 5
# 3.1.2
boot_resid_df <- data.frame(alpha = alpha_resid_boot)

p <- ggplot(boot_resid_df, aes(x = alpha)) +
  geom_histogram(binwidth = 0.005, fill = "cornflowerblue", color = "white") +
  geom_vline(aes(
    xintercept = mean(alpha_resid_boot), color = "Bootstrap mean",
    linetype = "Bootstrap mean"
  ), linewidth = 1) +
  scale_color_manual(name = NULL, values = c("Bootstrap mean" = "firebrick")) +
  scale_linetype_manual(name = NULL, values = c("Bootstrap mean" = "dashed")) +
  labs(x = "Krippendorff's α (interval, residuals)", y = "Count") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 15),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 11),
    legend.position = "bottom"
  )
p
save_plot(p, "krippendorff_alpha_residuals.png", width = 7, height = 5)
```

**What this shows:**

- The bootstrap distribution of interval Krippendorff's alpha computed on score residuals (score − median), centred on the bootstrap mean.
- A low alpha here indicates that scorer errors are largely independent — scorers do not share structured biases across the same CDAs.
- Comparing this distribution to the raw-score alpha (section 2) shows how much of the inter-rater agreement is driven by absolute severity ordering vs shared error patterns.

**Discussion:**
_Add notes here._

---

## 4. Pearson Correlation of Score Residuals

For each pair of scorers who shared at least one CDA, compute the Pearson correlation of their residuals across those shared CDAs. A high correlation means their errors tend to go in the same direction; a low or near-zero correlation means their errors are largely independent (good for ensemble/median robustness).

```{r}
# 4.1
# rownames(scorer_matrix) are in pivot order, not sorted, so map each scorer ID
# back to its matrix row and build pairs ordered by ID (Scorer.x < Scorer.y).
# This keeps every tile in the upper-left triangle like the other heatmaps.
scorer_ids <- as.integer(rownames(scorer_matrix))
row_of <- setNames(seq_along(scorer_ids), scorer_ids)

pairs <- expand.grid(Scorer.x = scorer_ids, Scorer.y = scorer_ids) %>%
  filter(Scorer.x < Scorer.y)

cor_results <- pairs %>%
  rowwise() %>%
  mutate(
    i = row_of[[as.character(Scorer.x)]],
    j = row_of[[as.character(Scorer.y)]],
    shared = sum(!is.na(residual_matrix[i, ]) & !is.na(residual_matrix[j, ])),
    correlation = if (shared >= 10) {
      cor(residual_matrix[i, ], residual_matrix[j, ],
        use = "pairwise.complete.obs"
      )
    } else {
      NA_real_
    }
  ) %>%
  ungroup() %>%
  select(Scorer.x, Scorer.y, shared, correlation)

print(
  cor_results %>%
    select(Scorer.x, Scorer.y, shared, correlation) %>%
    arrange(correlation)
)
```

```{r fig.width=11, fig.height=10}
#| fig-width: 11
#| fig-height: 10
# 4.2
p <- ggplot(
  cor_results,
  aes(
    x = factor(Scorer.x, levels = as.character(sort(unique(Scorer.x)))),
    y = factor(Scorer.y, levels = as.character(sort(unique(Scorer.y))))
  )
) +
  geom_tile(aes(fill = correlation), color = "black", linewidth = 1) +
  geom_text(
    aes(label = ifelse(is.na(correlation), "",
      sprintf("%.2f\n(n=%d)", correlation, shared)
    )),
    color = "black", size = 4, lineheight = 0.9
  ) +
  scale_fill_gradient2(
    low = "#3182bd", mid = "white", high = "firebrick",
    midpoint = 0, name = "Pearson r",
    na.value = "white", limits = c(-1, 1)
  ) +
  labs(x = "Scorer X", y = "Scorer Y") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 15),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 11)
  )
p
save_plot(p, "residual_correlation.png", width = 11, height = 10)
```

**What this shows:**

- Pairs with positive correlation share structured biases (e.g., both tend to score high or low on the same CDAs).
- Pairs near zero have largely independent errors — useful for the median being a robust ground truth.
- Blank / grey cells indicate pairs who shared too few CDAs (<10) to estimate a reliable correlation.

**Discussion:**
_Add notes here._
