---
title: "Pairwise Agreement and Per-Image Analysis"
format:
  html:
    code-fold: true
---

Date: 15/5/2024 – 10/6/2024

## GOHREP

**Goal:**

Investigate changes to average pairwise agreement when each rater is removed. Investigate differences in agreement across source agroinfiltration images.

**Hypothesis:**

Excluding scores from raters with low pairwise agreement will increase the overall average pairwise agreement. Some images will be scored more consistently than others due to differences in clarity or familiarity.

**Rationale:**

Differences in lighting or leaf clarity could cause some images to be scored less consistently. Identifying which raters most reduce overall agreement informs future data collection decisions.

---

## 1. Setup

### 1.1 Load required packages

```{r}
#| output: false
# 1.1
library(readr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(forcats)
library(patchwork)
library(scales)
library(here)
```

### 1.2 Load data

```{r}
# 1.2
path_data <- here("analyses", "01_generate_raw_data", "04_median_and_centre_coords", "combined_cda_data_median.csv")

raw_data <- read_csv(path_data, show_col_types = FALSE)

score_data <- raw_data %>%
  select(Basename, Row, Col, Pos, Scorer1, Score1, Scorer2, Score2, Scorer3, Score3, Median_Score) %>%
  mutate(
    across(c(Scorer1, Scorer2, Scorer3), as.character),
    across(c(Score1, Score2, Score3), as.character)
  )

score_data_no_0 <- score_data %>%
  filter(Median_Score != 0)

n_cdas <- nrow(score_data)
head(score_data)
```

### 1.3 Helper functions

```{r}
# 1.3.1
source(here("src", "data_helpers.R"))
source(here("src", "agreement_metrics.R"))
source(here("src", "plotting_helpers.R"))

save_plot <- make_save_plot("05_inter_rater_agreement")

theme_ira <- function() {
  theme_minimal() +
    theme(
      axis.text = element_text(size = 13),
      axis.title = element_text(size = 15),
      legend.title = element_text(size = 13),
      legend.text = element_text(size = 11),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_blank(),
      legend.key.size = unit(1.5, "lines")
    )
}
```

```{r}
# 1.3.2
score_data_long <- long_score_table(score_data)
score_data_long_no_0 <- long_score_table(score_data_no_0)
```

---

## 2. Pairwise % Agreement Between Raters

```{r}
# 2
merged_df <- pairwise_score_data(score_data_long)
merged_df_no_0 <- pairwise_score_data(score_data_long_no_0)
average_agreement <- average_pairwise_agreement(merged_df)
cat("Average pairwise agreement (all CDAs):", round(average_agreement, 2), "%\n")
cat("Average pairwise agreement (excl. median = 0):", round(average_pairwise_agreement(merged_df_no_0), 2), "%\n")
```

```{r fig.width=22, fig.height=10}
#| fig-width: 22
#| fig-height: 10
# 2
agree_lims <- range(c(merged_df$Percent_Agree, merged_df_no_0$Percent_Agree), na.rm = TRUE)

plot_agree <- function(df, title) {
  ggplot(df, aes(
    x = Scorer.x,
    y = factor(Scorer.y, levels = as.character(sort(unique(as.integer(Scorer.y))))),
    fill = Percent_Agree
  )) +
    geom_tile(color = "black", linewidth = 1) +
    geom_text(aes(label = ifelse(is.na(Percent_Agree), "", sprintf("%.1f%%", Percent_Agree))),
      color = "white", size = 6, vjust = 0.5, fontface = "bold"
    ) +
    scale_fill_gradient(
      low = "#3182bd", high = "firebrick", name = "% Agreement",
      na.value = "white", limits = agree_lims
    ) +
    scale_x_discrete(limits = as.character(1:10)) +
    scale_y_discrete(limits = as.character(1:10)) +
    coord_cartesian(xlim = c(1, 10), ylim = c(1, 10)) +
    labs(x = "Rater X", y = "Rater Y", title = title) +
    theme_ira() +
    theme(plot.title = element_text(size = 15, face = "bold"))
}

p_agree <- plot_agree(merged_df, "All CDAs") +
  plot_agree(merged_df_no_0, "Excluding median = 0") +
  plot_layout(guides = "collect")
p_agree
save_plot(p_agree, "pairwise_agreement.png", width = 22, height = 10)
```

### 2.1 Off-by-one disagreement rate

How often do two raters differ by *exactly* one severity point? Pooled across all
unordered rater pairs per CDA, the off-by-one rate is the near-miss rate (|diff| ≤ 1)
minus the exact rate (diff = 0).

```{r}
# 2.1
pairwise_diffs <- function(score_data_long) {
  d <- score_data_long %>%
    mutate(Score = as.numeric(Score)) %>%
    group_by(Basename, Row, Col, Pos) %>%
    mutate(id = row_number()) %>%
    ungroup()

  d %>%
    full_join(d,
      by = c("Basename", "Row", "Col", "Pos"),
      relationship = "many-to-many"
    ) %>%
    filter(id.x < id.y) %>%
    transmute(
      Median_Score = as.numeric(Median_Score.x),
      absdiff = abs(Score.x - Score.y)
    )
}

pdiff <- pairwise_diffs(score_data_long)

cat(sprintf("Pairwise comparisons:                     %d\n", nrow(pdiff)))
cat(sprintf("Exact agreement (diff = 0):               %.2f%%\n", mean(pdiff$absdiff == 0) * 100))
cat(sprintf("Near-miss agreement (diff <= 1):          %.2f%%\n", mean(pdiff$absdiff <= 1) * 100))
cat(sprintf("Off-by-one disagreement (diff == 1):      %.2f%%\n", mean(pdiff$absdiff == 1) * 100))
cat(sprintf(
  "Off-by-one disagreement, excl. median 0:  %.2f%%\n",
  mean(pdiff$absdiff[pdiff$Median_Score > 0] == 1) * 100
))
```

**What this shows:**

- Across all CDAs, two raters differ by exactly one severity point ~29% of the time. This is dominated by median-0 CDAs, on which raters almost always agree exactly.
- Restricting to non-zero-median CDAs, the off-by-one rate rises to ~45%: on the CDAs that actually carry disease signal, an off-by-one disagreement is the single most common outcome between any two raters.

---

## 3. Shared CDA Count per Rater Pair

How many CDAs did each pair of raters both score?

```{r fig.width=11, fig.height=10}
#| fig-width: 11
#| fig-height: 10
# 3
p <- ggplot(
  merged_df,
  aes(
    x = Scorer.x,
    y = factor(Scorer.y, levels = as.character(sort(unique(as.integer(Scorer.y))))),
    fill = Total
  )
) +
  geom_tile(color = "black", linewidth = 1) +
  geom_text(aes(label = ifelse(is.na(Total), "", Total)), color = "black", size = 5, vjust = 0.5) +
  scale_fill_gradient(low = "white", high = "firebrick", name = "Shared CDAs", na.value = "white") +
  labs(x = "Rater X", y = "Rater Y") +
  theme_ira()
p
save_plot(p, "shared_cda_count.png", width = 11, height = 10)
```

---

## 4. Effect of Removing Each Rater on Average Pairwise Agreement

```{r}
# 4
remove_scorer <- function(score_data_long, scorer, remove_2_disagree) {
  subset_score_data <- subset(score_data_long, Scorer != scorer)

  if (remove_2_disagree) {
    groups_with_issues <- subset_score_data %>%
      group_by(Basename, Row, Col, Pos) %>%
      filter(n() == 2, n_distinct(Score) > 1) %>%
      ungroup() %>%
      select(Basename, Row, Col, Pos)

    subset_score_data <- subset_score_data %>%
      anti_join(groups_with_issues, by = c("Basename", "Row", "Col", "Pos"))
  }
  subset_score_data
}

unique_scorers <- c("Original", as.character(sort(as.numeric(unique(score_data_long$Scorer)))))

results <- lapply(unique_scorers, function(scorer) {
  subset_T <- remove_scorer(score_data_long, scorer, TRUE)
  n_cdas_sub <- subset_T %>%
    distinct(Basename, Row, Col, Pos) %>%
    nrow()
  avg_T <- average_pairwise_agreement(pairwise_score_data(subset_T))
  data.frame(
    Rater_Removed = scorer, Score_Count = n_cdas_sub,
    Avg_Pairwise_Agreement = avg_T
  )
})

result_df <- bind_rows(results) %>%
  mutate(
    Images_Lost = n_cdas - Score_Count,
    Change_in_Agreement = Avg_Pairwise_Agreement - average_agreement,
    Agreement_Change_Per_Img_Lost = Change_in_Agreement / Images_Lost,
    Change_Per_Img_x1000 = round(Agreement_Change_Per_Img_Lost * 1000, 2)
  )

result_df %>%
  select(
    Rater_Removed, Score_Count, Avg_Pairwise_Agreement,
    Images_Lost, Agreement_Change_Per_Img_Lost, Change_Per_Img_x1000
  )
```

Why does pairwise agreement increase for most rater removals?

When a rater is removed, CDAs that now have only 2 remaining scores are also removed — but only if those 2 scores disagree. Since disagreement occurs ~36% of the time, each rater removal also removes a further 36% of the CDAs that rater uniquely shared, biasing the remaining pool toward agreement.

---

## 5. Verification: Random Sampling Gives the Same Average Agreement

```{r}
# 5
random_sample_pair_agree <- function(data, sample_num) {
  set.seed(42)
  CDA_set <- distinct(data, Basename, Row, Col, Pos)
  CDA_set <- CDA_set[sample(nrow(CDA_set)), ]
  sample_num <- min(sample_num, nrow(CDA_set))

  agreement_vector <- vapply(seq_len(sample_num), function(i) {
    sub <- data %>%
      filter(
        Basename == CDA_set$Basename[i], Row == CDA_set$Row[i],
        Col == CDA_set$Col[i], Pos == CDA_set$Pos[i]
      )
    if (nrow(sub) == 3) sub <- sub[-sample(1:3, 1), ]
    as.numeric(nrow(sub) == 2 && sub$Score[1] == sub$Score[2])
  }, numeric(1))

  mean(agreement_vector)
}

agreement_1000 <- random_sample_pair_agree(score_data_long, 1000)
cat("Original (weighted):", round(average_agreement, 2), "%\n")
cat("Random sample of 1,000 pairs:", round(agreement_1000 * 100, 2), "%\n")
```

---

## 6. Agreement Patterns Between Rater Pairs

```{r fig.width=22, fig.height=10}
#| fig-width: 22
#| fig-height: 10
# 6
gen_pairwise_counts <- function(score_data_long) {
  data <- score_data_long %>%
    mutate(Score = as.numeric(Score)) %>%
    group_by(Basename, Row, Col, Pos) %>%
    mutate(id = row_number()) %>%
    ungroup()

  pairwise <- data %>%
    full_join(data, by = c("Basename", "Row", "Col", "Pos"), relationship = "many-to-many") %>%
    filter(id.x < id.y) %>%
    select(Scorer.x, Score.x, Scorer.y, Score.y) %>%
    group_by(Scorer.x, Score.x, Scorer.y, Score.y) %>%
    summarise(Count = n(), .groups = "drop") %>%
    filter(!is.na(Score.x) & !is.na(Score.y))

  sym_copy <- pairwise
  names(sym_copy)[names(sym_copy) %in% c("Scorer.x", "Scorer.y")] <- c("Scorer.y", "Scorer.x")
  names(sym_copy)[names(sym_copy) %in% c("Score.x", "Score.y")] <- c("Score.y", "Score.x")
  rbind(pairwise, sym_copy) %>%
    mutate(Count_Binned = cut(Count,
      breaks = c(0, 1, 10, 100, 250, Inf),
      labels = c("1", "2-10", "11-100", "101-250", "251-1000"),
      include.lowest = TRUE
    ))
}

make_patterns_plot <- function(sdl) {
  counts <- gen_pairwise_counts(sdl)
  levels_xy <- as.character(1:10)
  counts <- counts %>%
    mutate(
      Scorer.x = factor(Scorer.x, levels = levels_xy),
      Scorer.y = factor(Scorer.y, levels = levels_xy)
    )
  complete_counts <- complete(counts,
    Scorer.x = factor(levels_xy, levels = levels_xy),
    Scorer.y = factor(levels_xy, levels = levels_xy),
    fill = list(Count = 0)
  ) %>%
    mutate(Scorer.y = fct_rev(Scorer.y))
  real_counts <- complete_counts %>% filter(Count > 0)

  ggplot(complete_counts, aes(x = Score.x, y = Score.y)) +
    geom_tile(aes(fill = Count_Binned), color = "black", linewidth = 0.5) +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey40") +
    scale_fill_manual(
      values = c(
        "1" = "#eff3ff",
        "2-10" = "#9ecae1",
        "11-100" = "#3182bd",
        "101-250" = "#fdae61",
        "251-1000" = "#d73027"
      ),
      name = "Frequency", na.value = "white", drop = FALSE
    ) +
    facet_grid(Scorer.y ~ Scorer.x, switch = "both") +
    theme_bw() +
    labs(x = "Rater X", y = "Rater Y") +
    theme(
      panel.spacing = unit(0.25, "lines"),
      strip.placement = "outside",
      panel.grid = element_blank(),
      strip.text = element_text(size = 11),
      axis.text = element_text(size = 10),
      axis.title = element_text(size = 13),
      legend.title = element_text(size = 13),
      legend.text = element_text(size = 11)
    )
}

p_patterns <- make_patterns_plot(score_data_long)
p_patterns
save_plot(p_patterns, "agreement_patterns.png", width = 11, height = 10)
```

**What this shows:**

- Rater 5 systematically scores higher than others across all facets.
- The Rater 8 vs Rater 10 facet shows a lighter 0–0 cell, consistent with their lower pairwise agreement.
- Removing 0s makes individual rater biases more visible.

---

## 7. Per-Rater Agreement with Median

Stacked bar chart showing each rater's exact agreement and within-±1 agreement with the CDA median score, computed across all CDAs scored by that individual. Adapted from `AutoCDAScorer_Models/analyses/0014_Scoring_Consistency/scoring_consistency.qmd` (human-only, no model results).

```{r fig.width=8, fig.height=6}
#| fig-width: 8
#| fig-height: 6
# 7
scorer_accuracy <- score_data_long %>%
  filter(!is.na(Score), Score != "[]") %>%
  mutate(Score = as.integer(Score)) %>%
  group_by(Scorer) %>%
  summarise(
    exact = mean(Score == Median_Score, na.rm = TRUE),
    within_one = mean(abs(Score - Median_Score) <= 1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(extra_within_one = within_one - exact)

avg_row <- scorer_accuracy %>%
  summarise(
    Scorer = factor("Avg"),
    exact = mean(exact),
    within_one = mean(within_one),
    extra_within_one = mean(extra_within_one)
  )

plot_acc <- bind_rows(scorer_accuracy, avg_row) %>%
  mutate(Scorer = factor(Scorer,
    levels = c(as.character(sort(as.numeric(
      unique(as.character(scorer_accuracy$Scorer))
    ))), "Avg")
  )) %>%
  pivot_longer(
    cols = c(exact, extra_within_one),
    names_to = "metric", values_to = "value"
  ) %>%
  mutate(metric = factor(metric,
    levels = c("extra_within_one", "exact"),
    labels = c("Additional within ±1", "Exact agreement")
  ))

p <- ggplot(plot_acc, aes(x = Scorer, y = value, fill = metric)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    values = c(
      "Exact agreement" = "steelblue",
      "Additional within ±1" = "cornflowerblue"
    ),
    name = NULL,
    breaks = c("Exact agreement", "Additional within ±1")
  ) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(x = "Rater", y = "Agreement with Median Severity Score") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 15),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 11)
  )
p
save_plot(p, "scorer_accuracy.png", width = 8, height = 6)
```

**What this shows:**

- Rater 5 stands out with substantially lower exact agreement and within-±1 agreement with the median, consistent with the systematic upward bias seen in other analyses.
- The majority of raters achieve >70% exact agreement and >95% within-±1 agreement.
- The "Avg" bar summarises the overall human performance baseline comparable to model evaluation metrics.

---

## 8. Agreement breakdown by median score

```{r}
# 8
# One row per CDA. Missing scores are stored as "[]"; treat them as NA.
# Agreement is defined by how many of the three scorers match the median score:
#   3 -> all three agreed, 2 -> two of three agreed, 1 -> all three differed.
conf_cda <- raw_data %>%
  select(Basename, Row, Col, Pos, Score1, Score2, Score3, Median_Score) %>%
  mutate(across(c(Score1, Score2, Score3), ~ na_if(as.character(.), "[]")),
    med = as.character(as.integer(Median_Score)),
    n_agree_median = (!is.na(Score1) & Score1 == med) +
      (!is.na(Score2) & Score2 == med) +
      (!is.na(Score3) & Score3 == med),
    Agreement = factor(
      case_when(
        n_agree_median == 3 ~ "All 3 agreed",
        n_agree_median == 2 ~ "2 of 3 agreed",
        TRUE ~ "All 3 differed"
      ),
      levels = c("All 3 agreed", "2 of 3 agreed", "All 3 differed")
    )
  )

confidence_table <- conf_cda %>%
  group_by(`Median Score` = as.integer(Median_Score)) %>%
  summarise(
    Frequency = n(),
    `All-3-Agree` = sum(Agreement == "All 3 agreed"),
    `All-3-Agree %` = round(mean(Agreement == "All 3 agreed") * 100, 1),
    `2-of-3-Agree` = sum(Agreement == "2 of 3 agreed"),
    `2-of-3-Agree %` = round(mean(Agreement == "2 of 3 agreed") * 100, 1),
    `All-3-Disagree` = sum(Agreement == "All 3 differed"),
    `All-3-Disagree %` = round(mean(Agreement == "All 3 differed") * 100, 1),
    .groups = "drop"
  )

knitr::kable(confidence_table, align = "c")
```

Each CDA is categorised by how many of its three raters agreed with the median: *All-3-Agree* (unanimous), *2-of-3-Agree* (a majority match the median, one differs), and *All-3-Disagree* (all three scores distinct). Percentages are out of the total number of CDAs at each median score. Note that 148 CDAs were scored by only two raters; in every case the two present scores match the median, so they fall in the *2-of-3-Agree* group.

---

## 9. Stacked bar of agreement by median score

```{r fig.width=8, fig.height=6}
#| fig-width: 8
#| fig-height: 6
# 9
conf_counts <- conf_cda %>%
  count(`Median Score` = as.integer(Median_Score), Agreement)

agree_pal <- c(
  "All 3 agreed" = "#1a9850",
  "2 of 3 agreed" = "#fdae61",
  "All 3 differed" = "#d73027"
)

p <- ggplot(conf_counts, aes(x = factor(`Median Score`), y = n, fill = Agreement)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    values = agree_pal, name = "Rater Agreement",
    guide = guide_legend(reverse = TRUE)
  ) +
  labs(x = "Median Severity Score", y = "Frequency") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 15),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 11)
  )
p
save_plot(p, "stackedbar.png", width = 8, height = 6)
```

**What this shows:**

- Median-0 CDAs are overwhelmingly unanimous, reflecting the dominance of negative controls and their visual clarity.
- Median-6 is the next most unanimous; both extremes are easier to identify consistently.
- Intermediate scores (1–5) are dominated by partial or full disagreement, highlighting the difficulty of distinguishing adjacent severity levels.
- All-three-differed is only possible for intermediate medians, since an extreme median (0 or 6) requires at least two raters to have given that extreme value.

---

## 10. Temporal test 2 — original scoring (image-ordered)

The original scoring was done in image order, with thousands of scores per rater — far more power, but
consecutive CDAs usually share an image. Each score is compared to the **leave-one-out consensus** (the
median, i.e. mean, of the *other two* raters on that CDA), giving a residual `score − consensus`. Scores
are matched to the canonical panel by image and CDA position; the leave-one-out reference removes the
self-inclusion that would otherwise force a score to match its own median.

```{r}
# 10
# Wald-Wolfowitz runs test on the sign of a series (p-value only; zeros dropped).
runs_p <- function(x) {
  s <- sign(x[x != 0])
  n <- length(s)
  n1 <- sum(s > 0)
  n2 <- sum(s < 0)
  if (n1 == 0 || n2 == 0 || n < 3) {
    return(NA_real_)
  }
  runs <- 1 + sum(s[-1] != s[-n])
  mu <- 1 + 2 * n1 * n2 / n
  v <- 2 * n1 * n2 * (2 * n1 * n2 - n) / (n^2 * (n - 1))
  2 * pnorm(-abs((runs - mu) / sqrt(v)))
}

# Canonical three-rater panel (one row per CDA).
canon <- read_csv(
  here("analyses", "01_generate_raw_data", "04_median_and_centre_coords", "combined_cda_data_median.csv"),
  show_col_types = FALSE
) %>%
  select(Basename, Row, Col, Pos, Scorer1, Score1, Scorer2, Score2, Scorer3, Score3, Median_Score) %>%
  mutate(
    across(c(Scorer1, Scorer2, Scorer3), as.integer),
    across(c(Score1, Score2, Score3, Median_Score), as.numeric)
  )

# data/cdascorer_anon/ holds one file per rater, named Scorer<ID>.csv - see
# the anonymisation script under scripts/.
read_raw_set <- function(set) {
  files <- list.files(here("data", "cdascorer_anon", set), pattern = "[.]csv$", full.names = TRUE)
  bind_rows(lapply(files, function(f) {
    scorer_id <- as.integer(sub("^Scorer", "", tools::file_path_sans_ext(basename(f))))
    read_csv(f, show_col_types = FALSE) %>%
      transmute(
        Set = set, Scorer = scorer_id, seq = row_number(),
        Basename = basename(img), Row = row, Col = col, Pos = pos, score = as.numeric(score)
      )
  }))
}

raw_resid <- bind_rows(read_raw_set("set1"), read_raw_set("set2")) %>%
  inner_join(canon %>% select(Basename, Row, Col, Pos, Score1, Score2, Score3),
    by = c("Basename", "Row", "Col", "Pos")
  ) %>%
  filter(!is.na(Score1), !is.na(Score2), !is.na(Score3)) %>%
  # keep only rows where the raw score is genuinely one of the three panel scores,
  # so (sum - score)/2 is exactly the median of the other two.
  filter(abs(score - Score1) < 1e-9 | abs(score - Score2) < 1e-9 | abs(score - Score3) < 1e-9) %>%
  mutate(
    loo_ref = (Score1 + Score2 + Score3 - score) / 2,
    resid = score - loo_ref
  )

cat(
  "Matched raw scores:", nrow(raw_resid), "across",
  n_distinct(paste(raw_resid$Scorer, raw_resid$Set)), "rater-sessions\n"
)

session_stats <- raw_resid %>%
  group_by(Scorer, Set) %>%
  arrange(seq, .by_group = TRUE) %>%
  summarise(
    n = n(), lag1_acf = cor(head(resid, -1), tail(resid, -1)),
    runs_p = runs_p(resid), .groups = "drop"
  )
print(as.data.frame(session_stats))
```

### 10.1 Within-image vs across-image autocorrelation

If the autocorrelation is genuine temporal momentum it should persist when consecutive scores come from
*different* images; if it is image-blocking it should vanish at image boundaries.

```{r}
# 10.1.1
raw_lag <- raw_resid %>%
  group_by(Scorer, Set) %>%
  arrange(seq, .by_group = TRUE) %>%
  mutate(lag_resid = lag(resid), lag_base = lag(Basename)) %>%
  ungroup() %>%
  filter(!is.na(lag_resid)) %>%
  mutate(boundary = if_else(lag_base == Basename, "Same image", "Different image"))

cat(
  "Consecutive pairs from the same image:",
  percent(mean(raw_lag$boundary == "Same image"), accuracy = 0.1), "\n\n"
)

boundary_fit <- bind_rows(lapply(c("Same image", "Different image"), function(g) {
  sub <- raw_lag %>% filter(boundary == g)
  f <- lm(resid ~ lag_resid + interaction(Scorer, Set), data = sub)
  ci <- confint(f, "lag_resid")
  tibble(
    boundary = g, coef = unname(coef(f)["lag_resid"]),
    lo = ci[1], hi = ci[2],
    p = summary(f)$coefficients["lag_resid", 4], n = nrow(sub)
  )
}))
boundary_fit$boundary <- factor(boundary_fit$boundary, levels = c("Same image", "Different image"))
print(boundary_fit)
```

```{r fig.width=6, fig.height=4.5}
#| fig-width: 6
#| fig-height: 4.5
# 10.1.2
p <- ggplot(boundary_fit, aes(boundary, coef)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(ymin = lo, ymax = hi), color = "steelblue", linewidth = 0.9, size = 0.9) +
  labs(x = NULL, y = "Lag-1 residual autocorrelation") +
  theme_minimal() +
  theme(axis.text = element_text(size = 13), axis.title = element_text(size = 14))
p
save_plot(p, "raw_image_boundary_autocorrelation.png", width = 6, height = 4.5)
```

**What this shows:**

- Pooled across the original scoring the residual is strongly positively autocorrelated — but this is **entirely a within-image effect**. ~97% of consecutive scores share an image; the lag-1 autocorrelation is large and highly significant for same-image pairs yet indistinguishable from zero across image boundaries (CI spans 0).
- This matches the rescore drift autocorrelation test in `06_intra_rater_reliability`: when the context changes — randomised rescore order, or crossing an image boundary — there is no residual momentum. The apparent clustering is image-blocking, not sequential anchoring.

---

## 11. Image-level structure of disagreement (confound-free)

The temporal tests show disagreement is organised at the image level, but the raw-score autocorrelation
conflates several things (the experimental layout, the abundance of agreed 0s, per-image bias). A cleaner
test, free of temporal order and of the self-vs-median issue, looks at the **difference between two
raters** on the same CDA: the true severity cancels, leaving only their relative disagreement. If that
difference is consistent within an image but shifts between images, disagreement has an image-level
component. This is the intraclass correlation (ICC) of the pairwise difference with image as the grouping
factor: the fraction of disagreement variance attributable to *which image* a CDA is on.

```{r}
# 11
# All within-CDA rater pairs, oriented lower-ID minus higher-ID, with image label.
mk_pair <- function(d, a, b) {
  d %>%
    transmute(Basename, Median_Score,
      sa = .data[[paste0("Scorer", a)]], sb = .data[[paste0("Scorer", b)]],
      ya = .data[[paste0("Score", a)]], yb = .data[[paste0("Score", b)]]
    ) %>%
    filter(!is.na(sa), !is.na(sb), !is.na(ya), !is.na(yb)) %>%
    mutate(
      pair = paste(pmin(sa, sb), pmax(sa, sb), sep = "-"),
      diff = if_else(sa < sb, ya - yb, yb - ya)
    )
}
pairs <- bind_rows(mk_pair(canon, 1, 2), mk_pair(canon, 1, 3), mk_pair(canon, 2, 3)) %>%
  select(pair, Basename, Median_Score, diff)

# Random-effects (one-way ANOVA) ICC of the difference by image, for one rater-pair.
icc_by_pair <- function(df) {
  g <- df %>%
    group_by(Basename) %>%
    summarise(ni = n(), mi = mean(diff), .groups = "drop") %>%
    filter(ni >= 2)
  df <- df %>% semi_join(g, by = "Basename")
  k <- nrow(g)
  N <- nrow(df)
  if (k < 3 || N < 10) {
    return(tibble(k = k, N = N, icc = NA_real_, btw_sd = NA, wth_sd = NA))
  }
  grand <- mean(df$diff)
  ss_b <- sum(g$ni * (g$mi - grand)^2)
  g_mi <- g %>% select(Basename, mi)
  ss_w <- df %>%
    left_join(g_mi, by = "Basename") %>%
    summarise(s = sum((diff - mi)^2)) %>%
    pull(s)
  ms_b <- ss_b / (k - 1)
  ms_w <- ss_w / (N - k)
  n0 <- (N - sum(g$ni^2) / N) / (k - 1)
  var_b <- max(0, (ms_b - ms_w) / n0)
  var_w <- ms_w
  tibble(k = k, N = N, icc = var_b / (var_b + var_w), btw_sd = sqrt(var_b), wth_sd = sqrt(var_w))
}

run_icc <- function(dat, label) {
  res <- dat %>%
    group_by(pair) %>%
    group_modify(~ icc_by_pair(.x)) %>%
    ungroup() %>%
    filter(!is.na(icc)) %>%
    mutate(subset = label)
  cat(sprintf(
    "[%s] %d pairs | median ICC = %.3f | within-image SD = %.2f | between-image SD = %.2f\n",
    label, nrow(res), median(res$icc), median(res$wth_sd), median(res$btw_sd)
  ))
  res
}
icc_all <- run_icc(pairs, "All CDAs (incl. 0s)")
icc_nz <- run_icc(pairs %>% filter(Median_Score > 0), "Non-zero median only")
icc_res <- bind_rows(icc_all, icc_nz) %>%
  mutate(subset = factor(subset, levels = c("All CDAs (incl. 0s)", "Non-zero median only")))
```

### 11.1 ICC of pairwise disagreement by image

```{r fig.width=7, fig.height=5}
#| fig-width: 7
#| fig-height: 5
# 11.1
p <- ggplot(icc_res, aes(subset, icc, fill = subset)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.5) +
  geom_jitter(width = 0.12, height = 0, size = 1.6, alpha = 0.7) +
  scale_fill_manual(values = c(
    "All CDAs (incl. 0s)" = "grey65",
    "Non-zero median only" = "#3182bd"
  ), guide = "none") +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = NULL, y = "ICC of pairwise difference by image\n(per rater-pair)") +
  theme_minimal() +
  theme(axis.text = element_text(size = 12), axis.title = element_text(size = 13))
p
save_plot(p, "pairwise_image_icc.png", width = 7, height = 5)
```

**What this shows:**

- Each point is one rater-pair; the ICC is the share of their disagreement variance explained by image.
- The median ICC is modest but clearly non-zero (≈0.14 with 0s, ≈0.22 without): two raters' relative offset has a real per-image component, while most disagreement remains CDA-specific (within-image SD ≫ between-image SD).
- Crucially, **removing the abundant agreed-0 CDAs raises the ICC** rather than lowering it — so the image-level structure is not an artefact of zero-heavy control regions; it is strongest among genuinely contested CDAs.

### 11.2 Illustration: per-image mean difference for the best-sampled pairs

```{r fig.width=12, fig.height=4.5}
#| fig-width: 12
#| fig-height: 4.5
# 11.2
top_pairs <- pairs %>%
  count(pair, Basename) %>%
  count(pair, name = "n_images") %>%
  slice_max(n_images, n = 3) %>%
  pull(pair)

img_means <- pairs %>%
  filter(pair %in% top_pairs) %>%
  group_by(pair, Basename) %>%
  summarise(
    ni = n(), mean_diff = mean(diff),
    se = ifelse(n() > 1, sd(diff) / sqrt(n()), NA), .groups = "drop"
  ) %>%
  filter(ni >= 5) %>%
  group_by(pair) %>%
  arrange(mean_diff, .by_group = TRUE) %>%
  mutate(img_rank = row_number()) %>%
  ungroup()

p <- ggplot(img_means, aes(img_rank, mean_diff)) +
  geom_hline(yintercept = 0, color = "grey60") +
  geom_pointrange(aes(ymin = mean_diff - 1.96 * se, ymax = mean_diff + 1.96 * se),
    color = "steelblue", size = 0.3
  ) +
  facet_wrap(~pair, scales = "free_x", labeller = label_both) +
  labs(
    x = "Image (ranked by mean difference)",
    y = "Mean score difference\n(rater A − rater B)"
  ) +
  theme_minimal() +
  theme(
    panel.spacing = unit(0.6, "lines"),
    axis.text = element_text(size = 11), axis.title = element_text(size = 13),
    strip.text = element_text(size = 11)
  )
p
save_plot(p, "pairwise_per_image_mean_diff.png", width = 12, height = 4.5)
```

**What this shows:**

- For the best-sampled rater-pairs, each point is the mean difference on one image with its 95% CI.
- The per-image means span a range (a between-image gradient) wider than the within-image error bars, illustrating the ICC: the two raters' relative offset shifts from image to image while being comparatively settled within each image.

---

## 12. Discussion

### 12.1 Pairwise % agreement between raters

_Add notes here._

### 12.2 Off-by-one disagreement rate

_Add notes here._

### 12.3 Shared CDA count per rater pair

_Add notes here._

### 12.4 Effect of removing each rater on average pairwise agreement

_Add notes here._

### 12.5 Verification: random sampling

_Add notes here._

### 12.6 Agreement patterns between rater pairs

_Add notes here._

### 12.7 Per-rater agreement with median

_Add notes here._

### 12.8 Agreement breakdown by median score

_Add notes here._

### 12.9 Stacked bar of agreement by median score

_Add notes here._

### 12.10 Temporal test 2 — original scoring (image-ordered)

_Add notes here._

### 12.11 ICC of pairwise disagreement by image

_Add notes here._

### 12.12 Illustration: per-image mean difference for the best-sampled pairs

_Add notes here._

### 12.13 Confounds and interpretation (image-level structure)

**Same image = same experiment.** Original scoring is not a random sequence. CDAs on one image are spots
on one leaf from one infiltration, so they share construct, leaf age, lighting and background. Two
effects follow: (i) the underlying severities are correlated, and (ii) a rater's *reading* of that leaf
(exposure, contrast, what counts as "necrotic" under that lighting) is correlated. The §10 within-image
autocorrelation and the §11 ICC both capture this image-level structure; neither is evidence of temporal
anchoring, which the randomised rescore (see `06_intra_rater_reliability`) and the across-image boundary
(§10.1) both rule out.

**The abundance of 0s and non-random layout.** Infiltration layouts group controls together, so a rater
genuinely does see runs of 0s — but those are near-universally agreed, so they contribute *agreement*,
not disagreement structure. This is why the analyses use **residuals / pairwise differences** rather than
raw scores (which would be trivially autocorrelated by the design), and why the §11 ICC is computed both
with and without 0s. The result is reassuring: excluding 0s *increases* the image-level ICC, so the
structure lives in the contested mid-range CDAs, not in the clustered controls.

**Is class balancing necessary?** For the *temporal* tests, no — and it would be harmful: down-sampling
or re-weighting score classes destroys the true adjacency of the sequence, so a "lag-1" on a balanced
series no longer means consecutive-in-time. The appropriate controls there are analysing residuals (not
raw scores) and the within/across-image split, both of which are done. For the *image-level* ICC, balance
is handled implicitly: the difference between two raters already removes the severity (class) signal, and
reporting the non-zero subset alongside the full set shows the conclusion is not driven by the dominant
0 class.

**Overall.** Over-/under-scoring does **not** have temporal momentum across contexts — a rater's last
score does not pull their next one once the image changes. What is real is a modest **image-level**
structure to disagreement: two raters' relative calibration drifts somewhat from image to image
(plausibly per-leaf lighting/background and per-session adaptation) while staying more consistent within
an image. Most disagreement, however, remains CDA-specific — the irreducible ambiguity of scoring an
individual cell-death area.

### 12.14 Overall narrative

_Add notes here._
