---
title: "Plots by source image / CDA image"
format:
  html:
    code-fold: true
---

Date: 08/06/2026

## GOHREP

**Goal:**

Visualise the image dataset in a way that allows all scores / images to be seen.

## 1. Setup

### 1.1 Load required packages

```{r}
#| output: false
# 1.1.1
library(readr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(here)
```

```{r}
# 1.1.2
source(here("src", "data_helpers.R"))
source(here("src", "plotting_helpers.R"))

save_plot <- make_save_plot("04_image_level_analysis")
```

### 1.2 Load data

```{r}
# 1.2
path_data <- here("analyses", "01_generate_raw_data", "04_median_and_centre_coords", "combined_cda_data_median.csv")
score_data <- read_csv(path_data, show_col_types = FALSE) %>%
  select(Basename, Row, Col, Pos, Scorer1, Score1, Scorer2, Score2, Scorer3, Score3, Median_Score) %>%
  mutate(
    across(c(Scorer1, Scorer2, Scorer3), as.character),
    across(c(Score1, Score2, Score3), ~ replace(as.character(.), . == "[]", NA))
  )
```

### 1.3 Create a long table with one rater and one score per row

```{r}
# 1.3
score_data_long <- long_score_table(score_data)
```

## 2. Dataset overview (Materials & Methods)

### 2.1 Load allocation metadata

```{r}
# 2.1.1
# image_info.csv has one row per source image. Path is recorded once per
# folder, so we fill it downwards; the top-level folder is the project,
# mapped to its anonymised label via project_anonymisation_key.csv.
project_key <- load_project_key()
project_lookup <- setNames(project_key$Anon_Label, project_key$Project)

image_info <- read_csv(
  here("analyses", "01_generate_raw_data", "01_allocation_info", "image_info.csv"),
  show_col_types = FALSE
) %>%
  fill(Path, .direction = "down") %>%
  mutate(Project = unname(project_lookup[sub("/.*", "", Path)]))
```

```{r}
# 2.1.2
# The allocation metadata over-counts CDAs-per-leaf for 16 source images:
# inspecting them confirmed these leaves carry only 5 CDAs, not the recorded
# 6 or 8. Corrected here in-memory only; image_info.csv (which drives the
# allocation pipeline) is left unchanged. Original value kept as CDA_Per_Leaf_Alloc.
layout5_images <- c(
  # Project E batch: allocation recorded 8 CDAs/leaf, true layout is 5
  "_DSC9258.tif", "_DSC9259.tif", "_DSC9260.tif", "_DSC9261.tif",
  # Project B batch: allocation recorded 6 CDAs/leaf, true layout is 5
  "DSC_2394.tif", "DSC_2395.tif", "DSC_2396.tif", "DSC_2397.tif",
  "DSC_2451.tif", "DSC_2452.tif", "DSC_2453.tif", "DSC_2454.tif",
  "DSC_2867.tif", "DSC_2868.tif", "DSC_2869.tif", "DSC_2870.tif"
)

image_info <- image_info %>%
  mutate(
    CDA_Per_Leaf_Alloc = CDA_Per_Leaf,
    CDA_Per_Leaf = if_else(Image %in% layout5_images, 5, CDA_Per_Leaf)
  )
```

### 2.2 Total number of source images

```{r}
# 2.2
n_from_files <- length(list.files(here("data", "raw_images_rotated"),
  pattern = "(?i)\\.tif$"
))

tibble(Source_Images = n_from_files) %>%
  knitr::kable(caption = "Total number of source images")

image_info %>%
  count(Set, name = "Source_Images") %>%
  mutate(Percent = round(100 * Source_Images / sum(Source_Images), 1)) %>%
  knitr::kable(caption = "Source images per set")
```

### 2.3 Number of raters per set

```{r}
# 2.3
raters_long <- score_data %>%
  select(Basename, Scorer1, Scorer2, Scorer3) %>%
  left_join(select(image_info, Image, Set), by = c("Basename" = "Image")) %>%
  pivot_longer(c(Scorer1, Scorer2, Scorer3), values_to = "Rater") %>%
  filter(!is.na(Rater), Rater != "[]")

raters_long %>%
  group_by(Set) %>%
  summarise(
    N_Raters = n_distinct(Rater),
    Raters = paste(sort(unique(as.numeric(Rater))), collapse = ", "),
    .groups = "drop"
  ) %>%
  knitr::kable(caption = "Distinct raters per set")
```

### 2.4 Distribution of source images across projects

```{r}
# 2.4.1
image_info %>%
  count(Set, Project, name = "Source_Images") %>%
  arrange(desc(Source_Images)) %>%
  mutate(Percent = round(100 * Source_Images / sum(Source_Images), 1)) %>%
  knitr::kable(caption = "Source images per project")
```

```{r}
# 2.4.2
score_data %>%
  left_join(select(image_info, Image, Set, Project), by = c("Basename" = "Image")) %>%
  count(Set, Project, name = "CDAs") %>%
  arrange(desc(CDAs)) %>%
  mutate(Percent = round(100 * CDAs / sum(CDAs), 1)) %>%
  knitr::kable(caption = "CDAs (cleaned dataset) per project")
```

### 2.5 Projects, layouts and counts (combined)

```{r}
# 2.5
image_info %>%
  count(Set, Project, CDA_Per_Leaf, name = "n") %>%
  arrange(Set, Project, CDA_Per_Leaf) %>%
  group_by(Set, Project) %>%
  summarise(
    Count            = sum(n),
    `CDAs Per Leaf`  = paste0(CDA_Per_Leaf, " (", n, ")", collapse = "<br>"),
    .groups          = "drop"
  ) %>%
  arrange(Set, desc(Count)) %>%
  knitr::kable(
    caption = "Set, project, source image count and leaf layout(s)",
    escape = FALSE
  )
```

### 2.6 Distribution of leaf layouts (cell death areas per leaf)

```{r}
# 2.6
# Layout derived independently from scored positions in combined_cda_data_median.csv:
# Pos indexes the CDA within a leaf, so an image's layout is the max Pos it
# contains (over all leaves, since individual leaves may have unscored positions).
layout_positions <- score_data %>%
  group_by(Basename) %>%
  summarise(CDA_Per_Leaf = max(Pos), .groups = "drop") %>%
  count(CDA_Per_Leaf, name = "Source_Images") %>%
  arrange(CDA_Per_Leaf)

layout_positions %>%
  knitr::kable(caption = "CDAs per leaf, derived from scored positions")
```

## 3. Score proportions per source image

### 3.1 Compute proportions

```{r}
# 3.1
score_data_grouped <- score_data %>%
  group_by(Basename) %>%
  summarize(
    Frequency_0 = sum(c(Score1, Score2, Score3) == 0, na.rm = TRUE),
    Frequency_1 = sum(c(Score1, Score2, Score3) == 1, na.rm = TRUE),
    Frequency_2 = sum(c(Score1, Score2, Score3) == 2, na.rm = TRUE),
    Frequency_3 = sum(c(Score1, Score2, Score3) == 3, na.rm = TRUE),
    Frequency_4 = sum(c(Score1, Score2, Score3) == 4, na.rm = TRUE),
    Frequency_5 = sum(c(Score1, Score2, Score3) == 5, na.rm = TRUE),
    Frequency_6 = sum(c(Score1, Score2, Score3) == 6, na.rm = TRUE),
    Total = n() * 3
  ) %>%
  mutate(
    Proportion_0 = Frequency_0 / Total,
    Proportion_1 = Frequency_1 / Total,
    Proportion_2 = Frequency_2 / Total,
    Proportion_3 = Frequency_3 / Total,
    Proportion_4 = Frequency_4 / Total,
    Proportion_5 = Frequency_5 / Total,
    Proportion_6 = Frequency_6 / Total
  ) %>%
  rowwise() %>%
  mutate(Most_Common_Score = which.max(c(
    Proportion_0, Proportion_1, Proportion_2, Proportion_3, Proportion_4, Proportion_5, Proportion_6
  )) - 1) %>%
  ungroup()

# Long form: one row per (image, score) with its proportion
score_data_heatmap <- score_data_grouped %>%
  pivot_longer(
    cols = starts_with("Proportion"),
    names_to = "Score",
    values_to = "Proportion"
  ) %>%
  mutate(Score = sub("Proportion_", "", Score))
```

### 3.2 Distribution of each score's proportion across images (violins)

```{r fig.width=10, fig.height=6}
#| fig-width: 10
#| fig-height: 6
# 3.2
score_box <- score_data_heatmap %>%
  mutate(Score = factor(Score, levels = as.character(0:6))) %>%
  group_by(Score) %>%
  mutate(is_outlier = Proportion > quantile(Proportion, 0.75) + 1.5 * IQR(Proportion)) %>%
  ungroup()

outlier_counts <- score_box %>%
  group_by(Score) %>%
  summarise(n_out = sum(is_outlier), y_lab = max(Proportion) + 0.04, .groups = "drop")

set.seed(1)
p <- ggplot(score_box, aes(x = Score, y = Proportion)) +
  geom_violin(aes(fill = Score), scale = "width", quantile.linetype = 1) +
  geom_point(
    data = subset(score_box, is_outlier),
    position = position_jitter(width = 0.08, height = 0), size = 0.8, alpha = 0.5
  ) +
  geom_text(
    data = outlier_counts, aes(x = Score, y = y_lab, label = n_out),
    size = 4, colour = "grey25"
  ) +
  scale_fill_manual(values = sev_pal, guide = "none") +
  labs(
    x = "Severity Score", y = "Proportion per Image",
    caption = "Number above each violin = images in the upper tail (> Q3 + 1.5 x IQR)"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 15),
    plot.caption = element_text(size = 10, colour = "grey35")
  )
p
save_plot(p, "score_proportion_violins.png", width = 10, height = 6)
```

### 3.3 Heatmap of proportions sorted by most common score

```{r fig.width=8, fig.height=16}
#| fig-width: 8
#| fig-height: 16
# 3.3
score_data_sorted <- score_data_grouped %>%
  arrange(Most_Common_Score, Basename)
breakpoints <- which(diff(score_data_sorted$Most_Common_Score) != 0) + 0.5

score_data_heatmap3 <- score_data_grouped %>%
  pivot_longer(
    cols = starts_with("Proportion"),
    names_to = "Score",
    values_to = "Proportion"
  ) %>%
  mutate(Score = sub("Proportion_", "", Score)) %>%
  select(Basename, Score, Proportion, Most_Common_Score)

p <- ggplot(score_data_heatmap3, aes(x = Score, y = reorder(Basename, Most_Common_Score), fill = Proportion)) +
  geom_tile() +
  geom_hline(yintercept = breakpoints, linetype = "dashed", color = "red", linewidth = 1) +
  scale_fill_gradient(low = "white", high = "firebrick") +
  labs(x = "Severity Score", y = "Agroinfiltration Image", fill = "Proportion") +
  theme_minimal() +
  theme(
    axis.text.x  = element_text(size = 13, angle = 45, hjust = 1),
    axis.text.y  = element_text(size = 4),
    axis.title   = element_text(size = 15),
    legend.title = element_text(size = 13),
    legend.text  = element_text(size = 11)
  )
p
save_plot(p, "heatmap_proportion_sorted_by_most_common_score.png", 8, 16)
```

## 4. Interpretation and Discussion

- Some images had no 0s. This is interesting because there are normally control infiltrations.
- Some images had almost all 0s.
- Non-0 scores rarely comprised more than 50% of cell death areas in a source image, with a couple of excepions for 2s and 6s.
- The images with no 0s are mostly images where two infiltrations were applied to the same leaf. In the case that there was none of the usual green cell death, there was an orange cell death of the other infiltration, and the annotators were instructed to ignore those cell death areas because they did not represent true "no cell death" examples.
- Like the CDA images themselves, in many of the source agroinfiltration images the most common score was 0.
- For images where the most common score was 5 or 6, there tended to be very few scores between 1 and 3.
- For images where the most common score was 3, there tended to be very few extreme scores (0 or 6).
- The images for which most scores were extreme (0 or 5/6) were likely from biological systems that showed strong presence/absence phenotypes.
- The images with very few 0s were likely from images where each cell death area received two infiltrations, and those with no green cell death were excluded from annotation.

_Add notes here._

## 5. Additional Information

### 5.1 Leaf layout distribution (from the allocation record)

```{r}
# 5.1
layout_alloc <- image_info %>%
  count(CDA_Per_Leaf, name = "Source_Images") %>%
  arrange(CDA_Per_Leaf)

layout_alloc %>%
  knitr::kable(caption = "CDAs per leaf, from the (corrected) allocation record")
```

### 5.2 Manually corrected layouts (allocated vs verified)

```{r}
# 5.2
image_info %>%
  filter(Image %in% layout5_images) %>%
  count(Project, CDA_Per_Leaf_Alloc, CDA_Per_Leaf, name = "Source_Images") %>%
  arrange(desc(Source_Images)) %>%
  knitr::kable(caption = "Manually corrected layouts (allocated vs verified)")
```
