---
title: "Plots by raw image / CDA image"
format:
  html:
    code-fold: true
---
 
Date: 17/5/2024

## GOHREP
**Goal:**

Visualise the image dataset in a way that allows all scores / images to be seen

**Hypothesis:**

NA

**Rationale:**

NA

**Experimental plan:**

1.

### Load required packages.

```{r}
library(readr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(here)
library(RColorBrewer)
```

### Load in the data. Create combined table. Anonymise scorers.

```{r}
path1 <- here("analyses", "05_by_raw_image", "output_combined.csv")
path2 <- here("analyses", "05_by_raw_image", "median_combined.csv")
path3 <- here("analyses", "05_by_raw_image", "scorer_key.csv")

output_data <- read_csv(path1, show_col_types = FALSE)
median_data <- read_csv(path2, show_col_types = FALSE)

scorer_key <- read_csv(path3, show_col_types = FALSE, col_types = cols(ID = col_character(), Scorer_Name = col_character()))

score_data <- output_data %>%
  select(Basename, Row, Col, Pos, Scorer1, Score1, Scorer2, Score2, Scorer3, Score3) %>%
  bind_cols(
      median_data %>%
        select(Median_Score)
  )

id_lookup <- setNames(scorer_key$ID, scorer_key$Scorer_Name)

score_data <- score_data %>%
  mutate(across(c(Scorer1, Scorer2, Scorer3), ~ id_lookup[.]))

score_data$Score1[score_data$Score1 == "[]"] <- NA
score_data$Score2[score_data$Score2 == "[]"] <- NA
score_data$Score3[score_data$Score3 == "[]"] <- NA

score_data$row_index <- seq_len(nrow(score_data))
```

Count by median score
```{r}
print(score_data)

median_score_counts <- score_data %>%
  group_by(Median_Score) %>%
  summarise(count = n())
print(median_score_counts)

```


### Create a long table with one scorer and one score in each row.

```{r}
long_score_table <- function(score_data) {
    scorer_data <- score_data %>%
    select(starts_with("Scorer")) %>%
    pivot_longer(cols = everything(), names_to = NULL, values_to = "Scorer")

    score_values <- score_data %>%
    select(starts_with("Score"), -starts_with("Scorer")) %>%
    pivot_longer(cols = everything(), names_to = NULL, values_to = "Score")

    additional_data <- score_data %>%
    select(Basename, Row, Col, Pos, Median_Score) %>%
    slice(rep(row_number(), each = 3))

    score_data_long <- bind_cols(scorer_data, score_values, additional_data)

    score_data_long$Scorer <- factor(score_data_long$Scorer, levels = sort(as.numeric(unique(score_data_long$Scorer))))

    score_data_long <- score_data_long %>%
    filter(Score != "[]")

    return(score_data_long)
}
score_data_long <- long_score_table(score_data)
```


### Pixel plots

Add Total, Range, R, G and B, Min, Med, Max, X, Y columns

```{r}
score_data_filtered <- score_data %>%
  filter(!is.na(as.numeric(Score1)),
         !is.na(as.numeric(Score2)),
         !is.na(as.numeric(Score3)))

# Ensure conversion to numeric is successful and there are no NAs
score_data_processed <- score_data_filtered %>%
  mutate(
    Score1 = as.numeric(Score1),
    Score2 = as.numeric(Score2),
    Score3 = as.numeric(Score3),
    Total = Score1 + Score2 + Score3,
    Min = pmin(Score1, Score2, Score3, na.rm = TRUE),
    Max = pmax(Score1, Score2, Score3, na.rm = TRUE),
    Range = Max - Min
  )

x_coords <- rep(1:79, times = 79)[1:nrow(score_data_processed)]
y_coords <- rep(79:1, each = 79)[1:nrow(score_data_processed)]

score_data_with_coords <- score_data_processed %>%
  mutate(X = x_coords, Y = y_coords)

score_data_with_coords <- mutate(score_data_with_coords,
  Red_scaled = Score1 * (255/6),
  Green_scaled = Score2 * (255/6),
  Blue_scaled = Score3 * (255/6)
)

score_data_with_coords <- score_data_with_coords %>%
  mutate(
    RGB = rgb(Red_scaled, Green_scaled, Blue_scaled, maxColorValue = 255)
  )

score_data_with_coords$Max_scaled <- score_data_with_coords$Max * (256/7)
score_data_with_coords$Median_scaled <- score_data_with_coords$Median_Score * (256/7)
score_data_with_coords$Min_scaled <- score_data_with_coords$Min * (256/7)

score_data_with_coords$RGB2 <- rgb(score_data_with_coords$Max_scaled, 
                                  score_data_with_coords$Median_scaled, 
                                  score_data_with_coords$Min_scaled, 
                                  maxColorValue = 256)
```

#### By median score

```{r fig.width=11, fig.height=10}
#| fig-width: 11
#| fig-height: 10
custom_theme <- theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank()
  )

ggplot(score_data_with_coords, aes(x = X, y = Y, fill = Median_Score)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(x = "X Coordinate", y = "Y Coordinate", fill = "Median Score") +
  custom_theme
```

What does this plot show?

- This plot clearly represents the consistent infiltration patterns. Going from left to right on a line, there is often a pattern of scores that cycles within an agroinfiltration image across leaves.
- The patterns clearly differ between agroinfiltration images, for example the number of zeros in leaf.
- There are some areas where this pattern is less clear, where there are long stretches with all or no zeros.

Hypothesis forming:

1. N/A

#### By range

```{r fig.width=11, fig.height=10}
#| fig-width: 11
#| fig-height: 10
ggplot(score_data_with_coords, aes(x = X, y = Y, fill = Range)) +
  geom_tile() +
  scale_fill_viridis_c() +  # You can choose any color scale you prefer
  labs(x = "X Coordinate", y = "Y Coordinate", fill = "Range of Scores") +
  custom_theme
```

What does this plot show?

- For the majority of scores, the range of scores is 0 or 1.
- For a couple of CDAs, the range is 3 or higher.
- There appear to be two clusters towards the top of the image with a higher proportion of CDAs with a range of 2. Since the order of CDAs is like lines in a book (left to right in lines), these are likely to come from similar images.

Hypothesis forming:

1. Some raw agroinfiltration images were scored less consistently than others
2. Some CDA images had largely disagreeing scores. This could be due to error.

#### RGB as Scorer1, Scorer2, Scorer3

```{r fig.width=10, fig.height=10}
#| fig-width: 10
#| fig-height: 10
ggplot(score_data_with_coords, aes(x = X, y = Y, fill = RGB)) +
  geom_tile() +
  labs(x = "X Coordinate", y = "Y Coordinate") +
  scale_fill_identity() +
  guides(fill = "none") +
  custom_theme
```

What does this show?

- This plot is challenging to interpret
- R is Scorer1, G is Scorer2, and B is Scorer 3, so pixels which look stronger for one of those colours indicate a higher score by that scorer.
- I don't think has a meaningful interpretation, since Scorer1, Scorer2, and Scorer3 differ between images.

Hypothesis forming:

1. N/A

#### RGB as Max, Median, Min

```{r fig.width=10, fig.height=10}
#| fig-width: 10
#| fig-height: 10
ggplot(score_data_with_coords, aes(x = X, y = Y, fill = RGB2)) +
  geom_tile() +
  labs(x = "X Coordinate", y = "Y Coordinate") +
  scale_fill_identity() +
  guides(fill = "none") +
  custom_theme
```

What does this show?

- In this plot, R G and B have been ordered, so that R is always the highest score.
- Pixels that are greyscale have agreeing scores, and the more red a pixel is, the more its maximum score differs from the other two.
- The more brown a pixel is, the more its minimum score differs from the other two (mixing green and red).
- This is hard to interpret and the colour scheme isn't colourblind friendly.

Hypothesis forming:

1. N/A

### Dotplot of scores

```{r fig.width=14, fig.height=8}
#| fig-width: 14
#| fig-height: 8
ggplot(score_data, aes(x = row_index, y = Score)) +
  geom_point(aes(y = as.numeric(Score1)), alpha=0.1) +
  geom_point(aes(y = as.numeric(Score2)), alpha=0.1) +
  geom_point(aes(y = as.numeric(Score3)), alpha=0.1) +
  scale_y_continuous(limits = c(0, 6)) +
  labs(x = "CDA Index", y = "Score") +
  theme_minimal()
```

```{r}

print(sum(!is.na(score_data$Score1)) + sum(!is.na(score_data$Score2)) + sum(!is.na(score_data$Score3)))
```

What does this plot show?

- When the points are plotted in order of their row number, it's hard to see any trends in the data due to the large number of datapoints.

Hypothesis forming:

1. N/A

#### Sorted by Median Score
```{r fig.width=14, fig.height=8}
#| fig-width: 14
#| fig-height: 8
score_data_sorted <- score_data %>%
  arrange(Median_Score) %>%
  mutate(row_index = row_number())

boundary_df <- score_data_sorted %>%
  group_by(Median_Score) %>%
  summarize(boundary = max(row_index)) %>%
  mutate(boundary = boundary - 0.5)

ggplot(score_data_sorted, aes(x = row_index, y = Score)) +
  geom_point(aes(y = as.numeric(Score1)), alpha=0.1) +
  geom_point(aes(y = as.numeric(Score2)), alpha=0.1) +
  geom_point(aes(y = as.numeric(Score3)), alpha=0.1) +
  geom_vline(data = boundary_df, aes(xintercept = boundary), color = "red", linetype = "dashed") +
  scale_y_continuous(limits = c(0, 6)) +
  labs(x = "CDA Index", y = "Score") +
  theme_minimal()
```

What does this plot show?

- The difference in class sizes can be seen clearly here.
- Even with an alpha of 0.1, it is challenging to see trends.

Hypothesis forming:

1. N/A

#### Faceted by Median Score
```{r fig.width=10, fig.height=14}
#| fig-width: 10
#| fig-height: 14


set.seed(123)
score_data_random <- score_data[sample(nrow(score_data)), ]
score_data_random$row_index <- seq_len(nrow(score_data_random))

desired_order <- c("6", "5", "4", "3", "2", "1", "0")
score_data_random$Median_Score <- factor(score_data_random$Median_Score, levels = desired_order)


allscorespath <- here("analyses", "05_by_raw_image", "all_scores.png")

png(allscorespath, width = 1500, height = 1000)

ggplot(score_data_random, aes(x = row_index, y = as.numeric(Score))) +
  geom_point(aes(y = as.numeric(Score1)), color = "black", alpha = 0.2) +
  geom_point(aes(y = as.numeric(Score2)), color = "black", alpha = 0.2) +
  geom_point(aes(y = as.numeric(Score3)), color = "black", alpha = 0.2) +
  scale_y_continuous(limits = c(0, 6)) +
  labs(x = "CDA Index", y = "Score") +
  theme_minimal() +
  facet_wrap(~ Median_Score, ncol = 2) +
  theme(
    text = element_text(size = 25),
    strip.text = element_text(size = 25)
  )

dev.off()
```

What does this plot show?

- When the datapoints are more spread out like this, the variability of scores can be seen more clearly.
- For CDA images of median 0, disagreements are less common, and are typically 1s, although there are a couple of scores between 3-6 which could potentially be errors.
- For intermediate scores, there is more variation of scores collected within a median score class. Many images have scores 1 above or below the median, and there are also a reasonable amount 2 above or below the median. Three above or below is very rare.

Hypothesis forming:

1. Some CDA images had largely disagreeing scores. This could be due to error.

### Heatmap of scores per raw_img (original agroinfiltration image)

#### Frequency
```{r fig.width=8, fig.height=16}
#| fig-width: 8
#| fig-height: 16
score_data_grouped <- score_data %>%
  group_by(Basename) %>%
  summarize(
    Frequency_0 = sum(c(Score1, Score2, Score3) == 0, na.rm = TRUE),
    Frequency_1 = sum(c(Score1, Score2, Score3) == 1, na.rm = TRUE),
    Frequency_2 = sum(c(Score1, Score2, Score3) == 2, na.rm = TRUE),
    Frequency_3 = sum(c(Score1, Score2, Score3) == 3, na.rm = TRUE),
    Frequency_4 = sum(c(Score1, Score2, Score3) == 4, na.rm = TRUE),
    Frequency_5 = sum(c(Score1, Score2, Score3) == 5, na.rm = TRUE),
    Frequency_6 = sum(c(Score1, Score2, Score3) == 6, na.rm = TRUE)
  )

# Reshape the data for heatmap visualization
score_data_heatmap <- score_data_grouped %>%
  pivot_longer(cols = starts_with("Frequency"),
               names_to = "Score",
               values_to = "Frequency")

# Create the heatmap
ggplot(score_data_heatmap, aes(x = Score, y = Basename, fill = Frequency)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(x = "Score", y = "Agroinfiltration Image", fill = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

What does this plot show?

- Some agroinfiltration images contained more CDAs than others, since some have pale colours for all scores.

#### Proportions

```{r fig.width=8, fig.height=16}
#| fig-width: 8
#| fig-height: 16
score_data_grouped <- score_data %>%
  group_by(Basename) %>%
  summarize(
    Frequency_0 = sum(c(Score1, Score2, Score3) == 0, na.rm = TRUE),
    Frequency_1 = sum(c(Score1, Score2, Score3) == 1, na.rm = TRUE),
    Frequency_2 = sum(c(Score1, Score2, Score3) == 2, na.rm = TRUE),
    Frequency_3 = sum(c(Score1, Score2, Score3) == 3, na.rm = TRUE),
    Frequency_4 = sum(c(Score1, Score2, Score3) == 4, na.rm = TRUE),
    Frequency_5 = sum(c(Score1, Score2, Score3) == 5, na.rm = TRUE),
    Frequency_6 = sum(c(Score1, Score2, Score3) == 6, na.rm = TRUE),
    Total = n() * 3  # Adjust if some scores might be NA
  ) %>%
  mutate(
    Proportion_0 = Frequency_0 / Total,
    Proportion_1 = Frequency_1 / Total,
    Proportion_2 = Frequency_2 / Total,
    Proportion_3 = Frequency_3 / Total,
    Proportion_4 = Frequency_4 / Total,
    Proportion_5 = Frequency_5 / Total,
    Proportion_6 = Frequency_6 / Total
  ) %>%
  rowwise() %>%  # Ensure which.max is calculated per row
  mutate(Most_Common_Score = which.max(c(Proportion_0, Proportion_1, Proportion_2, Proportion_3, Proportion_4, Proportion_5, Proportion_6)) - 1) %>%
  ungroup()  # Always a good practice to ungroup after rowwise operations

# Reshape the data for heatmap visualization
score_data_heatmap <- score_data_grouped %>%
  pivot_longer(cols = starts_with("Proportion"),
               names_to = "Score",
               values_to = "Proportion") %>%
  mutate(Score = sub("Proportion_", "", Score))  # Clean up column names for display
```

```{r fig.width=8, fig.height=16}
#| fig-width: 8
#| fig-height: 16
ggplot(score_data_heatmap, aes(x = Score, y = Basename, fill = Proportion)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(x = "Score", y = "Agroinfiltration Image", fill = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

What does this plot show?

- Between agroinfiltration images, the proportion of scores collected varied, which may reflect the underlying experimental differences. For example some images contained many 0s and others contained none.

#### Sorted by proportion of 0s
```{r fig.width=8, fig.height=16}
#| fig-width: 8
#| fig-height: 16
score_data_heatmap2 <- score_data_grouped %>%
  select(Basename, Proportion_0) %>%
  left_join(score_data_heatmap, by = "Basename")

ggplot(score_data_heatmap2, aes(x = Score, y = reorder(Basename, -Proportion_0), fill = Proportion)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(x = "Score", y = "Agroinfiltration Image", fill = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

What does this plot show?

- Some images had no 0s, which is interesting because there is normally control infiltrations. Could this be something to do with overall image brightness affecting scoring?

Hypothesis forming:

1. There may be a relationship between the images which had no 0s scored for them. They may be very bright for example, such that all CDAs appear more strongly.

#### Sorted by most common score
```{r fig.width=8, fig.height=16}
#| fig-width: 8
#| fig-height: 16
score_data_sorted <- score_data_grouped %>%
  arrange(Most_Common_Score, Basename)
breakpoints <- which(diff(score_data_sorted$Most_Common_Score) != 0) + 0.5

score_data_heatmap3 <- score_data_grouped %>%
  pivot_longer(
    cols = starts_with("Proportion"),
    names_to = "Score",
    values_to = "Proportion"
  ) %>%
  mutate(Score = sub("Proportion_", "", Score)) %>%  # Clean up column names for display
  select(Basename, Score, Proportion, Most_Common_Score)  # Ensure Most_Common_Score is carried over

# Create the heatmap with proportions sorted by the most common score
ggplot(score_data_heatmap3, aes(x = Score, y = reorder(Basename, Most_Common_Score), fill = Proportion)) +
  geom_tile() +
  geom_hline(yintercept = breakpoints, linetype = "dashed", color = "red", size = 1) +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(x = "Score", y = "Agroinfiltration Image", fill = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

What does this plot show?

- Like the CDA images themselves, in many of the raw agroinfiltration images the most common score was 0.
- There were examples of images for which the most commoon score was each other score.
- Of the images where 6 was the most common score, it appears common that the scoring was often between 0 and 5/6 - the presence/absence type. Similar is true of 5s, where in both cases, 1, 2, and 3 are uncommon.

### Count of scores per raw image.
```{r fig.width=14, fig.height=10}
#| fig-width: 14
#| fig-height: 10
score_data_totals <- score_data %>%
  mutate(
    Score1 = ifelse(Score1 %in% c("[]", "NA"), NA, 1),
    Score2 = ifelse(Score2 %in% c("[]", "NA"), NA, 1),
    Score3 = ifelse(Score3 %in% c("[]", "NA"), NA, 1)
  ) %>%
  rowwise() %>%
  mutate(Total_Scores = sum(c(Score1, Score2, Score3), na.rm = TRUE)) %>%
  group_by(Basename) %>%
  summarize(Total_Counts = sum(Total_Scores, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(Total_Counts))  # Sort by Total_Counts in descending order


ggplot(score_data_totals, aes(x = reorder(Basename, -Total_Counts), y = Total_Counts)) +
  geom_point(stat = "identity", size = 2, color = "black", alpha=0.5) +
  theme_minimal() +
  labs(x = "Agroinfiltration Image", y = "Total Num Scores Collected") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

What does this plot show?

- There is a large range in the number of scores collected for each miage. 6 images had more than 300 scores collected for them. There were some images with 16 leaves of 8 CDAs (128), which, multiplied by 3 is 384.
- There is a large groupimages for which there were around 90 scores collected. There were many images with a dice-5 leaf pattern and 6 CDAs per leaf (30 CDAs), which, multiplied by 3 is 90. The trailing end to the right of that could be due to damaged leaves or missing CDAs.
- There were some images for which very few scores were collected (~20 total). There was an image with only 2 leaves of 4 CDAs each, which, multiplied by 3 would give 24.